```{r setup,echo=FALSE}
library(shiny)
```

# Understanding UI

As you saw in the previous chapter, Shiny encourages separation of the code that generates your user interface (`ui`) from the code that drives your app's behavior (`server`). In this chapter, we'll dive much deeper into the UI side of things.

## It's just HTML

**Fundamentally, Shiny UI is all about generating HTML.** Almost all of the functions we'll encounter below are merely thin wrappers around HTML generating functions from the `htmltools` package. These functions not only generate HTML elements, but also perform JavaScript/CSS dependency packaging, automatically escape text, and much more.

If you're used to writing your own HTML, using R functions to generate HTML for you may not sound appealing at first. But if you give it a chance, you'll find that this approach to HTML generation comes with many benefits, while retaining all the power and flexibility of manual coding. That said, if you really want to write your own angle brackets—or if you're collaborating with a designer who wants to give you HTML—Shiny will let you do that too, as we'll see at the end of this chapter (XREF).

To start, though, let's focus on the basic UI elements that are most commonly used in Shiny apps.

## Inputs

As we saw in the previous chapter, functions like `sliderInput`, `selectInput`, `textInput`, and `numericInput` are used to insert input controls into your UI.

The first parameter of an input function is always the input ID; this is a simple string that is composed of alphanumeric characters and/or underscore (no spaces, dashes, periods, or other special characters please!). Generally, there is a second parameter `label` that is used to create a human-readable label for the control. Any remaining parameters are specific to the particular input function, and can be used to customize the input control.

For example, a typical call to `sliderInput` might look something like this:

```r
shiny::sliderInput("min", "Limit (minimum)", min = 0, max = 100, value = 50)
```

```{r echo=FALSE, fig.height=0.2, message=FALSE, cache=TRUE}

library(shiny)
shinyApp(
  fluidPage(shiny::sliderInput("min", "Limit (minimum)", min = 0, max = 100, value = 50)),
  function(input, output, session) {}
)
```

In the server function, the value of this slider would be accessed via `input$min`.

It's absolutely vital that each input have a *unique* ID. Using the same ID value for more than one input or output in the same app will result in errors or incorrect results.

Shiny itself comes with a variety of input functions out of the box:

- `sliderInput`
- `selectInput`/`selectizeInput`
- `checkboxGroupInput`/`checkboxInput`/`radioButtons`
- `dateInput`/`dateRangeInput`
- `fileInput`
- `numericInput`
- `textInput`/`passwordInput`
- `actionButton`

Each input function has its own unique look and functionality, and takes different arguments. But they all share the same two properties of 1) taking a unique input ID, and 2) exposing values to the server function via a slot in the `input` object.

## Outputs

Output functions are used to tell Shiny _where_ and _how_ to place outputs that are defined in the app's server.

Like inputs, outputs take a unique ID as their first argument. These IDs must be unique among all inputs _and_ outputs!

Unlike inputs, outputs generally start out as empty rectangles, that need to be fed data from the server in order to actually appear.

## Layouts and panels

Shiny includes several classes of UI functions that behave like neither inputs nor outputs. Rather, these functions help with the layout and formatting of your UI.

### Page functions

The first function you're likely to encounter in a Shiny UI is a page function. Page functions expect to be the outermost function call in your UI, and set up your web page to contain other content.

The most common page function is `fluidPage`.

```r
fluidPage(..., title = NULL, theme = NULL)
```

`fluidPage` sets up your page to use the Bootstrap CSS framework. Bootstrap provides your web page with attractive settings for typography and spacing, and also preloads dozens of CSS rules that can be invoked to visually organize and enhance specific areas of your UI. We'll take advantage of quite a few of these Bootstrap rules as we proceed through this chapter.

The "fluid" in `fluidPage` means the page's content may resize its width (but not height) as the size of the browser window changes. (The other option is "fixed", which means the page contents will never exceed 960 pixels in width.)

* sidebar
* tabset
* bootstrap grid

## Generating HTML

To understand how UI functions in R work, let's first talk about HTML, in case you're not familiar with it (or its more general descendant, XML).

HTML is a _markup language_ for describing web pages. A markup language is just a document format that contains plain text content, plus embedded instructions for annotating, or "marking up", specific sections of that content. These instructions can control the appearance, layout, and behavior of the text they mark up, and also provide structure to the document.

Here's a simple snippet of HTML:

```html
This time I <em>really</em> mean it!
```

The `<em>` and `</em>` markup instructions indicate that the word `really` should be displayed with special <strong>em</strong>phasis (italics):

```{r echo=FALSE}
tags$blockquote(HTML("This time I <em>really</em> mean it!"))
```

`<em>` is an example of a _start tag_, and `</em>` (note the slash character) is an example of an _end tag_.

### Inline formatting tags

`em` is just one of many HTML tags that are used to format text:

* `<strong>...</strong>` makes text <strong>bold</strong>
* `<u>...</u>` makes text <u>underlined</u>
* `<s>...</s>` makes text <s>strikeout</s>

### Block tags

Another class of tags is used to wrap entire blocks of text. You can use `<p>...</p>` to break text into distinct paragraphs, or `<h3>...</h3>` to turn a line into a subheading.

```{css echo=FALSE}
.sourceCode.html {
  white-space: pre-wrap !important;
}
```

```html
<h3>Chapter I. Down the Rabbit-Hole</h3>

<p>Alice was beginning to get very tired of sitting by her sister on the bank, and of having nothing to do: once or twice she had peeped into the book her sister was reading, but it had no pictures or conversations in it, ‘and what is the use of a book,’ thought Alice ‘without pictures or conversations?’</p>

<p>So she was considering in her own mind (as well as she could, for the hot day made her feel very sleepy and stupid), whether the pleasure of making a daisy-chain would be worth the trouble of getting up and picking the daisies, when suddenly a White Rabbit with pink eyes ran close by her.</p>
```

When rendered, this HTML looks like:

```{r echo=FALSE}
tags$blockquote(
  HTML("<h3>Chapter I. Down the Rabbit-Hole</h3>
<p>Alice was beginning to get very tired of sitting by her sister on the bank, and of having nothing to do: once or twice she had peeped into the book her sister was reading, but it had no pictures or conversations in it, ‘and what is the use of a book,’ thought Alice ‘without pictures or conversations?’</p>
<p>So she was considering in her own mind (as well as she could, for the hot day made her feel very sleepy and stupid), whether the pleasure of making a daisy-chain would be worth the trouble of getting up and picking the daisies, when suddenly a White Rabbit with pink eyes ran close by her.</p>")
)
```

### Tag attributes

Some tags need to do more than just demarcate some text. An `<a>` (for "anchor") tag is used to create a hyperlink. It's not enough to just wrap `<a>...</a>` around the link's text, as you also need to specify where the hyperlink points to.

Start tags let you include _attributes_ that customize the appearance or behavior of the tag. In this case, we'll add an `href` attribute to our `<a>` start tag:

```html
<p>Learn more about <strong>Shiny</strong> at <a href="https://shiny.rstudio.com">this website</a>.</p>
```

```{r echo=FALSE}
tags$blockquote(HTML('<p>Learn more about <strong>Shiny</strong> at <a href="https://shiny.rstudio.com">this website</a>.</p>'))
```

There are [dozens of attributes](https://developer.mozilla.org/en-US/docs/Web/HTML/Global_attributes) that all tags accept, and hundreds of attributes that are specific to particular tags. You don't have to worry about memorizing them all--even full-time web developers don't do that. There are two attributes that are used constantly, though.

The `id` attribute uniquely identifies a tag in a document. That is, no two tags in a single document should share the same `id` value, and each tag can have zero or one `id` value. As far as the web browser is concerned, the `id` attribute is completely optional and has no intrinsic effect on the appearance or behavior of the rendered tag. However, it's incredibly useful for identifying a tag for special treatment by CSS or JavaScript, and as such, plays a crucial role for Shiny apps.

The `class` attribute provides a way of classifying tags in a document. Unlike `id`, any number of tags can have the same class, and each tag can have multiple classes (space separated). Again, classes don't have an intrinsic effect, but are hugely helpful for using CSS or JavaScript to target groups of tags.

In the following example, we've given a `<p>` tag an id and two classes:

```html
<p id="storage-low-message" class="message warning">Storage space is running low!</p>
```

```{r echo=FALSE}
tags$blockquote(HTML('<p id="storage-low-message" class="message warning">Storage space is running low!</p>'))
```

Here, the `id` and `class` values have had no discernible effect. But we could, for example, write CSS that any elements with the `message` class should appear at the top of the page, and that any elements with the `warning` class should have a yellow background and bold text; and we could write JavaScript that automatically dismisses the message if the storage situation improves.

### Parents and children

In the example above, we have a `<p>` tag that contains some text that contains `<strong>` and `<a>` tag.s We can refer to `<p>` as the _parent_ of `<strong>`/`<a>`, and `<strong>`/`<a>` as the _children_ of `<p>`. And naturally, `<strong>` and `<a>` are called _siblings_.

It's often helpful to think of tags and text as forming a tree structure:

```
<p>
├── "Learn more about"
├── <strong>
│   └── "Shiny"
├── "at"
├── <a href="...">
│   └── "this website"
└── "."
```

### Comments

Just as you can use the `#` character to comment out a line of R code, HTML lets you comment out parts of your web page. Use `<!--` to start a comment, and `-->` to end one. Anything between these delimiters will be ignored during the rendering of the web page, although it will still be visible to anyone who looks at your raw HTML by using the browser's View Source command.

```html
<p>This HTML will be seen.</p>

<!-- <p>This HTML will not.</p> -->

<!--
<p>
Nor will this.
</p>
-->
```

```{r echo=FALSE}
tags$blockquote(HTML("<p>This HTML will be seen.</p>

<!-- <p>This HTML will not.</p> -->

<!--
<p>
Nor will this.
</p>
-->"))
```

### Escaping

Any markup language like HTML, where there are characters that have special meaning, needs to provide a way to "escape" those special characters--that is, to insert a special character into the document without invoking its special meaning.

For example, the `<` character in HTML has a special meaning, as it indicates the start of a tag. What if you actually want to insert a `<` character into the rendered document--or, let's say, an entire `<p>` tag?

```html
<p>In HTML, you start paragraphs with "<p>" and end them with "</p>".</p>
```

```{r echo=FALSE}
tags$blockquote(HTML('<p>In HTML, you start paragraphs with "<p>" and end them with "</p>".</p>'))
```

That doesn't look as we intended at all! The browser has no way of knowing that we meant the outer `<p>` and `</p>` to be interpreted as markup, and the inner `<p>` and `</p>` to be interpreted as text.

Instead, we need to escape the inner tags so they become text. The escaped version of `<` is `&lt;`, and `>` is `&gt;`.

```html
<p>In HTML, you start paragraphs with "&lt;p&gt;" and end them with "&lt;/p&gt;".</p>
```

```{r echo=FALSE}
tags$blockquote(HTML('<p>In HTML, you start paragraphs with "&lt;p&gt;" and end them with "&lt;/p&gt;".</p>'))
```

(Yes, escaped characters look pretty ugly. That's just how it is.)

Each escaped character in HTML starts with `&` and ends with `;`. There are lots of valid sequences of characters that go between, but besides `lt` (less than) and `gt` (greater than), the only one you're likely to need to know is `amp`; `&amp;` is how you insert a `&` character into HTML.

Escaping `<`, `>`, and `&` is mandatory if you don't want them interpreted as special characters; other characters can be expressed as escape sequences, but it's generally not necessary. Escaping `<`, `>`, and `&` is so common and crucial that every web framework contains a function for doing it (in our case it's `htmltools::htmlEscape`), but as we'll see in a moment, Shiny will usually do this for you automatically.

### HTML tag vocabulary

That concludes our whirlwind tour of HTML syntax. I've glossed over some minutiae, but by and large, this is all anyone needs to know about the structure of HTML documents, unless they're building a web browser.

The much larger part of learning HTML is getting to know the actual tags that are available to you, what attributes they offer, and how they work with each other.

### Expressing HTML in R






## Customizing with CSS and JavaScript

The native languages of the web are HTML (for content), CSS (for styling), and JavaScript (for behavior). Shiny is designed to be accessible for R users who aren't familiar with any of those languages. But if you do speak these languages, you can take full advantage of them with Shiny to customize your apps or extend the Shiny framework.



* Custom CSS
* www directory

## Alternate toolkits

* flexdashboard
* shinyBS
* shinyDashboard
* shinyjs
* shinythemes

## HTML templates
