# Reactive programming

* lexical scoping
* reactive primitives
* side effects
* reactive graph



In this chapter, we'll form a mental model of reactive programming. Without reactive programming, Shiny would not exist, and in educated hands it is by far the most elegant, robust, and productive way we know of for creating dynamic user interfaces.

The biggest drawback of reactive programming is that it is not intuitive. Even experienced R users can trouble getting their heads around it as they get started with Shiny, and those with deep experience in software engineering may dismiss it as too "magical". But once we've formed an accurate mental model, you'll see that there is nothing to fear.

## Prerequisite 1: Understanding lexical scoping

Before we can understand reactive programming, we need to understand *lexical scoping*. "Scoping" means the algorithm a programming language uses to look for variables, and lexical scoping is by far the most common type of scoping among languages commonly used today.

If you've written a function before, you know that variables you define inside a function body are local to that function:

```{r error=TRUE}
func1 <- function() {
  # A local variable
  x <- 1
  print(x)
}
func1()
x
```

In this example, the `x` variable is defined inside the function, so attempts to read this variable from outside the function fail.

Hopefully, you're also familiar with global variables. These can be read from inside of any function, unless a variable of the same name has been defined inside the function.

```{r}
# A global variable (defined outside any function)
x <- 1

func2 <- function() {
  print(x) # prints `1`
  x <- 2
  print(x) # prints `2`
}
func2()
x # The global variable is still `1`
```

You can get quite far as an R programmer knowing only about global and local variable scopes. What you may not have realized is that there can be additional scopes between the global and local ones! These intermediate scopes play a particularly important role in Shiny apps, as we'll soon see.

The body of the following `createCounter` function defines two functions: `value` and `increment`. When a function is defined inside of another function, we call it a _nested function_.

```{r}
createCounter <- function(startingValue = 0) {
  counter <- startingValue
  
  value <- function() {
    counter
  }
  
  increment <- function() {
    counter <<- counter + 1L
  }
  
  list(
    value = value,
    increment = increment
  )
}

counter1 <- createCounter()
counter1$value()
counter1$increment()
counter1$value()
```

The `value` function returns the value of `counter`. In this case, `counter` is present in neither the local scope (i.e. defined inside of the `value` function) nor the global scope (i.e. defined outside of any function), but in a _parent scope_, that is, the local variable scope of the containing function. This type of parent/child scoping is also recursive, meaning that if the nested function `value` itself defined a (doubly-)nested function, that new function would also have access to the `counter` variable.

Think of it this way. The line `counter <- startingValue` defines a variable. To figure out where that variable may be used from, scan backwards for the opening curly brace `{` that directly contains this line of code. All of the code between that opening curly brace and its matching closing curly brace `}` is able to read `counter`, including from inside of nested functions. Any functions whose text appears outside of those braces cannot read `counter`.

This is the way things work by default in R, and it's called _lexical scoping_. It's called "lexical" because you can figure out the relationships between different variable scopes by looking at the positions of curly braces in the source code. (In case you're curious, the other type of scoping is called "dynamic scoping" and it's also used in R, though in smaller doses. See the `base::with` function for one example of dynamic scoping.)

Shiny relies heavily on lexical scoping to allow reactive inputs, expressions, outputs, and observers to have relationships with each other, as we'll see soon.

## Prerequisite 2: Side effects

You can think of all R functions as falling into one of two categories:

1. **Functions that are only interesting for the result that they return**, e.g. `c()`, `list()`, `lm()`, `predict()`, `read.csv()`, `ls()`.
2. **Functions that perform some kind of action** (besides calculating a result), e.g. `print()`, `plot()`, `write.csv()`, `rm()`, `install.packages()`.

_Side effects_ are what we call the kinds of actions in the second category. Unlike with pharmaceuticals, where side effects are always unintentional and usually negative, we simply mean any effects apart from a function's return value. Changing a file on disk is a side effect. Printing words to the console is a side effect. Sending a message to another computer is a side effect.

As a shorthand, you can think of functions without side effects as _calculations_, and functions with side effects as _actions_.^[Computer scientists and Haskell programmers refer to functions without side effects as "pure" and functions with side effects as "impure"; though their definition of side effects includes the reading of any variable or state that isn't a function argument. For Shiny purposes, we don't need to be so strict.] 

As we learn to write more complicated Shiny apps, it's important to divide your code into calculations and actions. Shiny has opinions about what reactive abstractions you should use to wrap each kind of code, and if you structure your app in accordance with those opinions you'll end up with code that's more reliable, better performing, and easier to maintain.


