# Dynamic UI {#action-dynamic}

```{r, include = FALSE}
source("common.R")
```

So far, we've seen a clean separation between the user interface and the server function. The user interface is defined statically, when the app is launched. That means it can't respond to anything that happens in the app. In this chapter, you'll learn how to create __dynamic__ user interfaces, using code run in the server function. 

There are three key techniques for creating dynamic user interfaces:

* Using the `update` functions to modify the parameters of input controls.

* Using `tabsetPanel()` to conditionally show and hide different UI.

* Using `uiOutput()` and `renderUI()` to generate entinre parts of the user 
  interface from code. 

These three tools give you considerable power to respond to the user's actions by modifying the user interface. I'll demonstrate some of the more useful ways in which you can apply them, but ultimately, you are only constrained by your creativity. That said, these tools can make your app substantially more difficult to reason about, so you should deploy them sparingly, and always use the simplest technique that solves your problem.

```{r setup}
library(shiny)
```

## Updating inputs

We'll begin with the simplest technique that allows you to tweak the parameters of an input after it has been created. Every input control, e.g. `textInput()`, is paired with a update function, e.g. `updateTextInput()`, that allows you to modify the control after it has been created. Each update function has the same arguments as the corresponding constructor, allowing you to change any of arguments after construction.

Take this very simple example. The app has three two inputs that control the range (the `min` and `max`) of the a slider. The key idea is to use `observeEvent()`[^observe] to trigger `updateNumericInput()` whenever the `min` or `max` inputs change. `updateNumericInput()` looks a little different to other Shiny functions: it takes the current `session` as the first argument, and the name of the input (as a string) as the second. The remaining arguments match the arguments to the input constructor.

```{r}
ui <- fluidPage(
  numericInput("min", "min", 0),
  numericInput("max", "max", 3),
  sliderInput("n", "n", min = 0, max = 3, value = 1)
)
server <- function(input, output, session) {
  observeEvent(input$min, {
    updateNumericInput(session, "n", min = input$min)
  })  
  observeEvent(input$max, {
    updateNumericInput(session, "n", max = input$max)
  })
}
```

<!-- figure showing four screenshots after updating min and max  -->

[^observe]: Note that I've used `observeEvent()` here, although `observe()` would also have worked and would yield shorter code. I generally prefer `observeEvent()` because its argument cleanly separate the event you're listening for from the action you want to take in response.

To help you get the hang of the update functions, I'll next show a couple more simple examples. Then we'll dive into a more complicated case study using hierarchical select boxes. I'll finish off by discussing circular references, a major challenge with update functions, and the related problem of multiple sources of truth.

### Simple uses

The simplest uses of the update functions are to provide small conveniences for the user. For example, maybe you want to make it easy to reset some parameters back to their starting place. The following snippet shows how you might combine an `actionButton()`, `observeEvent()` and `updateNumericInput()`:

```{r}
ui <- fluidPage(
  sliderInput("x1", "x1", 0, min = -10, max = 10),
  sliderInput("x2", "x2", 0, min = -10, max = 10),
  sliderInput("x3", "x3", 0, min = -10, max = 10),
  actionButton("reset", "Reset")
)

server <- function(input, output, session) {
  observeEvent(input$reset, {
    updateNumericInput(session, "x1", value = 0)
    updateNumericInput(session, "x2", value = 0)
    updateNumericInput(session, "x3", value = 0)
  })
}
```

A similar application is to tweak the text of an action button so you know exactly what it's going to do:

```{r}
ui <- fluidPage(
  numericInput("n", "Simulations", 10),
  actionButton("simulate", "Simulate")
)

server <- function(input, output, session) {
  observeEvent(input$n, {
    label <- paste0("Simulate ", input$n, " times")
    updateActionButton(session, "simulate", label = label)
  })
}
```

There are many ways to use update functions in this way; be on the look out for ways to give more information to the user when you are working on sophisticated applications. A particularly important application is making it easier to select from a long list of possible options, through step-by-step filtering. That's often a problem for "hierarchical select boxes".

### Hierarchical select boxes

A more complicated, but particularly useful, application of the update functions is to allow interactive drill down across multiple categories. I'm going to start some imaginary data for a sales dashboard, coming from <https://www.kaggle.com/kyanyoga/sample-sales-data>.

For our purposes, I'm going to focus on a natural hierarchy in the data:

* Each territory contains customers
* Each customer has multiple orders
* Each order contains rows

I want to create a user interface where you can:

* Select a territory to see all customers.
* Select customer to see all orders.
* Select order to see data.

The basic idea is to start with UI that contains three select boxes and one output table. The choices for the `customername` and `ordernumber` will be dynamically supplied so I start with `choices = NULL`.

```{r}
sales <- vroom::vroom("sales-dashboard/sales_data_sample.csv")

ui <- fluidPage(
  selectInput("territory", "Territory", choices = unique(sales$TERRITORY)),
  selectInput("customername", "Customer", choices = NULL),
  selectInput("ordernumber", "Order number", choices = NULL),
  tableOutput("data")
)
```

In the server function, I work top-down:

1.  I create a reactive, `territory`, that contains `sales` just for the
    selected territory 

1.  Whenever `territory()` changes, I update the list of possible customer
    choices.
    
1.  I create a reactive, `customer`,  that contains only the rows from
    `territory` matching the selected customer.
    
1.  Whenever `customer` changes, I update the `choices` for the order
    select box.
    
1.  I display the selected orders in `output$data`.

```{r}
server <- function(input, output, session) {
  territory <- reactive({
    filter(sales, TERRITORY == input$territory)
  })
  observeEvent(territory(), {
    choices <- unique(territory()$CUSTOMERNAME)
    updateSelectInput(session, "customername", choices = choices) 
  })
  
  customer <- reactive({
    filter(territory(), CUSTOMERNAME == input$customername)
  })
  observeEvent(input$customername, {
    choices <- unique(customer()$ORDERNUMBER)
    updateSelectInput(session, "ordernumber", choices = choices)
  })
  
  output$data <- renderTable({
    filter(customer(), ORDERNUMBER == input$ordernumber)
  })
}
```

<!-- TODO: diagram -->

You can see a more fleshed out application of this principle in <https://github.com/hadley/mastering-shiny/tree/master/sales-dashboard>.

Consider using this, or a similar technique, whenever the user has to select from a large number of choices. You're free to use any other inputs for filtering: if you (e.g.) wanted to make it easier to find orders based on the total value, you might want to use `sliderInput()`s instead.

### Circular references

If you want to use update functions to change the `value`[^other-params] of inputs, there's an important issue we need to discuss. From Shiny's perspectve, using an update function to modify `value` is no different to the user modifying the value. That means an update function can trigger reactive updates in exactly the same way that a human can. You are now stepping outside of the bounds of pure reactive programming, so you need to start worrying about circular references and creating infinite loops. 

[^other-params]: This is generally only a concern when you are changing the `value`, but be aware that changing some other parameters can occassionally change the value, as when changing set of `choices` for `selectInput()`, or when changing `min` and `max` for `sliderInput()`.

For example, take the following simple app. It contains a single input control and a observer that increments its value by one. Every time `updateNumericInput()` runs, it invalidates `input$n`, causing `updateNumericInput()` to run again, so the app is stuck in an infinite loop constantly increasing the value of `input$n`.

```{r}
ui <- fluidPage(
  numericInput("n", "n", 0)
)
server <- function(input, output, session) {
  observeEvent(input$n,
    updateNumericInput(session, "n", value = input$n + 1)
  )
}
```

You're unlikely to create such an obvious problem in your own app, but beware this potential problem if you are updating multiple controls that depend on one another. You will need to carefully reason through the updates to ensure that you're not creating an infinite loop of updates.

### Inter-related inputs

One place where it's easy to end up with circular references if you're attempting to have multiple "sources of truth" in an app. For example, imagine you want to create a temperatue conversion app where you can either enter the temperature in Celsius or in Fahrenheit:

```{r}
ui <- fluidPage(
  numericInput("temp_c", "Celsius", NA),
  numericInput("temp_f", "Fahrenheit", NA)
)

server <- function(input, output, session) {
  observeEvent(input$temp_f, {
    c <- round((input$temp_f - 32) * 5 / 9)
    updateNumericInput(session, "temp_c", value = c)
  })
  
  observeEvent(input$temp_c, {
    f <- round((input$temp_c * 9 / 5) + 32)
    updateNumericInput(session, "temp_f", value = f)
  })
}
```

If you play around with this app, you'll notice that it _mostly_ works, but you might notice that it'll sometimes trigger multiple changes. For example:

* Set 120 F, then click the down arrow.
* F changes to 119, and C is updated to 48.
* 48 C converts to 118 F, so F changes again to 118.
* Fortunately 118 F is still 48 C, so the updates stop there.

There's no way around this problem because you have one idea (the temperature) with two expressions in the app (Celsius and Fahrenheit). Here we are lucky that cycle quickly converges to a value that satisfies both constraints.

<!-- Similar problem: https://community.rstudio.com/t/mutually-dependent-numericinput-in-shiny/29307 -->

## Showing and hiding

The next step up in complexity is to selectively show and hide parts of the UI. Later, when you learn more about CSS and JS, you'll learn more approaches, but there's one powerful technique that you can use already: concealing optional UI in a tabset. This is a clever hack, that allows you to show and hide UI as needed, without having to re-generate it from scratch (as you'll learn how to do in the next section).

Currently[^tab-future] you need a dash of css to make it work: in the following UI specification `tags$style("#switcher { display:none; }")` hides the tab switcher. If you adapt this for your own code make sure to repace `switcher` with id of your the `tabsetPanel()`.

[^tab-future]: Hopefully, it will be built into `tabsetPanel()` in the future; follow <https://github.com/rstudio/shiny/issues/2680> for details.

```{r}
ui <- fluidPage(
  tags$style("#switcher { display:none; }"),
  sidebarLayout(
    sidebarPanel(
      selectInput("controller", "Show", choices = paste0("panel", 1:3))
    ),
    mainPanel(
      tabsetPanel(
        id = "switcher",
        tabPanel("panel1", "Panel 1 content"),
        tabPanel("panel2", "Panel 2 content"),
        tabPanel("panel3", "Panel 3 content")
      )
    )
  )
)

server <- function(input, output, session) {
  observeEvent(input$controller, {
    updateTabsetPanel(session, "switcher", selected = input$controller)
  })
}
```

There are two main ideas here:

* Use tabset panel with hidden tabs.
* Use `updateTabsetPanel()` to switch tabs as needed.

This is a simple idea, but it gives you a considerable amount of flexibility. And with a little creativity you can do some neat things. The following two sections illustrate how you might use it to collect different information depending on another selection, or how you might create a wizard like interface.

### Conditional UI

Imagine that you want to create a shiny app that allows the user to simulate from the normal, uniform, and expontential distributions. Each distribution has different parameters so how do you let the user pick different ones?

We can use a tabset to provide a different set of controls for each distribution:

```{r}
param_tabs <- tagList(
  tags$style("#params { display:none; }"),
  tabsetPanel(id = "params",
    tabPanel("normal",
      numericInput("mean", "mean", value = 1),
      numericInput("sd", "standard deviation", min = 0, value = 1)
    ),
    tabPanel("uniform", 
      numericInput("min", "min", value = 0),
      numericInput("max", "max", value = 1)
    ),
    tabPanel("exponential",
      numericInput("rate", "rate", value = 1, min = 0),
    )
  )
)
```

We then embed that inside a fuller UI. Here I'm going to show a histogram of the results, and the values individually so you can copy and paste them.

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("dist", "Distribution", choices = c("normal", "uniform", "exponential")),
      numericInput("n", "Number of samples", value = 100),
      param_tabs,
    ),
    mainPanel(
      plotOutput("hist"),
      textOutput("values")
    )
  )
)
```

The server function has two main components: an `observeEvent()` to automatically switch controls when the distribution changes, and a reactive that samples from the selected distribution.

```{r}
server <- function(input, output, session) {
  observeEvent(input$dist, {
    updateTabsetPanel(session, "params", selected = input$dist)
  }) 
  
  sample <- reactive({
    switch(input$dist,
      normal = rnorm(input$n, input$mean, input$sd),
      uniform = runif(input$n, input$min, input$max),
      exponential = rexp(input$n, input$rate)
    )
  })
  output$hist <- renderPlot(hist(sample()))
  output$values <- renderText(round(sample(), 3))
}
shinyApp(ui, server)
```

### Wizard interface

The following code shows you how might create a "wizard", a type of interface that makes it easier to collect a lot of information by breaking it down into a progression of simple steps. 

```{r}
ui <- fluidPage(
  tags$style("#wizard { display:none; }"),
  tabsetPanel(id = "wizard",
    tabPanel("page1", 
      "Welcome!",
      actionButton("page12", "next")
    ),
    tabPanel("page2", 
      "Only one page to go",
      actionButton("page21", "prev"),
      actionButton("page23", "next")
    ),
    tabPanel("page3", 
      "You're done!",
      actionButton("page32", "prev")
    )
  )
)

server <- function(input, output, session) {
  switch_tab <- function(page) {
    updateTabsetPanel(session, "wizard", selected = page)
  }
  
  observeEvent(input$page12, switch_tab("page2"))
  observeEvent(input$page21, switch_tab("page1"))
  observeEvent(input$page23, switch_tab("page3"))
  observeEvent(input$page32, switch_tab("page2"))
}
```

You can imagine how you might change this code so to only show certain pages if some conditions are met. For example, you'd only collect additional information if some condition was true.

## Dynamic UI

Sometimes modifying the values of an existing control is not enough, and you need control the number of type of controls. There's a special pair of output and render functions that allow you to do UI on the server: `uiOutput()` and `renderUI()`

Here's a simple example to illustrate the basic idea. It dynamically creates a different type of input control depending on an input:

```{r}
ui <- fluidPage(
  textInput("label", "label"),
  selectInput("type", "type", c("slider", "numeric")),
  uiOutput("numeric")
)
server <- function(input, output, session) {
  output$numeric <- renderUI({
    if (input$type == "slider") {
      sliderInput("dynamic", input$label, value = 0, min = 0, max = 100)
    } else {
      numericInput("dynamic", input$label, value = 0, min = 0, max = 100)  
    }
  })
}
```

If you run this code yourself, you'll notice that it takes a fraction of a second to appear after the app loads - that's because it has to be rendered by the server function. Relying too much on `renderUI()` can create a laggy UI. Want to keep as much as of the fixed structure in UI as possible. Better peformance. Simpler reasoning.

Note that you are now creating IDs in two places so that when you add to `ui`, you have to be careful not to call the control `dynamic`. Later, in Chapter \@ref(action-modules) we'll see how modules can help avoid this problem.

There's one other problem with this approach: when you change controls, you lose the value that the user selected. It's possible to fix by setting the value to the current value. We haven't talked about `isolate()` yet; allows you to get a reactive value _without_ creating a reactive dependency.

```{r}
server <- function(input, output, session) {
  output$numeric <- renderUI({
    if (input$type == "slider") {
      sliderInput("dynamic", input$label, isolate(input$dynamic), min = 0, max = 100)
    } else {
      numericInput("dynamic", input$label, isolate(input$dynamic), min = 0, max = 100)  
    }
  })
}
```

Maintaining existing user input is one of the big challenges of working with `renderUI()`. This is one reason that selecting showing and hiding UI (as above) is a better appraoch it works for you - because you're not destroying and recreating the controls, you don't need to do anything to prserve the values.

Dynamic UI is most useful when you are generating an arbitrary number or type of controls. That typically means that you'll be generating UI with code, and function programming is a good fit for this sort of task. If you're not familiar with the basic `map()` and `reduce()` of functional programming, you might want to take a brief detour to read <https://adv-r.hadley.nz/functionals.html> before continuing.

```{r}
library(purrr)
```

### Multiple controls

A more realistic use of `dynamicUI` is when you need to create an arbitrary number of controls. For example, imagine you'd like the user to be able to supply their own palette of colours. 

The UI side is pretty simple: we have a numeric input that controls that will control the number of text inputs; and then a UI output where the text boxes will go.

```{r}
ui <- fluidPage(
  numericInput("n", "Number of colours", value = 5, min = 1),
  uiOutput("col"),
  textOutput("palette")
)
```

Then in the server function I first a make a reactive that will provide the names for each control, 

* I create the dynamic UI by using `map()` to create a `textInput()` once for 
  each name.
  
* To generate the output, I need to use use a new trick. So far we've always
  accessed the components of `input` with `$`. But we can also use `[[`, which
  is what we need here because we have the name of the component stored in a 
  character vector. I use `map_chr()` to collect them all into a character
  vector.

```{r}
server <- function(input, output, session) {
  col_names <- reactive(paste0("col", seq_len(input$n)))
  
  output$col <- renderUI({
    map(col_names(), ~ textInput(.x, NULL))
  })
  
  output$palette <- renderText({
    map_chr(col_names(), ~ input[[.x]])
  })
}
```

If ran this app, you'd discover a really annyoing behaviour: whenever you change the number of colours, all the data you've entered disappears. We can fix this problem by using the same technique as before: setting `value` to the current value (using `isolate()` so we don't create an reactive dependency that we don't want).

I'll also tweak the apperance to look a little nicer, including displaying the selected colours in a plot.

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      numericInput("n", "Number of colours", value = 5, min = 1),
      uiOutput("col"),
    ),
    mainPanel(
      plotOutput("plot")  
    )
  )
)
server <- function(input, output, session) {
  col_names <- reactive(paste0("col", seq_len(input$n)))
  
  output$col <- renderUI({
    map(col_names(), ~ textInput(.x, NULL, value = isolate(input[[.x]])))
  })
  
  output$plot <- renderPlot({
    cols <- map_chr(col_names(), ~ input[[.x]] %||% "")
    cols[cols == ""] <- NA
    
    barplot(
      rep(1, length(cols)), 
      col = cols,
      space = 0, 
      axes = FALSE
    )
  })
}
shinyApp(ui, server)
```

### Dynamic filtering

To finish off the chapter, I'm going to create an app that generates a filtering interface for any data frame. 

To begin I'll create a function that generates either a slider for numeric inputs, or a multi-select for factor inputs:

```{r}
make_ui <- function(df, var) {
  x <- df[[var]]
  if (is.numeric(x)) {
    rng <- range(x, na.rm = TRUE)
    sliderInput(var, var, min = rng[1], max = rng[2], value = rng)
  } else if (is.factor(x)) {
    levs <- levels(x)
    selectInput(var, var, choices = levs, selected = levs, multiple = TRUE)
  } else {
    # Not supported
    NULL
  }
}
```

And then I'll write a function that takes the value of that control and returns a logical vector saying whether or not to include it. I chose to return a logical vector here because it'll make it easy to combine the results from multiple columns.

```{r}
filter_var <- function(input, df, var) {
  x <- df[[var]]

  if (is.numeric(x)) {
    x >= input[[var]][1] & x <= input[[var]][2]
  } else if (is.factor(x)) {
    x %in% input[[var]]
  } else {
    # No control
    TRUE
  }
}
```

I'll illustrate how these bits fit together on the iris dataset:

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      make_ui(iris, "Sepal.Length"),
      make_ui(iris, "Sepal.Width"),
      make_ui(iris, "Species")
    ),
    mainPanel(
      tableOutput("data")
    )
  )
)
server <- function(input, output, session) {
  selected <- reactive({
    filter_var(input, iris, "Sepal.Length") &
      filter_var(input, iris, "Sepal.Width") &
      filter_var(input, iris, "Species")
  })
  
  output$data <- renderTable(iris[selected(), ])
}
```

I got sick of copying and pasting so the above app only works with three columns. I can generalise to all the columns by using some functional programming:

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      tagList(map(names(iris), ~ make_ui(iris, .x)))
    ),
    mainPanel(
      tableOutput("data")
    )
  )
)
server <- function(input, output, session) {
  selected <- reactive({
    each_var <- map(names(iris), ~ filter_var(input, iris, .x))
    reduce(each_var, `&`)
  })
  
  output$data <- renderTable(iris[selected(), ])
}
```

From there, it's a simple generalisation to allow the user to provide the dataset.

```{r}
dfs <- keep(ls("package:datasets"), ~ is.data.frame(get(.x, "package:datasets")))

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("dataset", label = "Dataset", choices = dfs),
      uiOutput("filter")
    ),
    mainPanel(
      tableOutput("data")
    )
  )
)
server <- function(input, output, session) {
  data <- reactive({
    get(input$dataset, "package:datasets")
  })
  vars <- reactive(names(data()))
  
  output$filter <- renderUI(
    tagList(map(vars(), ~ make_ui(data(), .x)))
  )
  
  selected <- reactive({
    each_var <- map(vars(), ~ filter_var(input, data(), .x))
    reduce(each_var, `&`)
  })
  
  output$data <- renderTable(data()[selected(), ])
}
```

