# Uploads and downloads {#action-transfer}

```{r, include = FALSE}
source("common.R")
```

## Upload

`fileInput()`

Returns a data frame with columns:

* `name`: name of the file.

* `size`: file size, in bytes. By default the file size is capped at 5 meg. 
   If you want to increase this set `options(shiny.maxRequestSize)` to number 
   of bytes.
   

* `type`: the "mime type" of the file.

* `datapath`: the path to where the path has been uploaded to the server.
  Temporary directory. Made up name (number). This is a transient path; if the 
  user uploads more files it will go away.

You can explore with a simple app:

```{r}
ui <- fluidPage(
  fileInput("upload", NULL, multiple = TRUE),
  tableOutput("files")
)
server <- function(input, output, session) {
  output$files <- renderTable(input$upload)
}
```

Use of `req()`

```{r}
ui <- fluidPage(
  fileInput("file", NULL),
  tableOutput("description")
)

server <- function(input, output, session) {
  data <- reactive({
    req(input$file)
    ext <- tools::file_ext(input$file$filename)
    validate(need(ext == "csv", "Please upload a CSV file"))
    
    read.csv(input$file$datapath)
  })
  
  output$description <- renderTable({
    data()
  })
}
```

## Download

