# Bookmarking {#action-bookmark}

```{r, include = FALSE}
source("common.R")
```

By default, Shiny apps have one major drawback compared to most apps you'll on the internet: the app url does not capture the current state of the app. This means that you can't bookmark the current state and return to it in the future, and there's no way to share your current location with someone else in an email or text. While unfortunately Shiny can't do this by default, fortunately it's a behaviour that you can opt-in to with a little extra work. This chapter will show you how.

```{r}
library(shiny)
```

## Basic idea

Let's take a simple app that we want to make bookmarkable. This app draws Lissajous figures which replicate the motion of a pendulum. There are two parameters to tweak, and having found an interesting pattern you might want to share it with a friend:

```{r}
ui <- fluidPage(
  sliderInput("omega", "omega", value = 0, min = -2, max = 2, step = 0.01),
  sliderInput("delta", "delta", value = pi / 2, min = 0, max = 2, step = 0.01),
  plotOutput("fig")
)
server <- function(input, output, session) {
  t <- seq(0, 100, length = 1e4)
  x <- reactive(sin(input$omega * t + input$delta))
  y <- reactive(sin(t))
  
  output$fig <- renderPlot({
    plot(x(), y(), axes = FALSE, xlab = "", ylab = "", type = "l", lwd = 2)
  })
}
```

There are three things we need to do to make this app bookmarkable:

1.  Add `bookmarkButton()` to the UI. This will generate a button that the 
    user clicks to generate the bookmarkable url.

1.  Turn `ui` into a function. This is needed because of the way that 
    bookmarkable state has to replay previous values: there's no longer really
    one UI, but many possible starting UIs depending on the previously 
    bookmarked values.
    
1.  Add `enableBookmarking = "url"` to the `shinyApp()` call.

This gives us:

```{r}
ui <- function(request) {
  fluidPage(
    sliderInput("omega", "omega", value = 1, min = -2, max = 2, step = 0.01),
    sliderInput("delta", "delta", value = pi / 2, min = 0, max = 2, step = 0.01),
    plotOutput("fig"),
    bookmarkButton()
  )
}
```
```{r, eval = FALSE}
shinyApp(ui, server, enableBookmarking = "url")
```

### Updating the url

Instead of providing an explicit button, another options is to automatically update the url in the browser. This means that the usual bookmark action will work, and you can copy and paste the current url.

```{r}
server <- function(input, output, session) {
  t <- seq(0, 100, length = 1e4)
  x <- reactive(sin(input$omega * t + input$delta))
  y <- reactive(sin(t))
  
  output$fig <- renderPlot({
    plot(x(), y(), axes = FALSE, xlab = "", ylab = "", type = "l", lwd = 2)
  })
  
  # Automatically bookmark every time an input changes
  observe({
    reactiveValuesToList(input)
    session$doBookmark()
  })
  # Update the query string
  onBookmarked(updateQueryString)
}
```
```{r, eval = FALSE}
shinyApp(ui, server, enableBookmarking = "url")
```

You could of course now remove the bookmark button if you were so inclined.

### Storing state

To bookmark an app means to store the current state of the inputs so it can be later replayed. There are ways you can do this:

* `enableBookmarking = "url"`, which we've used so far. This stores the state
  directly in the URL. This has the advantage of being very simple. The primary
  downside is that if you have many or complex inputs, it will create very 
  long URLs. This limitation also means that it can't work with file uploads.
  
* `enableBookmarking = "server"`, saves the state to an `.rds` file on the 
  server. This always generates a short URL but requires additional storage 
  on the server. While your testing locally you'll see these files saved in 
  `shiny_bookmarks/` in the working directory. Doesn't work on shinyapps.io.

## Challenges

When the application is restored using that state, the inputs are seeded with the saved values. This will replay all reactive expressions and outputs, yielding the same app that you see, as long as the application’s reactive graph structure is straightforward. 

* If your app uses random numbers, by default the random numbers will be 
  different between the apps. You can work around this by either using
  `set.seed()` to set the global seed or using `repeatable()` to set the
  random seed locally. 

* If you have tabs and want to bookmark and restore the active tab,
  make sure to supply an `id` in your call to `tabsetPanel()`.

* If there are inputs that should not be bookmarked, for example because 
  they're private information that should be shared include a called to 
  `setBookmarkExclude()` somewhere in your server function. For example,
  `setBookmarkExclude(c("secret1", "secret2"))` will ensure that the
  `secret1` and `secret2` inputs are not bookmarked.

* If you are manually managing reactive state in your own `reactiveValues()`
  object (as we'll discuss in Chapter XYZ), you'll need to use the 
  `onBookmark()` and `onRestore()` callbacks to manually save and load your 
  additional state. See 
  <https://shiny.rstudio.com/articles/advanced-bookmarking.html> for more 
  details.
