# Functions {#action-functions}

```{r, include = FALSE}
source("common.R")
```

When you start out writing an app, it's easy to get into the habit of solving related problems through copy and paste. Copy and paste is a great starting technique but if you rely on it too much you end up with two major problems with your app:

* It's hard to understand because it's hard to see the differences in blocks 
  of repeated code.
  
* It's hard to update because it's whenever you change one repeated block, you
  have to remember to change every other block in the same way.

At some point, you need to switch from copy and paste to writing functions. A good rule of thumb is that if you have copy and pasted something three times, you should write a function to reduce the duplication. In this chapter, I'll show how to use functions to reuse duplication in your UI and server. 

There are four main advantages to using a function. You can:

* Give it an informative name
* Isolate code
* Put it in a separate file
* Combine with functional programming techniques

If you're newer to writing functions, start by writing functions that extract out existing repeated patterns in your code. As you get more experience, you can begin to anticipate future needs.

I assume that you're already familiar with the basics of writing functions in R.  If not, you might want to read the [Functions chapter](https://r4ds.had.co.nz/functions.html) of _R for Data Science_ before continuing. The of this chapter goal is to activate your existing function skills, reminding you that Shiny code is R code, and then show some specific examples of how you might use functions in your Shiny app.

This chapter focusses on independent duplication in either the UI or server function. As your apps get more complex you'll also discover cases where the code in the UI is bound to the code in your server function. To solve that problem, you'll need modules, the topic of Chapter \@ref(scaling-modules).

```{r setup}
library(shiny)
```

## UI functions

Let's start with a concrete example of some duplicated code. For example, imagine that you're creating a bunch of sliders that need to each run from 0 to 1, starting at 0.5, with a 0.1 step. You _could_ do a bunch of copy and paste to generate all the sliders:

```{r}
ui <- fluidRow(
  sliderInput("alpha", "alpha", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("beta",  "beta",  min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("gamma", "gamma", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("delta", "delta", min = 0, max = 1, value = 0.5, step = 0.1)
)
```

But even for this simple case, I think it's worthwhile to pull worthwhile to recognise the repeated pattern and extract out a function:

```{r}
sliderInput01 <- function(id, label = id) {
  sliderInput(id, label, min = 0, max = 1, value = 0.5, step = 0.1)
}
```

Then the UI specification becomes much more concise:

```{r}
ui <- fluidRow(
  sliderInput01("alpha"),
  sliderInput01("beta"),
  sliderInput01("gamma"),
  sliderInput01("delta")
)
```

Functions provide a few benefits even in this simple case:

* We can give the function a evocative name, making it easier to understand 
  what's going on when we re-read the code in the future.
  
* If we need to change the behaviour (e.g. maybe we want to make the step size
  smaller), we only need to do it in one place

### Own file

Another advantage of functions is that they don't need to live in the same `app.R` file as your `ui` and `server()`. For large complex functions, I recommend putting them in `R/{function-name}.R`, but you might collect simple functions like `sliderInput01()` in a `R/utils.R` or similar. If you're using Shiny 1.5.0, running the app will automatically load all files in the `R/` directory, so there's nothing else to do!

For more complicated apps, particularly apps that multiple people contribute to, there are substantial advantages to turning your app into a package. In that case, you might want to check out the [golem](https://thinkr-open.github.io/golem/) package and accompanying ["Buidling Big Shiny Apps"](https://thinkr-open.github.io/building-shiny-apps-workflow/) book. I think a particularly important power is that the package infrastructure makes it possible to test your helper functions.

### Functional programming

If you're comfortable with functional programming, you could reduce the code still further:

```{r}
library(purrr)

vars <- c("alpha", "beta", "gamma", "delta")
sliders <- map(vars, sliderInput01)
ui <- fluidRow(!!!sliders)
```

There are two big ideas here:

* `map()` calls `sliderInput01()` once for each string stored in `vars`. 
  It returns a list of sliders.
  
* `!!!` __splats__ the elements `sliders` into individual arguments to 
  `fluidRow()`. `fluidRow(!!!list(a, b))` is equivalent to `fluidRow(a, b)`.

If you're like to learn more about `map()` (or it's base equivalent, `lapply()`), you might enjoy the [Functionals chapter](https://adv-r.hadley.nz/functionals.html) of _Advanced R_.

Don't worry if this code looks like gibberish to you: you can continue to use copy and paste instead. But in the long-run, I'd recommend learning more about functional programming, because it gives you such a wonderful ability to concisely express otherwise long-winded concepts.

### Varying arguments

It's possible to generalise this further if the controls have more than one varying input. First, we create an inline data frame that defines the parameters of each control, using `tibble::tribble()`:

```{r}
vars <- tibble::tribble(
  ~ id,   ~ min, ~ max,
  "alpha",     0,     1,
  "beta",      0,    10,
  "gamma",    -1,     1,
  "delta",     0,     1,
)
```

Then we create a function where the argument names match the column names:

```{r}
mySliderInput <- function(id, label = id, min = 0, max = 1) {
  sliderInput(id, label, min = 0, max = 1, value = 0.5, step = 0.1)
}
```

Then finally we use `purrr::pmap()` to call `mySliderInput()` once for each row of `vars`:

```{r}
sliders <- pmap(vars, mySliderInput)
```

### More advanced

See Section \@ref(programming-ui) for more examples of using these techniques to generate dynamic UI in response to user actions.

## Server functions

The key benefits of a function in the UI tend to be around reducing duplication. The key benefits of functions in a server tend to be around isolation and testing. When looking at a reactive expression or output, there's no way to easily tell exactly what values it depends on, except by carefully reading the code block. The function definition is a nice signpost that tells you  exactly what to inspect.  

Even without repetition, it's sometimes useful to pull complex code out of a reactive, and move it into a function. Take this server from Section \@ref(uploading-data). It contains a moderately complex `reactive()`:

```{r}
server <- function(input, output, session) {
  data <- reactive({
    req(input$file)
    
    ext <- tools::file_ext(input$file$name)
    switch(ext,
      csv = vroom::vroom(input$file$datapath, delim = ","),
      tsv = vroom::vroom(input$file$datapath, delim = "\t"),
      validate("Invalid file; Please upload a .csv or .tsv file")
    )
  })
  
  output$head <- renderTable({
    head(data(), input$n)
  })
}
```

If this was a real app, I'd seriously considering extracting out a function specifically for reading uploading files into its own function:

```{r}
load_file <- function(name, path) {
  ext <- tools::file_ext(name)
  switch(ext,
    csv = vroom::vroom(path, delim = ","),
    tsv = vroom::vroom(path, delim = "\t"),
    validate("Invalid file; Please upload a .csv or .tsv file")
  )
}
```

When extracting out such helpers, where possible avoiding taking reactives as input or returning outputs. Instead, pass them in through the arguments, and assume the caller will turn into a reactive. This isn't a hard and fast rule; sometime it will make sense for your functions to input or output reactives. But generally, I think it's better to keep the reactive and non-reactive parts of your app as separate a possible. In this case, I'm still using `validate()`; that works because outside of Shiny `validate()` works similar to `stop()`. But I keep the `req()` in the server, because it shouldn't the be responsibility of the file parsing code to know when it's run.

Since this is now an indepedent function, it could live in its own file (`R/load_file.R`, say), keeping the `server()` svelte. This helps keep the server function focussed on the big picture
of reactivity, rather than the smaller details underlying each component.

```{r}
server <- function(input, output, session) {
  data <- reactive({
    req(input$file)
    load_file(input$file$name, input$file$datapath)
  })
  
  output$head <- renderTable({
    head(data(), input$n)
  })
}
```

The other big advantage is that you can play with `load_file()` at the console, outside of your Shiny app. If you move towards formal testing of your app (see Chapter XXX), this also makes that code easier to test.

## Case studies
