# Functions {#action-functions}

```{r, include = FALSE}
source("common.R")
```

If you have been creating a lot of your app via copy and paste, parts of your app may be very similar to each other. Copy and paste is a great starting technique but if you rely on it too much you end up with two major problems:

* Your app is hard to understand because it's hard to see the important 
  differences when you have a lot of copy and pasted code.
  
* Your app is hard to update because it's easy to forget to update one of the 
  places that you've duplicated code). 
  
A good rule of thumb is that if you have copy and pasted something three times, it's a good time use some other technique to reduce the amount of duplication. In this chapter, we'll show how you can use functions to reuse duplication in code on either the UI or the server. This chapter is fairly brief; the goal is to activate your existing function skills and remind you that Shiny code is R code.

There are four main advantages to using a function. You can:

* Put it in a separate file
* Give it an informative name
* Isolate code
* Combine with functional programming techniques

This isn't the only technique you'll need in Shiny, but it's a good place to start. In Chapter \@ref(scaling-modules), you'll learn about __modules__, which allow you to reduce duplicated code that requires coordination across the UI and server.

```{r setup}
library(shiny)
```

## Organisation

Once you've extracted your code out into a function, you can put that code in another file.

```{r, eval = FALSE}
library(shiny)

source("R/my-helper-function.R")

server <- function(input, output, session) {
  data <- reactive(my_helper(input$x, input$y, input$z))
}
```

(A future version of shiny will automatically source all files in `R/`, <https://github.com/rstudio/shiny/pull/2547>, so you'll be able to remove the `source()` line.)

As your collection of helper functions grow, you might want to pull them out into their own files. I recommend putting that file in a `R/` directory underneath the app directory. Then load it at the top of your `app.R`:

For more complicated apps, particularly apps that multiple people contribute to, there are substantial advantages to turning your app into a package. In that case, you might want to check out the [golem](https://thinkr-open.github.io/golem/) package and accompanying ["Buidling Big Shiny Apps"](https://thinkr-open.github.io/building-shiny-apps-workflow/) book. I think a particularly important power is that the package infrastructure makes it possible to test your helper functions; and they can be tested independently of reactivity.

If you're newer to writing functions, start by writing functions that extract out existing repeated patterns in your code. As you get more experience, you can begin to anticipate future needs.

## UI

You can apply these same ideas to generating your UI. If you have a bunch of controls that you use again and again and again, it's worth doing some up front work to make a function that saves some typing.

This can be useful even if all you're doing is changing three or four default arguments. For example, imagine that you're creating a bunch of sliders that need to each run from 0 to 1, starting at 0.5, with a 0.1 step. You _could_ do a bunch of copy and paste:

```{r}
ui <- fluidRow(
  sliderInput("alpha", "alpha", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("beta", "beta", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("gamma", "gamma", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("delta", "delta", min = 0, max = 1, value = 0.5, step = 0.1)
)
```

But even for this simple case, I think it's worthwhile to pull out the repeated code into a function:

```{r}
sliderInput01 <- function(id, label = id) {
  sliderInput(id, label, min = 0, max = 1, value = 0.5, step = 0.1)
}

ui <- fluidRow(
  sliderInput01("alpha"),
  sliderInput01("beta"),
  sliderInput01("gamma"),
  sliderInput01("delta")
)
```

### Functional programming

If you're comfortable with functional programming, you could reduce the code still further:

```{r}
library(purrr)

vars <- c("alpha", "beta", "gamma", "delta")
sliders <- vars %>% map(sliderInput01)
ui <- fluidRow(!!!sliders)
```

There are two big ideas here:

* Using a function like `map()` to generate a list containing one slider for 
  each element of `vars`.
  
* Using `!!!` to __splat__ the elements of the list into individual arguments
  to `fluidRow()`. `fluidRow(!!!list(a, b))` is equivalent to `fluidRow(a, b)`.

I'm not going to teach functional programming in this book, but I will show a few aspirational examples. It's a good example of where improving your general R programming skills pays off in your Shiny apps. 

### Varying arguments

```{r}
vars <- tibble::tribble(
  ~ id,   ~ min, ~ max,
  "alpha",     0,     1,
  "beta",      0,    10,
  "gamma",    -1,     1,
  "delta",     0,     1,
)

mySliderInput <- function(id, label = id, min = 0, max = 1) {
  sliderInput(id, label, min = 0, max = 1, value = 0.5, step = 0.1)
}

sliders <- vars %>% pmap(mySliderInput)
```

Don't worry if this all looks like gibberish: you can just use copy and paste instead. 

## Server

<!-- https://community.rstudio.com/t/r-shiny-apply-custom-function-to-datatable/39790/3 -->

If, given specific values, your app requires complex calculation, first start by pulling that calculation out into separate function:

```{r}
server <- function(input, output, session) {
  data <- reactive({
    # complex data calculation involving input$x, input$y, input$z
  })
}
```

```{r}
my_helper <- function(x, y, z) {
  ...
}

server <- function(input, output, session) {
  data <- reactive(my_helper(input$x, input$y, input$z))
}
```

When extracting out such helpers, avoid putting any reactive component inside the function. Instead, pass them in through the arguments. 

Reactivity should only live in the server function, or a module server function, which we'll talk about next.

A function also _enforces_ this independence --- if you try and refer to an input that you did not pass into the function, you'll get an error. This enforced independence becomes increasingly important as you create bigger and bigger apps because it ensures that pieces of your app are independent and can be analysed in isolation. 

When looking at a reactive expression or output, there's no way to easily tell exactly what values it depends on, except by carefully reading the code block. The function definition is a nice signpost that tells you  exactly what to inspect.  

### Functional programming

If you're a functional programming aficionado, you could even generate the call to `dplyr::filter()`:

```{r}
library(rlang)
select <-  map(vars$var, function(v) expr(.data[[!!v]] == input[[!!v]]))
select
```

If you haven't seen `.data` before, it comes from tidy evaluation, the system that allows you to program with tidyverse packages that are designed for interactive exploration (like dplyr). It's not necessary when writing interactive code (and it's not strictly necessary here) but it makes the parallel between the data frame and the inputs more clear. We'll talk more about tidy evaluation in Chapter \@ref(action-tidy).

Again we'd use `!!!` to splat the generated expressions into `filter()`:

```{r, eval = FALSE}
filter(ngo, !!!select)
```
