# Modules {#action-modules}

```{r, include = FALSE}
source("common.R")
```

As your app grows, sandwhiching all of your code into a single `app.R` will start to become painful. This chapter describes the techniques you can use to break your app apart into independent pieces, focussing on shiny modules.

If you learend Shiny with an older version of Shiny, you might be more familiar with using separate files for the front end (`ui.R`) and back end (`server.R`). That organisation continues to work, but is no longer recommended: if you have an older app, I recommend doing a little copy and paste to combine the two files into a single `app.R`. Similarly, if you're using `global.R` inline it into `app.R`.

DRY principle. Rule of three.

## Helper functions

If, given specific values, your app requires complex calculation, first start by pulling that calculation out into separate function, and then put that function in its own file. I recommend put that file in a `R/` directory underneath the app directory. Then load it at the top of your `app.R`:

```{r, eval = FALSE}
library(shiny)

source("R/my-helper-function.R")
ui <- fluidPage(
  ...
)
```

(A future version of shiny will automatically source all files in `R/`, <https://github.com/rstudio/shiny/pull/2547>, so you'll be able to remove the `source()` line.)

## Reducing duplication with functions

Sometimes you can extract out duplicated code using functions. For example, if you've copied and pasted some UI code to create variants with different names:

Or you have a self contained set of reactives:

However, a function alone with only take you so far because typically you'll have some connection between the front end and back end, and you need some way to coordinate the two. Shiny uses identifiers so you need some way to share them. This gives rise to Shiny __modules__.

### Case study

<!-- https://community.rstudio.com/t/trying-to-create-a-reactive-graph-that-will-take-in-multiple-different-inputs/38506 -->

Repeated code makes differences hard to see.

```{r eval = FALSE}
fluidRow(
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("traffickingType",
      label = "Choose a trafficking type: ",
      choices = sort(unique(ngo$Trafficking.Type)),
      multiple = TRUE
    )
  ),
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("traffickingSubType",
      label = "Choose a trafficking sub type: ",
      choices = sort(unique(ngo$Trafficking.Sub.Type)),
      multiple = TRUE
    )
  ),
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("gender",
      label = "Choose a gender: ",
      choices = sort(unique(ngo$Victim.Gender)),
      multiple = TRUE
    )
  )
)
```

```{r, eval = FALSE}
ngoSelectInput <- function(var, label, multiple = TRUE) {
  choices <- sort(unique(ngo[[var]]))
  label <- paste0("Choose a ", label, ": ")
  selectInput(var, label, choices = choices, multiple = multiple)
}
boxHeader <- function(...) {
  box(width = 4, solidHeader = TRUE, ...)
}

fluidRow(
  boxHeader(ngoSelectInput("Trafficking.Type", "trafficking type")),
  boxHeader(ngoSelectInput("Trafficking.Sub.Type", "trafficking sub type")),
  boxHeader(ngoSelectInput("Victim.Gender", "gender"))
)
```

I made one simplifying assumption that would also require changes on the server side: when filtering based on a variable, the input name should be the same as the variable name. I think this sort of consistency generally makes for code that's easier to read and remember.

```{r}
ngo_filtered <- reactive({
  filter(ngo, 
    Trafficking.Type %in% input$Trafficking.Type,
    Trafficking.Sub.Type %in% input$Trafficking.Sub.Type,
    Victim.Gender %in% input$Victim.Gender
  )
})
```

If you had a lot more controls, I'd consider using functional programming to generate them. I won't talk about more here, but you might want to learn more about it if this is intriguing to you. 

```{r, eval = FALSE}
library(purrr)
vars <- tibble::tribble(
  ~ var,                  ~ label,
  "Trafficking.Type",     "trafficking type",
  "Trafficking.Sub.Type", "trafficking sub type",
  "Victim.Gender",        "gender"
)
vars %>% 
  pmap(ngoSelectInput) %>% # create one select input for each row
  map(boxHeader) %>%       # wrap each in a boxHeader()
  exec(fluidRow)           # collapse into a single fluidRow()
```

In this case generating the dplyr code would require yet still more FP skills:

```{r, eval = FALSE}
library(rlang)
select <-  map(vars$var, function(v) expr(!!sym(v) == input[[!!v]]))
select
filter(ngo, !!!select)
```

Point is that if you already have these skils, don't forget that you can apply them with Shiny!

## Isolating behaviour with modules

Modules are way to create an app within an app. They force isolation of behaviour so that one module can't affect another, and code outside of a module can only affect the inside in a way that the module allows.

A Shiny module is a pair of functions, corresponding to the front end UI and the backend server function.

### Front end

The front end function has one special convention: the first argument must be `id`, and it uses `shiny::NS()` to generate unique names within the function. `NS()` is what's known as a function factory: it returns a function that you then use to generate unique identifiers. You can see it in operation below:

```{r}
ns <- NS("welcome")
ns("name")
ns("age")
```

The key idea is that the argument to `NS()` is supplied by the person using the component, and the arguments to the function it produces is supply by the person who wrote the component. This two-phase creation makes ensures that the final name combines properties needed by both the app author and the module author. This is a bit confusing at first, because you're likely to be both the app and module author.

`NS()` is so simple that you might wonder why you can't just implement it yourself. You certainly can, but using `NS()` makes it immediately obvious that you're generating a module.

### Back end

The back end looks like a regular server function. You don't need to worry about namespacing as Shiny takes care of that when you insert the module in your function.

## Developing a module

```{r, eval = FALSE}
ui <- fluidPage(
  my_module_ui("foofy")
)
server <- function(input, output, session) {
  callModule(my_module, "foofy")
}
```

## Case studies

### Limited selection + other

Consider the following app, which provdies a way to select gender that is sensitive to the many possible ways that people can express their gender.[^gender]

[^gender]: For a deeper dive on this issue, and a discussion of why many commonly used way of asking about gender can be hurtful to some people, I recommend reading "Designing forms for gender diversity and inclusion" by Sabrina Fonseca: <https://uxdesign.cc/d8194cf1f51>.

```{r}
ui <- fluidPage(
  radioButtons("gender", "Gender:",
    choiceValues = list("male", "female", "self-described", "na"),
    choiceNames = list(
      "Male",
      "Female",
      textInput("gender_self", NULL, placeholder = "Self-described"),
      "Prefer not to say"
    ),
    selected = "na",
  ),
  textOutput("txt")
)

server <- function(input, output, session) {
  observeEvent(input$gender_self, {
    req(input$gender_self)
    updateRadioButtons(session, "gender", selected = "self-described")
  })
  
  gender <- reactive({
    if (input$gender == "self-described") {
      input$gender_self
    } else {
      input$gender
    }
  })
  
  output$txt <- renderText({
    paste("You chose", gender())
  })
}
```

The `gender` and `gender_self` components are tightly bound together. We haven't talked about `updateRadioButtons()` yet, but this is just a small convenience so that if you start typing a self-described gender, that radio button is automatically selected.

Convert to a module and generalise a little.

```{r}
radioButtonsWithOther <- function(id, label, choices, selected = NULL, placeholder = NULL) {
  ns <- NS(id)
  
  radioButtons(ns("primary"), "Gender:",
    choiceValues = c(names(choices), "other"),
    choiceNames = c(
      unname(choices),
      list(textInput(ns("other"), NULL, placeholder = NULL))
    ),
    selected = selected
  )
}

radioButtonsWithOtherServer <- function(input, output, session) {
  observeEvent(input$primary, {
    req(input$other)
    updateRadioButtons(session, "primary", selected = "other")
  })
  reactive({
    if (input$primary == "other") {
      input$other
    } else {
      input$primary
    }
  })
}

ui <- fluidPage(
  radioButtonsWithOther("gender", 
    label = "Gender", 
    choices = list(
      male = "Male",
      female = "Female",
      na = "Prefer not to say"
    ), 
    placeholder = "Self-described", 
    selected = "na"
  ),
  textOutput("txt")
)

server <- function(input, output, session) {
  gender <- callModule(radioButtonsWithOtherServer, "gender")

  output$txt <- renderText({
    paste("You chose", gender())
  })
}
```

### Hierarchical select boxes

```{r, eval = FALSE}
library(tidyverse)
country_df <- countrycode::codelist %>% 
  as_tibble() %>% 
  select(iso3c, continent, country = cow.name) %>% 
  filter(!is.na(continent), !is.na(country))

continents <- sort(unique(country_df$continent))

ui <- fluidPage(
  selectInput("continent", "Continent", choices = continents),
  selectInput("country", "country", choices = NULL)
)

server <- function(input, output, session) {
  countries <- reactive({
    country_df[country_df$continent == input$continent, , drop = FALSE]
  })
  
  observeEvent(input$continent, {
    updateSelectInput(session, "country", choice = countries()$country)
  })
}
shinyApp(ui, server)

```


## Packages

For more complicated apps, particularly apps that multiple people contribute to, there are substantial advantages to turning your app into a package. In that case, you might want to check out the [golem](https://thinkr-open.github.io/golem/) package and accompanying ["Buidling Big Shiny Apps"](https://thinkr-open.github.io/building-shiny-apps-workflow/) book.
