# User feedback {#action-feedback}

```{r, include = FALSE}
source("common.R")
source("demo.R")
```

It's important to give the user feedback as they interact with your app. Some feedback occurs naturally through outputs, but often you'll need a side-channel to let the user know that they've given the app an invalid input or its taking a little while for some action to occur. In this chapter, you'll learn how to keep the user up to date with what's happening with validation message and notifications. Many of the techniques described here are best thought of as side-effects; they are things that happen at the end of the reactive graph.

* Validation
* Notification
* Progress bars
* Dialog boxes

```{r setup}
library(shiny)
```

## Validation

The first and most important feedback you can give to the user is that they've given you bad input. 

It's particularly important to think through how the user might use your app incorrectly so that you can give them informative messages in the UI, rather than allowing errors to trickle through into the R code where the error messages might not be so important.

In this section, you'll learn about the `req()` and `validate()` functions.

### Ignore missing inputs

Sometimes you don't want to do anything until the user has performed some action. By default, a reactive will perform computation as soon as the app loads, but sometimes you want to wait until the user has provided some data.

This occurs primarily with three controls:

* `textInput()` starts off blank (`""`) so you may want to delay computation
  until the user has entered some text.
  
* In `inputSelect()` you may have provided an empty choice, `""`, to indicate
  no selection.
  
* `fileInput()` will have an empty result until the user has uploaded a file.
  See Section \@ref(#upload) for more details.

The easiest way to fix this problem is with `req()`: it checks that the input has been set before proceeding. `req()` sends a special signal to Shiny telling it that the reactive isn't ready yet. For example, consider the following app which will generate a greeting in English or Maori. If you run this app, you'll see an error, as in Figure \@ref(fig:require-simple). That's because there's no entry in the `greetings` vector that corresponds to the choice of `""`.


```{r}
ui <- fluidPage(
  selectInput("language", "Language", choices = c("", "English", "Maori")),
  textInput("name", "Name"),
  textOutput("greeting")
)

server <- function(input, output, session) {
  greetings <- c(
    English = "Hello", 
    Maori = "Ki ora"
  )
  output$greeting <- renderText({
    paste0(greetings[[input$language]], " ", input$name, "!")
  })
}
```

```{r require-simple, echo = FALSE, message = FALSE, cache = FALSE, fig.cap = "The app displays an uninformation error when it is loaded because language hasn't been selected yet", out.width = "50%", fig.show = "hold"}
demo <- demo_inline("require-simple", ui, server)
demo$resize(400)
demo$screenshot()
```

You can fix this problem by using `req()`. Now nothing will be displayed until the user has suppled values for both language and name, as shown in Figure \@ref(fig:require-simple2).

```{r}
server <- function(input, output, session) {
  greetings <- c(
    English = "Hello", 
    Maori = "Ki ora"
  )
  output$greeting <- renderText({
    req(input$language, input$name)
    paste0(greetings[[input$language]], " ", input$name, "!")
  })
}
```

```{r require-simple2, fig.cap = "By using `req()`, the output is only shown once both language and name have been supplied", out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("require-simple2", ui, server)
demo$resize(400)
s1 <- demo$screenshot("on-load")
demo$set_values(language = "English")
s2 <- demo$screenshot("langauge")
demo$set_values(name = "Hadley")
s3 <- demo$screenshot("name")
knitr::include_graphics(c(s1, s2, s3))
```

`req()` is designed so that `req(input$x)` should just work regardless of what type of input `x` is: if you're interested you can read about the details in `?isTruthy`. You can also use `req()` with your own logical statement if needed. For example, `req(input$a > 0)` will only trigger computation when a is greater than 0.

In certain cases, you might want to leave the last output up. For example, imagine you allow the user to type the name of a dataset. As they're typing, reactives will be firing, and it'd be nice to only change the output once they've typed a complete name.

```{r}
ui <- fluidPage(
  textInput("dataset", "Dataset name"), 
  tableOutput("data")
)
server <- function(input, output, session) {
  output$data <- renderTable({
    req(input$dataset)
    req(exists(input$dataset, "package:datasets"), cancelOutput = TRUE)
    head(get(input$dataset, "package:datasets"))
  })
}
```

Also note that I've used `req()` twice; the first time prevents `exists("")` being called, which triggers an error.

`req()` works by signalling a special condition (condition is the term that refers jointly to errors, warnings, and messages). This special condition causes all downstream reactives and outputs to stop executing. Technically, it leaves any downstream reactive consumers in an invalidated state. We'll come back to this terminology in Chapter \@ref{reactive-components}.

### Validate input

If you want to give additional feedback to the user, a great way to do so is with the [shinyFeedback](https://github.com/merlinoa/shinyFeedback) package. There are two steps to use it. First, you add `useShinyFeedback()` to the ui:

```{r}
ui <- fluidPage(
  shinyFeedback::useShinyFeedback(),
  
  textInput("dataset", "Dataset name"), 
  tableOutput("data")
)
```

Then you call `feedbackDanger()` or `feedbackWarning()` in the server function. Since the logic is starting to get complicated, I've pulled out the validation into a reactive. The sequence of operations is basically the same as before except now that I pull out `exist` into a separate variable and use it for both `feedbackDanger()` and `req()`.

```{r}
server <- function(input, output, session) {
  data <- reactive({
    req(input$dataset)
    
    exists <- exists(input$dataset, "package:datasets")
    shinyFeedback::feedbackDanger("dataset", !exists, "Unknown dataset")
    req(exists, cancelOutput = TRUE)

    get(input$dataset, "package:datasets")
  })
  
  output$data <- renderTable({
    head(data())
  })
}
```

Note that `feedbackDanger()` is parameterised in the opposite direction to `req()` so it needs a `!` on its condition.

### Validate output

Sometimes the problem is not related to a single input, but is related to an invalid combination of inputs. In this case, rather than putting the error message next to the input, it might make more sense to put it in the output. You can do so with a tool built into shiny: `validate()`. When called inside a reactive or a output, `validate(message)` stops execution of the rest of the code and instead displays `message` in any downstream outputs.

```{r}
ui <- fluidPage(
  numericInput("x", "x", value = 0),
  selectInput("trans", "transformation", choices = c("square", "log", "square-root")),
  textOutput("out")
)

server <- function(input, output, server) {
  output$out <- renderText({
    if (input$x < 0 && input$trans %in% c("log", "square-root")) {
      validate("x can not be negative for this transformation")
    }
    
    switch(input$trans,
      square = input$x ^ 2,
      "square-root" = sqrt(input$x),
      log = log(input$x)
    )
  })
}
```

## Notifications

In other cases, their might not be a problem, but you just want to let the user know what's happening, particularly if you need to perform some operation that takes a while. You can use `showNotification()` to display a notification dialog. Notifications automatically automatically stack in the bottom right corner of the page.

### Transient notification

The simplest way to use `showNotification()` is to call it with the message you want to display to the user:

```{r}
ui <- fluidPage(
  actionButton("night", "Good night")
)
server <- function(input, output, session) {
  observeEvent(input$night, {
    showNotification("So long")
    Sys.sleep(1)
    showNotification("Farewell")
    Sys.sleep(1)
    showNotification("Auf Wiedersehen")
    Sys.sleep(1)
    showNotification("Adieu")
  })
}
```

By default, the message will disappear after 5 seconds, which you can override by setting `duration`. The user can also dismiss earlier by clicking on the close button. If you want to make the notification more prominent, you can set the `type`, to one of "message", "warning", or "error".

### Removing on completion

If you have a longer running action that you'd like to notify the user about, you can manually remove the notification once the action has completed. In this case you'll need:

* `duration = NULL` and `closeButton = FALSE` so that the notification 
  stays visibile until the data loading is done.
  
* To store `id` returned by `showNotification()`, and then call 
  `removeNotification()` to remove it once it's done. The most reliable way to 
  do that is to wrap it in `on.exit()`, which ensures that the notification is 
  removed regardless of how the reactive completes. It also has the nice property 
  that you can put the show and remove functions right next to each other.

The following example puts the pieces together to shows how you might keep the user up to date when reading in a large csv file[^csv]:

[^csv]: For this specific case you should also look at `data.table::fread()` and `vroom::vroom()` to read in the file; they can be orders of magnitude faster than `read.csv()`.

```{r}
server <- function(input, output, session) {
  data <- reactive({
    id <- showNotification("Reading data...", duration = NULL, closeButton = FALSE)
    on.exit(removeNotification(id), add = TRUE)
    
    read.csv(path())
  })
}
```

### Progressive updates

By default, multiple calls to `showNotification()` will create multiple notifications on the notification stack. You can update a single notification by capturing the `id` from the first call and using it in subsequent calls. 

```{r}
ui <- fluidPage(
  tableOutput("data")
)

server <- function(input, output, session) {
  notify <- function(msg, id = NULL) {
    showNotification(msg, id = id, duration = NULL, closeButton = FALSE)
  }

  data <- reactive({ 
    id <- notify("Reading data...")
    on.exit(removeNotification(id), add = TRUE)
    Sys.sleep(1)
      
    notify("Reticulating splines...", id = id)
    Sys.sleep(1)
    
    notify("Herding llamas...", id = id)
    Sys.sleep(1)

    notify("Notifying sherpas...", id = id)
    Sys.sleep(1)
        
    mtcars
  })
  
  output$data <- renderTable(head(data()))
}
```

## Progress bars

For long-running tasks, the best type of feedback is a progress bar. As well as telling you where you are in the process, it also helps you estimate how much longer it's going to be: should you take a deep breath, go get a coffee, or come back tomorrow? 

In this section I'll show two techniques for displaying progress bars, one build into Shiny, and one from the [waiter](https://waiter.john-coene.com/) package developed by John Coene. Both work roughly the same way, first creating an R6 object and then calling an update method after each step.

Unfortunately both techniques suffer from the same drawbacks. To use a progress bar you need to be able to divide the a big task into a known number of small pieces that each take roughly the same amount of time. This is often hard, particularly since the underlying code is often written in C and you have no insight into what it's doing. We are working on tools in the progress package so that packages like dplyr, readr, and vroom will generate progress bars that you can easily forward to Shiny, and I'll update this chapter with an example when that technology is mature

### Shiny

The following code block shows the basic lifecycle of a progress bar:

```{r, eval = FALSE}
progress <- Progress$new(max = 5)
progress$set(message = "Starting process")

for (i in 1:5) {
  progress$inc(1)
}

progress$close()
```

1.  Create a progress bar object with `Progress$new(max = number_of_steps)`.

1.  Display the progress bar by calling the `$set()` message , providing the a 
    title for the progress bar in the `message` argument.

1.  Call `$inc()` repeatedly, once for each step.

1.  When done, call `$close()` to terminate the progress bar.

And here's how that might look in a complete Shiny app:

```{r}
ui <- fluidPage(
  numericInput("steps", "How many steps?", 10),
  actionButton("go", "go")
)
server <- function(input, output, session) {
  observeEvent(input$go, {
    progress <- Progress$new(max = input$steps)
    on.exit(progress$close())
    
    progress$set(message = "Starting")
    for (i in seq_len(input$steps)) {
      Sys.sleep(0.5)
      progress$inc(1)
    }
  })
}
```

Here I'm using `Sys.sleep()` to simulate a long running operation; in your code this would be a moderately expensive operation.

### Waiter

[waiter](https://waiter.john-coene.com/)

Adapating the above code to work with Waiter is straightforward. In the UI, we add `use_waitress()`:

```{r}
library(waiter)

ui <- fluidPage(
  use_waitress(),
  numericInput("steps", "How many steps?", 10),
  actionButton("go", "go")
)
```

And then we replace `Progress` with `Waitress`:

```{r}
server <- function(input, output, session) {
  observeEvent(input$go, {
    waitress <- Waitress$new()
    waitress$start()
    on.exit(waitress$hide())
    
    for (i in seq_len(input$steps)) {
      Sys.sleep(0.5)
      waitress$increase(1 / input$steps * 100)
    }
  })
}
```

The neat thing about Waitress is that you can overlay it on an existing input or output.

```{r, eval = FALSE}
waitress <- Waitress$new(id = "#steps", theme = "overlay")
```

One very nice feature of waiter (indeed, the titular feature) is that if you don't know exactly how long the process will take you can use a spinner (a waiter) instead:

```{r, eval = FALSE}
server <- function(input, output, session) {
  observeEvent(input$go, {
    waiter <- Waiter$new(id = "steps")
    waiter$start()
    on.exit(waiter$hide())
    
    Sys.sleep(0.5 * input$steps)
  })
}
```

See the complete list of progreess animations at `?waiter::spinners`.

## Dialogs

Sometimes you need to give more information to the user than you can fit in a simple notification, or you want the user to take some action, or  or the information is so important that you want to force the user to read and respond.

This is called a "modal" dialog because it creates a new "mode" of interaction; you can't interact with the main application until you have dealt with the dialog.

### Simple

```{r}
delete_all <- modalDialog(
  "ALL FILES DELETED", 
  footer = NULL,
  easyClose = TRUE
)

modal_ui <- modalDialog(
  "This is really important!",
  "Are you sure you want to continue?",
  footer = list(
    actionButton("yes", "yes"),
    actionButton("no", "no")
  )
)
```

```{r}
ui <- fluidPage(
  actionButton("go", "Go")
)

server <- function(input, output, session) {
  observeEvent(input$go, {
    showModal(modal_ui)
  })
  
  observeEvent(input$yes, 
    showModal(delete_all)
  )
  observeEvent(input$no, 
    removeModal()
  )
}
```

### Complex 

A dialog box can contain any `UI` inside of it.

Another way of creating dynamic UI. See Chapter \@ref(action-dynamic) for more details.
