# Graphics {#action-graphics}

```{r, include = FALSE}
source("common.R")
```

We talked briefly about `renderPlot()` in Chapter \@ref(basic-ui); it's a powerful tool for displaying graphics in your app. This chapter will show you how to use it to its full extent, including how to make your plots interactive. You'll also learn two important related functions: `renderCachedPlot()`, which speeds up your app by caching frequently used plots, and `renderImage()`, which allows you to serve images created in other ways.

In this chapter, we'll show a few base graphics, but will focus on ggplot2.

```{r setup}
library(shiny)
library(ggplot2)
```

## Interactivity

One of the neatest things about `plotOutput()` is that it can also be used as an input, allowing you to create interactive graphics. This is a very powerful tool, and has a wide range of applications. Here I'll show the basics, and give you a few pointers about how you might extent for your own purposes.

There are four mouse events that you can respond to on a plot: `click`, `dblClick` (double click), `hover` (fired when the mouse stays in the same place for a little while), and `brush`. To turn these events into Shiny inputs, you supply a string to the corresponding argument, which then creates an input with that name. For example, imagine you wanted to create a plot that responded to click events. You'd set `plotOutput("plot", click = "plot_click")`. 

Here's a simple example that displays the location of the mouse click (note that the location is in data coordinates):

```{r}
ui <- basicPage(
  plotOutput("plot", click = "plot_click"),
  verbatimTextOutput("info")
)

server <- function(input, output) {
  output$plot <- renderPlot({
    plot(mtcars$wt, mtcars$mpg)
  })

  output$info <- renderPrint({
    req(input$plot_click)
    x <- round(input$plot_click$x, 2)
    y <- round(input$plot_click$y, 2)
    paste0("[", x, ", ", y, "]")
  })
}
```

(Note the use of `req()`; that makes sure the app doesn't do anything before the first click.)

The following sections describe the events in more details. We'll start with the "point" events, `click`, `dblClick` and `hover`, whose primary data is the `x` and  `y` location of the event (in various units). Then we'll discuss the `hover` event, which gives a rectangle defined by its four sides (`xmin`, `xmax`, `ymin`, and `ymax`). I'll give a brief example of using these events to maintain a logical selection vector; we'll come back to the underlying ideas later in Chapter XYZ.

If you're looking for more general interactivity, you might also want to consider the plotly package, as documented in the book [_Interactive web-based data visualization with R, plotly, and shiny_](https://plotly-r.com), by Carson Sievert.

### Points

The point events return a relatively rich data structure. The most components of the output `x` and `y` components that give the location of the event in data coordinates. But there are also:

* `coord_img` which give the raw pixel coordinates on the image.
* `coord_css` which uses `img_css_ratio` to compute the scaled css pixel
  coordinates on the image
* `mapping` is used for ggplot2 plots
* `domain` which gives the location of the data boundaries of the plot 
  (i.e. inside the axes) in data coordinates. `range` gives them in 
  terms of pixel coordinates on the image.
* `log` which tells you if either of the `x` or `y` axes are logged.

(If you want to learn more about these values, I think the best bet is to create an app that displays them so you can explore interactively.)

However, you'll only need to use the input directly in relatively rare situations because Shiny provides the `nearPoints()` helper, which finds data points near the event. You're best off using these helpers because they take care of a bunch of fiddly details. Here's a simple example of `nearPoints()` in action, showing a table of data about the points near the event: 

```{r}
ui <- fluidPage(
  plotOutput("plot", click = clickOpts("click")),
  tableOutput("data")
)
server <- function(input, output, session) {
  output$plot <- renderPlot(
    plot(mtcars$wt, mtcars$mpg)
  )
  
  output$data <- renderTable({
    nearPoints(mtcars, input$click, xvar = "wt", yvar = "mpg")
  })
}
```

Here `nearPoints()` takes four arguments: the data frame the underlies the plot, the input event, and the names of the `x` and `y` variables on the plot. If you use ggplot2 instead, you don't need to provide these arguments because they can be automatically imputed from the plot data structure. For that reason, I'll use ggplot2 throughout the rest of the chapter. Here's the previous Shiny app reimplemented with ggplot2:

```{r}
ui <- fluidPage(
  plotOutput("plot", click = "plot_click"),
  tableOutput("data")
)
server <- function(input, output, session) {
  output$plot <- renderPlot(
    ggplot(mtcars, aes(wt, mpg)) + geom_point()
  )
  
  output$data <- renderTable({
    nearPoints(mtcars, input$plot_)
  })
}
```

Another way to use `nearPoints()` is to compute the distance between each point and the event. `nearPoints(allRows = TRUE, addDist = TRUE)` to return all rows with distance; this allows you to rely on the distance calculations done by Shiny, but you can use the information in other ways. 


Same principle applies to `hover` and `dblClick`, it's only the user event that changes. You can get additional control over the events by supplying `clickOpts()`, `dblclickOpts()`, or `hoverOpts()` instead of a string giving the input id. These are rarely needed, so I won't discuss them hear; see the documentation for details.

### Brushing

Another way of selecting points on a plot is to use a **brush**, a rectangular selection on the plot. Once you've mastered click and `nearPoints()`, using the brush is just a matter of using the `hover` event and the `brushedPoints()` helper.

### Modifying data

So far we've displayed the results of the interaction in another output. But the true elegance of interactivty comes when you display the changes in the plot you are interacting with. Unfortuantely to achieve this, you'll need to use `reactiveValue()`, This creates a reactive object that behave just like an input, but you can modify in code (rather than relying on the user to modify in the browser). We're going to cover `reactiveValue()` in detail in Chapter XYZ, but I wanted to show it here because it's such a powerful technique. You might not understand everything with your first read through, but you'll hopefully see some sense of the potential and get a couple of techniques that you can copy and paste.

`reactiveValue()` works similarly to `reactive()`. It has one major difference: you can also update its value by supplying an argument when you call it. The following code shows the basic idea, but unfortunately it work work in the console because it must be run in an interactive environment. Working with `reactiveValues()` also requires you to improve your debugging skills.

```{r, eval = FALSE}
val <- reactiveVal(10)
val()
#> [1] 10
val(20)
val()
#> [1] 20
```

`reactiveValeu()` will almost always be paired with `observeEvent()`; you'll use this to update the reactive value when (e.g.) a mouse event occurs. For example, imagine that you want to visualise the distance between a click and the points on  the plot. First we create a reactive value that initialises the distance. Then, whenever the plot is clicked, we update this value using `nearPoints()`:

```{r, eval = FALSE}
dist <- reactiveVal(rep(1, nrow(df)))

observeEvent(input$plot_click,
  dist(nearPoints(df, input$plot_click, allRows = TRUE, addDist = TRUE)$dist_)  
)
```

Putting this all together, we get: 

```{r}
df <- data.frame(x = rnorm(100), y = rnorm(100))

ui <- fluidPage(
  plotOutput("plot", click = "plot_click")
)
server <- function(input, output, session) {
  dist <- reactiveVal(rep(1, 100))
  observeEvent(input$plot_click,
    dist(nearPoints(df, input$plot_click, allRows = TRUE, addDist = TRUE)$dist_)  
  )
  
  output$plot <- renderPlot({
    df$dist <- dist()
    ggplot(df, aes(x, y, size = dist)) + 
      geom_point() + 
      scale_size_area(limits = c(0, 1000), max_size = 10, guide = NULL)
  })
}
```

(Note the arguments to `scale_size_area()`: I've set the limits so that .  This is one of the biggest challenges in interactive graphics. Keeping plots comparable over time.)

Here's a more complicated idea. I'm going to allow the user to click points to add or remove them from a model fit. I use `nearPoint()` to find which points are near the click, then `ifelse()` to toggle their values: if they were previously excluded they'll be included; if they were previously included, they'll be excluded.

```{r, eval = FALSE}
selected <- reactiveVal(rep(TRUE, nrow(mtcars)))

observeEvent(input$click, {
  clicked <- nearPoints(mtcars, input$click, allRows = TRUE)$selected_
  selected(ifelse(clicked, !selected(), selected()))
})
```

If you're plotting with ggplot2, you'll need to set the limits on the colour scale. Otherwise the colour scale will change after the first click because it gains a new level ("FALSE").

```{r}
ui <- fluidPage(
  plotOutput("plot", click = clickOpts("click")),
  tableOutput("data")
)
server <- function(input, output, session) {
  selected <- reactiveVal(rep(TRUE, nrow(mtcars)))
  
  output$plot <- renderPlot({
    mtcars$sel <- selected()
    ggplot(mtcars, aes(wt, mpg)) + 
      geom_point(aes(colour = sel)) +
      scale_colour_discrete(limits = c("TRUE", "FALSE"))
  })
  
  observeEvent(input$click, {
    clicked <- nearPoints(mtcars, input$click, allRows = TRUE)$selected_
    selected(ifelse(clicked, !selected(), selected()))
  })
}
```

Note that in both cases I add the reactive data to the data frame before plotting. I think this is good practice in ggplot2.

### Data flow

1.  Javascript captures mouse event
1.  Shiny send javascript mouse event back to R, invalidating the input.
1.  All downstream reactives are recomputed. `plotOutput()` generates a new
    PNG and sends it to the browser.

For local app, the bottleneck tends to be the time taken to draw the plot. But for hosted apps, you also have to take into account the time needed to transmit the event from the browser to the R, and then the rendered plot back from R to the browser. This means this interactivity is never going to be highly fluid (needs <30 ms response time). 

### Dynamic height and width

One other useful trick. `height` and `width` can be zero-argument functions that return the size in pixels. These are evaluated in a reactive environment so that you can make the size of your plot dynamic.

Here's the basic idea:

```{r}
ui <- fluidPage(
  sliderInput("height", "height", min = 100, max = 500, value = 250),
  sliderInput("width", "width", min = 100, max = 500, value = 250),
  sliderInput("n", "n", min = 10, max = 100, value = 25),
  plotOutput("plot", width = 250, height = 250)
)
server <- function(input, output, session) {
  output$plot <- renderPlot(
    width = function() input$width,
    height = function() input$height,
    {
      plot(rnorm(input$n), rnorm(input$n))
    }
  )
}
```

Note that the plot is re-drawn, but the code is not rerun (i.e. the random values say the same). This is the same behaviour as when you resize a plot with a dynamic height/width (e.g. 100%).

In real cases, you'd use more complicated expressions in the `width` and `height` functions. For example, if you're using a faceted plot in ggplot2, you might use it to increase the size of the plot to keep the individual facet sizes roughly the same (unfortunately there's no easy way to keep them exactly the same because it's currently v. difficult to find out the size of the fixed elements around the borders of the plot.)

## `renderCachedPlot()`

Really useful if a plot is seen by multiple users, 

<https://shiny.rstudio.com/articles/plot-caching.html>

Mostly a matter of changing `renderPlot()` to `renderCachedPlot()`. But you also need to supply a `cacheKeyExpr`. This is some code that returns an object that basically represents the "state" of the plot; whenever that value changes, the plot will be recomputed.

BASIC EXAMPLE

### Cache key

Best to keep it as simple as possible - should be a list of vectors. 

* Input parameters.
* Dataset reactive. Avoid using very large datasets as a cache key - it
  will take some time
* Timestamp. `proc.time()[[3]] %/% 3600`

Also includes id, and plot size. Also implicitly depends on the current user.

### Sizing policy

Normally plots are rendered to the exact size needed. But that's not very useful for caching: if everyone has a slightly different sized plot, then the cache will never be used. `renderCachePlot()` resolves this problem by only caching plots with a fixed sizes and then resizing as needed. 

It uses an exponential rounding strategy determined by the `sizingPolicy` argument. It's unlikely that you'll need to tweak this, but if you do, see the `sizeGrowthRatio()` for details.

Another option worth considering is setting the plots to a fixed size by setting the `width` argument to `plotOutput` to a fixed unit like `"400px"` (the default `height` is already fixed). That will give you the best cache performance because every plot will have the same size.

### Scoping

By default, the plot cache is stored in memory, and shared across all users of the app. If needed, you can override these defaults with:

* `cache = "session"`: the cache lifetime will tied to a single user. It
  will be created when the session starts (i.e. when someone first visits
  you app), and deleted when the session ends. 
  
* `cache = diskCache(...)`: shares across multiple users, multiple     
  processes, and app-restarts. Beware that restarting the app will no 
  longer clear the cache, so if you change the plotting code, you'll also
  need to manually reset the cache by deleting the directory. 

It's also possible to store in a database, or right your own backend. See <https://shiny.rstudio.com/articles/plot-caching.html#can-i-write-my-own-caching-backend> for more details.

The default memory cache size of 10 MB can hold plenty of plots – a cached plot object is typically between 50 and 250 kB. You can override with `shinyOptions(cache = memoryCache(size = 20e6))`.

## Images

If you need to display images that you already have saved on disk. These might come from an external source (maybe a directory of photographs), or perhaps you've generated them with another package (e.g. a 3d scene with`rgl::rgl.snapshot()`).

<https://shiny.rstudio.com/articles/images.html>

Unlike `renderPlot()`, `renderImage()` must return a list:

* `src`: (local) path to the image file. 

* `contentType` entry specifying the MIME type of the image. If one is not 
  provided, will try to detect the type from the file extension. You should
  only need to set this if you're generating images without file extensions,
  which generally you won't do.

* Other components like `width`, `height`

* `class`, `alt` will be added as attributes to the `<img>` tag.

### Static images

Sample app shows directory full of files.

`www/images`: another option is to put images in the `www/` directory. This is somewhat more efficient as it allows the browser to use caching automatically.

`?shiny::addResourcePath`

> Static files under the www/ directory are automatically made available under a request path that begins with /.

Then use `img()` + `renderUI()` to generate img tags.

### Dynamic images

In most cases, you're better off figuring out how to use `renderPlot()` because it takes care of so many details for you, it's easily cached, and the vast majority of R packages that generate images already provided some way to .

* Note the `deleteFile` argument. Set it to `TRUE` if you're rendering the image live, and it should be deleted after its been sent to the user. `outfile <- tempfile(fileext = '.png')`

* You can detect the size of the plot in the browserwith `session$clientData$output_myImage_width` and `session$clientData$output_myImage_height`. (This works for all outputs, but isn't usually needed. ) Also `session$clientData$pixelratio`.





## Exercises

1. Make a plot with click handle that shows all the data returned in the input.

1. Make a plot with click, dblclick, hover, and brush output handlers
   and nicely display the current selection in the sidebar. Plot the 
   plot in the main panel.

1. Compute the limits of the distance scale using the size of the plot.

```{r}
output_size <- function(id) {
  reactive(c(
    session$clientData[[paste0("output_", id, "_width")]],
    session$clientData[[paste0("output_", id, "_height")]]
  ))
}
```
  
