# Reducing duplication with functions {#action-functions}

```{r, include = FALSE}
source("common.R")
```

Sometimes you can extract out duplicated code using functions. For example, if you've copied and pasted some UI code to create variants with different names:

Or you have a self contained set of reactives:

However, a function alone with only take you so far because typically you'll have some connection between the front end and back end, and you need some way to coordinate the two. Shiny uses identifiers so you need some way to share them. This gives rise to Shiny __modules__.

```{r setup}
library(shiny)
```

## Helper functions

If, given specific values, your app requires complex calculation, first start by pulling that calculation out into separate function:

```{r}
server <- function(input, output, session) {
  data <- reactive({
    # complex data calculation involving input$x, input$y, input$z
  })
}
```

```{r}
my_helper <- function(x, y, z) {
  ...
}

server <- function(input, output, session) {
  data <- reactive(my_helper(input$x, input$y, input$z))
}
```

When extracting out such helpers, avoid putting any reactive component inside the function. Instead, pass them in through the arguments.

There are two advantages to using a function:

* It allows you to move it to a separate file

* It makes it clear from the outside exactly what inputs your function
  takes. When looking at a reactive expression or output, there's no way to
  easily tell exactly what values it depends on, except by carefully reading
  the code block. The function definition is a nice signpost that tells you 
  exactly what to inspect.  

A function also _enforces_ this independence --- if you try and refer to an input that you did not pass into the function, you'll get an error. This enforced independence becomes increasingly important as you create bigger and bigger apps because it ensures that pieces of your app are independent and can be analysed in isolation.

As your collection of helper functions grow, you might want to pull them out into their own files. I recommend putting that file in a `R/` directory underneath the app directory. Then load it at the top of your `app.R`:

```{r, eval = FALSE}
library(shiny)

source("R/my-helper-function.R")

server <- function(input, output, session) {
  data <- reactive(my_helper(input$x, input$y, input$z))
}
```

(A future version of shiny will automatically source all files in `R/`, <https://github.com/rstudio/shiny/pull/2547>, so you'll be able to remove the `source()` line.)

## UI functions

You can apply these same ideas to generating your UI. If you have a bunch of controls that you use again and again and again, it's worth doing some up front work to make a function that saves some typing.

This can be useful even if all you're doing is changing three or four default arguments. For example, imagine that you're creating a bunch of sliders that need to each run from 0 to 1, starting at 0.5, with a 0.1 step. You _could_ do a bunch of copy and paste:

```{r}
ui <- fluidRow(
  sliderInput("alpha", "alpha", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("beta", "beta", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("gamma", "gamma", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("delta", "delta", min = 0, max = 1, value = 0.5, step = 0.1)
)
```

But even for this simple case, I think it's worthwhile to pull out the repeated code into a function:

```{r}
sliderInput01 <- function(id, label = id) {
  sliderInput(id, label, min = 0, max = 1, value = 0.5, step = 0.1)
}

ui <- fluidRow(
  sliderInput01("alpha"),
  sliderInput01("beta"),
  sliderInput01("gamma"),
  sliderInput01("delta")
)
```

If you're comfortable with functional programming, you could reduce the code still further as below. htmltools (the package that provides the underlying html code to Shiny) supports tidy dots only in the development version. `fluidRow(!!!list(a, b))` is equivalent to `fluidRow(a, b)`. This technique is sometimes called splatting because you're splatting the elements of a list into the arguments of a function.

```{r}
vars <- c("alpha", "beta", "gamma", "delta")
sliders <- purrr::map(vars, sliderInput01)
ui <- fluidRow(!!!sliders)
```

I'm not going to teach functional programming here, but I will show off some examples. It's a good example of where improving your general R programming skills pays off in your Shiny apps. 

### Reactives

<!-- https://community.rstudio.com/t/r-shiny-apply-custom-function-to-datatable/39790/3 -->

Note that you want to keep as much reactivity inside the server function as possible. So it takes a generic `path` and it returns a data frame, not a reactive.

### Case study

Lets explore this idea with a realistic Shiny app, inspired by a post, <https://community.rstudio.com/t/38506>, on the RStudio community forum. The post contained some code that looks like this:

```{r eval = FALSE}
fluidRow(
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("traffickingType",
      label = "Choose a trafficking type: ",
      choices = sort(unique(ngo$Trafficking.Type)),
      multiple = TRUE
    )
  ),
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("traffickingSubType",
      label = "Choose a trafficking sub type: ",
      choices = sort(unique(ngo$Trafficking.Sub.Type)),
      multiple = TRUE
    )
  ),
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("gender",
      label = "Choose a gender: ",
      choices = sort(unique(ngo$Victim.Gender)),
      multiple = TRUE
    )
  )
)
```

It's a little hard to see what's going on here because repeated code makes the differences harder to see. When looking at this code I see two places where I could extract out a function:

* The call to `box()` repeats `width = 4` and `solidHeader = TRUE`.
  It appears that the intent of this code is making a header, so I'll call the
  function `headerBox`.

* The calls to `selectInput()` repeat `multiple = TRUE` and all use the 
  same strategy for determining the choices: pulling unique values from 
  a data frame column. This function is tied to a specific dataset,
  so I'll call it `ngoSelectInput()`.

That leads me to:

```{r, eval = FALSE}
ngoSelectInput <- function(var, label, multiple = TRUE) {
  choices <- sort(unique(ngo[[var]]))
  label <- paste0("Choose a ", label, ": ")
  selectInput(var, label, choices = choices, multiple = multiple)
}
boxHeader <- function(...) {
  box(width = 4, solidHeader = TRUE, ...)
}

fluidRow(
  boxHeader(ngoSelectInput("Trafficking.Type", "trafficking type")),
  boxHeader(ngoSelectInput("Trafficking.Sub.Type", "trafficking sub type")),
  boxHeader(ngoSelectInput("Victim.Gender", "gender"))
)
```

I made one simplifying assumption that would also require changes on the server side: when filtering based on a variable, the input name should be the same as the variable name. I think this sort of consistency generally makes for code that's easier to read and remember. For example, the names of the new inputs will match up perfectly to the data frame columns if I produce a reactive with only the selected rows:

```{r}
ngo_filtered <- reactive({
  filter(ngo, 
    Trafficking.Type %in% input$Trafficking.Type,
    Trafficking.Sub.Type %in% input$Trafficking.Sub.Type,
    Victim.Gender %in% input$Victim.Gender
  )
})
```

You might consider genearalising to handle multiple datasets:

```{r}
dfSelectInput <- function(df, var, label, multiple = TRUE) {
  choices <- sort(unique(df[[var]]))
  label <- paste0("Choose a ", label, ": ")
  selectInput(var, label, choices = choices, multiple = multiple)
}
```

This would be a good idea if you saw that pattern repeated in multiple places. But you'll probably also need to introduce some additional component for the id. Otherwise `dfSelect(df1, "x")` and `dfSelect(df2, "x")` would generate a control with the same id, which is obviously going to cause problems. This is the problem of namespacing; we want somehow to have a hierarchy in the names. We'll come back to this in modules, as this is one of the big problems that they solve.


If you had a lot more controls, I'd consider using functional programming to generate them. Again, I'll just show an example so if you're already familiar with FP you can see my basic approach. The key idea is to capture all the data you need to generate the columns in a single data frame, which is convenient to create with `tibble::tribble()`. A data frame is useful here because it easily generalises to any number of arguments

```{r}
library(purrr)
vars <- tibble::tribble(
  ~ var,                  ~ label,
  "Trafficking.Type",     "trafficking type",
  "Trafficking.Sub.Type", "trafficking sub type",
  "Victim.Gender",        "gender"
)
```

Then we use `purrr::pmap()` to turn each row in the data frame to a call to `ngoSelectInput()`, use `map()` to wrap each select input into a boxHeader, and then `!!!` to 

```{r, eval = FALSE}
vars %>% 
  pmap(ngoSelectInput) %>% # create one select input for each row
  map(boxHeader) %>%       # wrap each in a boxHeader()
  fluidRow(!!!.)           # collapse into a single fluidRow()
```

If you have really advanced FP skills, you can even generate the call to `dplyr::filter()`:

```{r}
library(rlang)
select <-  map(vars$var, function(v) expr(.data[[!!v]] == input[[!!v]]))
select
```

If you haven't seen `.data` before, it comes from tidy evaluation, the system that allows you to program with tidyverse packages that are designed for interactive exploration (like dplyr). It's not necessary when writing interactive code (and it's not strictly necessary here) but it makes the parallel between the data frame and the inputs more clear. We'll talk more about tidy evaluation in Chapter XXX.

Again we'd use `!!!` to splat the generated expressions into `filter()`:

```{r, eval = FALSE}
filter(ngo, !!!select)
```

Don't worry if this all looks like gibberish: you can just use copy and paste instead. 
