# Functions {#action-functions}

```{r, include = FALSE}
source("common.R")
```

If you have been creating a lot of your app via copy and paste, parts of your app may be very similar to each other. Copy and paste is a great starting technique but if you rely on it too much you end up with two major problems:

* Your app is hard to understand because it's hard to see the important 
  differences when you have a lot of copy and pasted code.
  
* Your app is hard to update because it's easy to forget to update one of the 
  places that you've duplicated code). 
  
A good rule of thumb is that if you have copy and pasted something three times, it's a good time use some other technique to reduce the amount of duplication. In this chapter, we'll show how you can use functions to reuse duplication in code on either the UI or the server. This isn't the only technique you'll need in Shiny, but it's a good place to start. In the next chapter, you'll learn about __modules__, which allow you to reduce duplicated code that requires coordination across the UI and server.

There are two advantages to using a function:

* It allows you to move it to a separate file. 

* It makes it clear from the outside exactly what inputs your function
  takes. 

```{r setup}
library(shiny)
```

## Organisation
  
```{r, eval = FALSE}
library(shiny)

source("R/my-helper-function.R")

server <- function(input, output, session) {
  data <- reactive(my_helper(input$x, input$y, input$z))
}
```

(A future version of shiny will automatically source all files in `R/`, <https://github.com/rstudio/shiny/pull/2547>, so you'll be able to remove the `source()` line.)

As your collection of helper functions grow, you might want to pull them out into their own files. I recommend putting that file in a `R/` directory underneath the app directory. Then load it at the top of your `app.R`:

For more complicated apps, particularly apps that multiple people contribute to, there are substantial advantages to turning your app into a package. In that case, you might want to check out the [golem](https://thinkr-open.github.io/golem/) package and accompanying ["Buidling Big Shiny Apps"](https://thinkr-open.github.io/building-shiny-apps-workflow/) book. I think a particularly important power is that the package infrastructure makes it possible to test your helper functions; and they can be tested independently of reactivity.

## Duplication in UI

You can apply these same ideas to generating your UI. If you have a bunch of controls that you use again and again and again, it's worth doing some up front work to make a function that saves some typing.

### Basic idea

This can be useful even if all you're doing is changing three or four default arguments. For example, imagine that you're creating a bunch of sliders that need to each run from 0 to 1, starting at 0.5, with a 0.1 step. You _could_ do a bunch of copy and paste:

```{r}
ui <- fluidRow(
  sliderInput("alpha", "alpha", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("beta", "beta", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("gamma", "gamma", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("delta", "delta", min = 0, max = 1, value = 0.5, step = 0.1)
)
```

But even for this simple case, I think it's worthwhile to pull out the repeated code into a function:

```{r}
sliderInput01 <- function(id, label = id) {
  sliderInput(id, label, min = 0, max = 1, value = 0.5, step = 0.1)
}

ui <- fluidRow(
  sliderInput01("alpha"),
  sliderInput01("beta"),
  sliderInput01("gamma"),
  sliderInput01("delta")
)
```

If you're comfortable with functional programming, you could reduce the code still further:

```{r}
vars <- c("alpha", "beta", "gamma", "delta")
sliders <- purrr::map(vars, sliderInput01)
ui <- fluidRow(!!!sliders)
```

There are two big ideas here:

* Using a function like `map()` to generate a list containing one slider for 
  each element of `vars`.
  
* Using `!!!` to __splat__ the elements of the list into individual arguments
  to `fluidRow()`. `fluidRow(!!!list(a, b))` is equivalent to `fluidRow(a, b)`.

I'm not going to teach functional programming in this book, but I will show a few aspirational examples. It's a good example of where improving your general R programming skills pays off in your Shiny apps. 

### Case study

Lets explore this idea with a realistic Shiny app, inspired by a post, <https://community.rstudio.com/t/38506>, on the RStudio community forum. The post contained some code that looks like this:

```{r eval = FALSE}
fluidRow(
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("traffickingType",
      label = "Choose a trafficking type: ",
      choices = sort(unique(ngo$Trafficking.Type)),
      multiple = TRUE
    )
  ),
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("traffickingSubType",
      label = "Choose a trafficking sub type: ",
      choices = sort(unique(ngo$Trafficking.Sub.Type)),
      multiple = TRUE
    )
  ),
  box(
    width = 4,
    solidHeader = TRUE,
    selectInput("gender",
      label = "Choose a gender: ",
      choices = sort(unique(ngo$Victim.Gender)),
      multiple = TRUE
    )
  )
)
```

It's a little hard to see what's going on here because repeated code makes the differences harder to see. When looking at this code I see two places where I could extract out a function:

* The call to `box()` repeats `width = 4` and `solidHeader = TRUE`.
  It appears that the intent of this code is making a header, so I'll call the
  function `headerBox`.

* The calls to `selectInput()` repeat `multiple = TRUE` and all use the 
  same strategy for determining the choices: pulling unique values from 
  a data frame column. This function is tied to a specific dataset,
  so I'll call it `ngoSelectInput()`.

That leads me to:

```{r, eval = FALSE}
ngoSelectInput <- function(var, label, multiple = TRUE) {
  choices <- sort(unique(ngo[[var]]))
  label <- paste0("Choose a ", label, ": ")
  selectInput(var, label, choices = choices, multiple = multiple)
}
boxHeader <- function(...) {
  box(width = 4, solidHeader = TRUE, ...)
}

fluidRow(
  boxHeader(ngoSelectInput("Trafficking.Type", "trafficking type")),
  boxHeader(ngoSelectInput("Trafficking.Sub.Type", "trafficking sub type")),
  boxHeader(ngoSelectInput("Victim.Gender", "gender"))
)
```

I made one simplifying assumption that would also require changes on the server side: when filtering based on a variable, the input name should be the same as the variable name. I think this sort of consistency generally makes for code that's easier to read and remember. For example, the names of the new inputs will match up perfectly to the data frame columns if I produce a reactive with only the selected rows:

```{r}
ngo_filtered <- reactive({
  filter(ngo, 
    Trafficking.Type %in% input$Trafficking.Type,
    Trafficking.Sub.Type %in% input$Trafficking.Sub.Type,
    Victim.Gender %in% input$Victim.Gender
  )
})
```

You might consider generalising to handle multiple datasets:

```{r}
dfSelectInput <- function(df, var, label, multiple = TRUE) {
  choices <- sort(unique(df[[var]]))
  label <- paste0("Choose a ", label, ": ")
  selectInput(var, label, choices = choices, multiple = multiple)
}
```

This would be a good idea if you saw that pattern repeated in multiple places. But you'll probably also need to introduce some additional component for the id. Otherwise `dfSelect(df1, "x")` and `dfSelect(df2, "x")` would generate a control with the same id, which is obviously going to cause problems. This is the problem of namespacing; we want somehow to have a hierarchy in the names. We'll come back to this in modules, as this is one of the big problems that they solve.

If you had a lot more controls, I'd consider using functional programming to generate them. Again, I'll just show an example so if you're already familiar with functional programming you can see my basic approach. The key idea is to capture all the data you need to generate the columns in a single data frame, which is convenient to create with `tibble::tribble()`. A data frame is useful here because it easily generalises to any number of arguments

```{r}
library(purrr)
vars <- tibble::tribble(
  ~ var,                  ~ label,
  "Trafficking.Type",     "trafficking type",
  "Trafficking.Sub.Type", "trafficking sub type",
  "Victim.Gender",        "gender"
)
```

Then we use `purrr::pmap()` to turn each row in the data frame to a call to `ngoSelectInput()`, use `map()` to wrap each select input into a boxHeader, and then `!!!` to splat into a row.

```{r, eval = FALSE}
vars %>% 
  pmap(ngoSelectInput) %>% # create one select input for each row
  map(boxHeader) %>%       # wrap each in a boxHeader()
  fluidRow(!!!.)           # collapse into a single fluidRow()
```

## Duplication in server

<!-- https://community.rstudio.com/t/r-shiny-apply-custom-function-to-datatable/39790/3 -->

If, given specific values, your app requires complex calculation, first start by pulling that calculation out into separate function:

```{r}
server <- function(input, output, session) {
  data <- reactive({
    # complex data calculation involving input$x, input$y, input$z
  })
}
```

```{r}
my_helper <- function(x, y, z) {
  ...
}

server <- function(input, output, session) {
  data <- reactive(my_helper(input$x, input$y, input$z))
}
```

When extracting out such helpers, avoid putting any reactive component inside the function. Instead, pass them in through the arguments. 

Reactivity should only live in the server function, or a module server function, which we'll talk about next.

A function also _enforces_ this independence --- if you try and refer to an input that you did not pass into the function, you'll get an error. This enforced independence becomes increasingly important as you create bigger and bigger apps because it ensures that pieces of your app are independent and can be analysed in isolation. 

When looking at a reactive expression or output, there's no way to easily tell exactly what values it depends on, except by carefully reading the code block. The function definition is a nice signpost that tells you  exactly what to inspect.  

### Functional programming

If you're a functional programming aficionado, you could even generate the call to `dplyr::filter()`:

```{r}
library(rlang)
select <-  map(vars$var, function(v) expr(.data[[!!v]] == input[[!!v]]))
select
```

If you haven't seen `.data` before, it comes from tidy evaluation, the system that allows you to program with tidyverse packages that are designed for interactive exploration (like dplyr). It's not necessary when writing interactive code (and it's not strictly necessary here) but it makes the parallel between the data frame and the inputs more clear. We'll talk more about tidy evaluation in Chapter XXX.

Again we'd use `!!!` to splat the generated expressions into `filter()`:

```{r, eval = FALSE}
filter(ngo, !!!select)
```

Don't worry if this all looks like gibberish: you can just use copy and paste instead. 

## Limitation of functions

Functions are great when you have duplicated code in the UI or server that is unconnected to the other side. 

To illustrate why we need modules, and can't just use regular functions, consider the following simple app. It allows the user to input their birthday as a string: this is a little faster than using a `dateInput()` since there's no need to scroll through a calendar. But it means that we need to check that they've entered a correct date and give an informative message if they haven't.

```{r}
library(lubridate)

ui <- fluidPage(
  textInput("date", "When were you born? (yyyy-mm-dd)"),
  textOutput("error"),
  textOutput("age")
)

server <- function(input, output, session) {
  birthday <- reactive({
    req(input$date)
    ymd(input$date, quiet = TRUE)
  })
  
  output$error <- renderText({
    if (is.na(birthday())) {
      "Please enter valid date in yyyy-mm-dd form"
    }
  })
  age <- reactive({
    req(birthday())
    (birthday() %--% today()) %/% years(1)
  })
  output$age <- renderText({
    paste0("You are ", age(), " years old")
  })
}
```

It seems plausible that as your app gets bigger you might want to use this date control in multiple places, so lets have a go at extracting it out into functions. We'll need two functions: one to generate the UI, and one to do the computation on the server side:

```{r}
ymdInputUI <- function(label) {
  label <- paste0(label, " (yyyy-mm-dd)")
  
  fluidRow(
    textInput("date", label),
    textOutput("error")
  )
}

ymdInputServer <- function(input, output, session) {
  date <- reactive({
    req(input$date)
    ymd(input$date, quiet = TRUE)
  })
  
  output$error <- renderText({
    if (is.na(date())) {
      "Please enter valid date in yyyy-mm-dd form"
    }
  })

  date
}
```

That leads to the following app:

```{r}
ui <- fluidPage(
  ymdInputUI("When were you born?"),
  textOutput("age")
)

server <- function(input, output, session) {
  birthday <- ymdInputSever(input, output, session)
  
  age <- reactive({
    req(birthday())
    (birthday() %--% today()) %/% years(1)
  })
  
  output$age <- renderText({
    paste0("You are ", age(), " years old")
  })
}
```

There are two problems with this approach:

*   It always assumes that the control is called `date`. This means that 
    we can't have two controls in the same app.
  
    ```{r}
    ui <- fluidPage(
      ymdInputUI("When was your mother born?"),
      ymdInputUI("When was your father born?")
    )
    ```
  
*   The UI has a output with id `error` that you can't see when reading 
    the UI code. This makes it very easy to accidentally break the app by
    introducing two controls with the same id:
  
    ```{r}
    ui <- fluidPage(
      ymdInputUI("When were you born?"),
      textOutput("error")
    )
    ```

    Debugging the problem that this creates will be painful because it will 
    reveal itself through failure of reactivity -- the output won't update as you
    expect, or you'll get weird errors because two controls are fighting for 
    the same input value. 

These problems arise because we've used functions to isolate local variables; the code is simpler to understand because any variables created inside of `ymdInputUI()` and `ymdInputServer()` can't be accessed outside. But there's another important way that Shiny code can interface: through the names of input and output controls.

This is the problem that modules are designed to solve: creating inputs and reactives that are completely isolated from the rest of your app. Learning how to use modules will take a little time, but it will pay off by giving you the ability to write components that are guaranteed to be isolated from everything else in your app.
