# Dynamic UI {#action-dynamic}

```{r, include = FALSE}
source("common.R")
```

As well as updating the content of the outputs, sometimes you will actually want to add more, or modify the inputs, or generally change the apperance of the UI with code. These techniques should be used sparingly, because they begin to step outside of reactivity of they add substantially more complexity to your app, but when you need them, you need them.

In Chapter XYZ, we'll come back to the techniques (including `conditionalPanel()`, `insertUI()`, and `removeUI()`) that require you know a little more HTML/CSS/JS.

```{r setup}
library(shiny)
```

## Modify existing inputs

Every input control, e.g. `textInput()`, is paired with a update function, e.g. `updateTextInput()`, that allows you to modify the control after it has been created. Take this very simple example. The app has three numeric inputs, where the first two control the range (the `min` and `max`) of the last.

```{r}
ui <- fluidPage(
  numericInput("min", "min", 0),
  numericInput("max", "max", 10),
  numericInput("n", "n", 5)
)
server <- function(input, output, session) {
  observeEvent(input$min, {
    updateNumericInput(session, "n", min = input$min)
  })  
  observeEvent(input$max, {
    updateNumericInput(session, "n", max = input$max)
  })
}
```

Note that I've used `observeEvent()` here, although `observe()` would also have worked and would yield shorter code. I generally prefer `observeEvent()` because it cleanly separates what you are listening for, and what action you want to take in response.

You might notice that the call to `updateNumericInput()` looks a little different to other Shiny functions. It uses `session`, which is the third argument passed into `server`.  So if you're using an older `server` function template, you might not have `session` in the arguments, so make sure you add it if its missing.

The primary use case of the update functions is to make the user interface more user friendly. Next, you'll see a couple of simple examples of how you might use them, then we'll dive into a more complicated case study of hierarchical select boxes. We'll finish off by discussing circular references, a major challenge with update functions, and the related problem of multiple sources of truth.

### Simple uses

The simplest uses of the update functions are to provide small conveniences for the user. For example, maybe you want to make it easy to reset some parameters back to their starting place. The following snippet shows how you might combine an `ationButton()`, `observeEvent()` and `updateNumericInput()`:

```{r}
ui <- fluidPage(
  sliderInput("x1", "x1", 0, min = -10, max = 10),
  sliderInput("x2", "x2", 0, min = -10, max = 10),
  sliderInput("x3", "x3", 0, min = -10, max = 10),
  actionButton("reset", "Reset")
)

server <- function(input, output, session) {
  observeEvent(input$reset, {
    updateNumericInput(session, "x1", value = 0)
    updateNumericInput(session, "x2", value = 0)
    updateNumericInput(session, "x3", value = 0)
  })
}
```

A similar application is to tweak the text of an action button so you know exactly what it's going to do:

```{r}
ui <- fluidPage(
  numericInput("n", "Simulations", 10),
  actionButton("simulate", "Simulate")
)

server <- function(input, output, session) {
  observeEvent(input$n, {
    label <- paste0("Simulate ", input$n, " times")
    updateActionButton(session, "simulate", label = label)
  })
}
```

There are many ways to use update functions in this way; be on the look out for ways to give more information to the user when you are working on sophisticated applications. A particularly important application is making it easier to select from a long list of possible options, through step-by-step filtering. That's often a problem for "hierarchical select boxes".

### Hierarchical select boxes

A more complicated, but particularly useful, application of the update functions is to allow interactive drill down across multiple categories. I'm going to start some imaginary data for a sales dashboard, coming from <https://www.kaggle.com/kyanyoga/sample-sales-data>.

For our purposes, I'm going to focus on a natural hierarchy in the data:

* Each territory contains customers
* Each customer has multiple orders
* Each order contains rows

I want to create a user interface where you can:

* Select a territory to see all customers.
* Select customer to see all orders.
* Select order to see data.

The basic idea is to start with UI that contains three select boxes, and one output table. The choices for the `customername` and `ordernumber` will be dynamically supplied so, I explicitly set `choices = NULL`.

```{r}
sales <- vroom::vroom("sales-dashboard/sales_data_sample.csv")

ui <- fluidPage(
  selectInput("territory", "Territory", choices = unique(sales$TERRITORY)),
  selectInput("customername", "Customer", choices = NULL),
  selectInput("ordernumber", "Order number", choices = NULL),
  tableOutput("data")
)
```

Then in the server function, I work top-down, first creating a filtered subset that only contains the selected territory and using that to update `input$customername`, then creating a subset contains the given customer and using that to update `input$ordernumber`.

```{r}
server <- function(input, output, session) {
  territory <- reactive({
    filter(sales, TERRITORY == input$territory)
  })
  observeEvent(territory, {
    choices <- unique(territory()$CUSTOMERNAME)
    updateSelectInput(session, "customername", choices = choices) 
  })
  
  customer <- reactive({
    filter(territory(), CUSTOMERNAME == input$customername)
  })
  observeEvent(input$customername, {
    choices <- unique(customer()$ORDERNUMBER)
    updateSelectInput(session, "ordernumber", choices = choices)
  })
  
  output$data <- renderTable({
    filter(customer(), ORDERNUMBER == input$ordernumber)
  })
}
```

You can see a more fleshed out application of this principle in <https://github.com/hadley/mastering-shiny/tree/master/sales-dashboard>.

Consider using this, or similar techniques, whenever the user might have to select from a large number of choices. You're free to use any other inputs for filtering: if you (e.g.) wanted to make it easier to find orders based on the total value, you might want to use `sliderInput()`s instead.

### Circular references

If you want to use update functions to change the `value`[^other-params] of inputs, there's an important issue we need to discuss. From Shiny's perspectve, using an update function to modify `value` is no different to the user modifying the value. That means an update function can trigger reactive updates in exactly the same way that a human can. You are now stepping outside of the bounds of pure reactive programming, so you need to start worrying about circular references and creating infinite loops. 

[^other-params]: This is generally only a concern when you are changing the `value`, but be aware that changing some other parameters can occassionally change the value, as when changing set of `choices` for `inputSelect()`.

For example, take the following simple app. It contains a single input control and a observer that increments its value by one. Every time `updateNumericInput()` runs, it invalidates `input$n`, causing `updateNumericInput()` to run again, so the app is stuck in an infinite loop constantly increasing the value of `input$n`.

```{r}
ui <- fluidPage(
  numericInput("n", "n", 0)
)
server <- function(input, output, session) {
  observeEvent(input$n,
    updateNumericInput(session, "n", value = input$n + 1)
  )
}
```

You're unlikely to create such an obvious problem in your own app, but beware this potential problem if you are updating multiple controls that depend on one another. You will need to carefully reason through the updates to ensure that you're not creating an infinite loop of updates. This is a very good reason that you should only update functions for the most important cases.

### Inter-related inputs

One place where it's easy to end up with circular references if you're attempting to have multiple "sources of truth" in an app. For example, imagine you want to create a temperatue conversion app where you can either enter the temperature in Celsius or in Fahrenheit:

```{r}
ui <- fluidPage(
  numericInput("temp_c", "Celsius", NA),
  numericInput("temp_f", "Fahrenheit", NA)
)

server <- function(input, output, session) {
  observeEvent(input$temp_f, {
    c <- round((input$temp_f - 32) * 5 / 9)
    updateNumericInput(session, "temp_c", value = c)
  })
  
  observeEvent(input$temp_c, {
    f <- round((input$temp_c * 9 / 5) + 32)
    updateNumericInput(session, "temp_f", value = f)
  })
}
```

If you play around with this app, you'll notice that it _mostly_ works, but you might notice that it'll sometimes trigger multiple changes. For example:

* Set 120 F
* Click down. F changes to 119, and C is updated to 48.
* 48 C is converted to 118 F, so F changes again to 118.
* Fortunately 118 F is still 48 C, so the updates stop there.

There's no way around this problem because you have one idea (the temperature) with two expressions in the app (Celsius and Fahrenheit). Here we are lucky that cycle quickly converges to a value that satisfies both constraints.

<!-- Similar problem: https://community.rstudio.com/t/mutually-dependent-numericinput-in-shiny/29307 -->

## Dynamic tabs

```{r}
ui <- fluidPage(
  tabsetPanel(
    type = "pills",
    tabPanel("Plot", plotOutput("plot")),
    tabPanel("Summary", verbatimTextOutput("summary")),
    tabPanel("Table", tableOutput("table"))
  )
)
```

## Dynamic UI

Sometimes modifying the values of an existing control is not enough, and you need control the number of type of controls. There's a special pair of output and render functions that allow you to do UI on the server: `uiOutput()` and `renderUI()`

Here's a simple example that dynamically creates a different type of input control depending on an input:

```{r}
ui <- fluidPage(
  textInput("label", "label"),
  selectInput("type", "type", c("slider", "numeric")),
  uiOutput("numeric")
)
server <- function(input, output, session) {
  output$numeric <- renderUI({
    if (input$type == "slider") {
      sliderInput("dynamic", input$label, isolate(input$dynamic), min = 0, max = 100)
    } else {
      numericInput("dynamic", input$label, isolate(input$dynamic), min = 0, max = 100)  
    }
  })
}
```

(We haven't talked about `isolate()` yet; allows you to get a reactive value _without_ creating a reactive dependency.)

Note the way I set the default value to `input$dynamic`; that ensures the selected value is not lost when the control type changes. If you're generating a lot of dynamic UI, it's good practice to think about the user experience and make sure that existing data isn't lost.

Also note that it takes a fraction of a second to appear after the app loads - that's because it has to be rendered by the server function. Relying too much on `renderUI()` can create a laggy UI.

Note that you are now creating IDs in two places so that when you add to `ui`, you have to be careful not to call the control `dynamic`. Later, in Chapter \@ref(action-modules) we'll see how modules can help avoid this problem.

### Multiple controls

`tagList()` if you need mutliple controls. Want to keep as much as of the fixed structure in UI as possible. Better peformance. Simpler reasoning.

Using functional programming to generate multiple controls. And equivalent code on the reactive side.

### Dynamic filtering

Select dataset, then generate.


