# User feedback {#action-feedback}

```{r, include = FALSE}
source("common.R")
```

Important to respond to the user as they interact with your app. In this chapter we'll show how you can dynamically push feedback out of R and into the browser.  

Many of the techniques here are best described as side-effects so that they can appear only at the end of the reactive graph.

```{r setup}
library(shiny)
```

## Validation

### Ignore missing inputs

Most frequent offender is `textInput()`, because it starts off blank. Can also use with `inputSelect()` with a choice of `""`. Another important use case is with `fileInput()`, see Section \@ref(#upload) for more details.

The easiest way to fix this problem is with `req()`: it checks that the input has been set before proceeding. `req()` sends a special signal to Shiny telling it that the reactive isn't ready yet[^technically]

[^technically]: Technically, it leaves any downstream reactive consumers in an invalidated state. We'll come back to this terminology in Chapter \@ref{reactive-components}.



You can use `req()` inside of an output, but it's good practice to use inside a `reactive()` because it isolates computation in one place.

Truthi-ness. `req()` is designed so that `req(input$x)` will do what you expect regardless of what type of input `x` is. You can see the details in `?isTruthy`



### Validate input

You can extend the same technique to provide other feedback.

`validate()`

With `tryCatch()`

## Notifications

### Progress

* Can divide the task into units that take roughly the same amount of time.
* Can divide the task into discrete steps

Progress bar. Simplest way is to use `withProgress()`

```{r}
ui <- fluidPage(
  numericInput("steps", "How many steps?", 5),
  actionButton("go", "go")
)
server <- function(input, output, session) {
  observeEvent(input$go, {
    withProgress(message = "Hi!", {
      for (i in seq_len(input$steps)) {
        Sys.sleep(0.1)
        setProgress(i / input$steps, message = NULL)
      }
    })
  })
}
```

If you need more control, see the details in `?Progress` and <https://shiny.rstudio.com/articles/progress.html>.

With progress package - i.e. forwarding the conditions it generates to `withProgress()`. Should be able to make simple wrapper.

```{r, echo = FALSE}
shinyProgress <- function(expr) {
  progress <- NULL
  on.exit(if (!is.null(progress)) progress$close())
  
  tryCatch(expr,
    progress_message = function(cnd) {
      switch(cnd$msgtype,
        add_job = {
          progress <<- Progress$new()
        },
        complete_job = {
          progress$close()
          progress <<- NULL
        },
        set_job_status = progress$set(message = cnd$status),
        set_job_output = progress$set(details = cnd$output),
        set_job_progress = progress$set(value = cnd$progress),
        add_job_progress = progress$inc(amount = cnd$increment)
      )
    }
  )
}
```

<!-- Indeterminate progress: https://github.com/rstudio/shiny/issues/609 -->

### Message

If you don't know how long some code will take, a better approach is to use notifications.

```{r}
server <- function(input, output, session) {
  data <- reactive({
    id <- showNotification("Reading data...", duration = NULL, closeButton = FALSE)
    on.exit(removeNotification(id), add = TRUE)
    
    read.csv(path())
  })
}
```

Two important ideas here:

* We captured the notification id created by `showNotification()`.

* We use `on.exit()` to automatically cancel the notification when the 
  reactive complete, regardless of whether it returns a value or throws
  an error.
  
Also note the use of `duration = NULL` and `closeButton = FALSE` that makes the notification most suitable for this task, ensuring that is stays visibile until the data loading is done.
  
(For this specific case you should also look at `data.table::fread()` and `vroom::vroom()` to read in the file; they can be orders of magnitude faster than `read.csv()`. And pointer to chapter about performance/promises/future)

You can slightly extend this approach to send multiple notifications if there are discrete steps (where you also don't know how long they'll take)

```{r}
server <- function(input, output, session) {
  data <- reactive({
    id <- showNotification("Reading data...", duration = NULL, closeButton = FALSE)
    on.exit(removeNotification(id), add = TRUE)
    Sys.sleep(0.5)
      
    showNotification("Reticulating spines...", id = id, duration = NULL, closeButton = FALSE)
    Sys.sleep(0.5)
  })
}
```

### Modal 

Sometimes you need to give more information to the user than you can fit in a simple notification; or the information is so important that you want to ensure that the 

I think it's a good idea to put the modal ui 

```{r}
ui <- fluidPage(
  actionButton("go", "Go")
)
modal_ui <- modalDialog(
  "This is really important!",
  "Are you sure you want to continue?",
  footer = list(
    actionButton("yes", "yes"),
    actionButton("no", "no")
  ),
  easyClose = FALSE
)

server <- function(input, output, session) {
  observeEvent(input$go, {
    showModal(modal_ui)
  })

  observeEvent(input$yes, 
    showModal(modalDialog("DELETING ALL FILES", size = "l", footer = NULL))
  )
  observeEvent(input$no, 
    removeModal()
  )
}
```

You'll see another technique when we come back to modules.
