# Dynamic UI {#action-dynamic}

```{r, include = FALSE}
source("common.R")
source("demo.R")
```

So far, we've seen a clean separation between the user interface and the server function. The user interface is defined statically, when the app is launched. That means it can't respond to anything that happens in the app. In this chapter, you'll learn how to create __dynamic__ user interfaces, using code run in the server function. 
There are three key techniques for creating dynamic user interfaces:

* Using the `update` functions to modify the parameters of input controls.

* Using `tabsetPanel()` to conditionally show and hide parts of the user 
  interface.

* Using `uiOutput()` and `renderUI()` to generate selected parts of the user 
  interface with code. 

These three tools give you considerable power to respond to the user by modifying inputs and outputs. I'll demonstrate some of the more useful ways in which you can apply them, but ultimately, you are only constrained by your creativity. That said, these tools can make your app substantially more difficult to reason about, so deploy them sparingly, and always strive for the simplest technique that solves your problem.

```{r setup}
library(shiny)
```

## Updating inputs

We'll begin with a simple technique that allows you to modify an input after it has been created. Every input control, e.g. `textInput()`, is paired with an __update function__, e.g. `updateTextInput()`, that allows you to modify the control after it has been created. The update funtions look a little different to other Shiny functions: they all takes the current `session` as the first argument, and the name of the input (as a string) as the second. The remaining arguments match the arguments to the input constructor.

Take the example in the code below, with the results shown in Figure \@ref(fig:update-basics). The app has two inputs that control the range (the `min` and `max`) of another input, a slider. The key idea is to use `observeEvent()`[^observe] to trigger `updateSliderInput()` whenever the `min` or `max` inputs change.

```{r}
ui <- fluidPage(
  numericInput("min", "Minimum", 0),
  numericInput("max", "Maximum", 3),
  sliderInput("n", "n", min = 0, max = 3, value = 1)
)
server <- function(input, output, session) {
  observeEvent(input$min, {
    updateSliderInput(session, "n", min = input$min)
  })  
  observeEvent(input$max, {
    updateSliderInput(session, "n", max = input$max)
  })
}
```

```{r update-basics, fig.cap = demo$caption("The app on load (left), after increasing max (middle), and then decreasing min (right)."), out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("update-basics", ui, server)
demo$resize(350)
s1 <- demo$screenshot("onload")
demo$set_values(max = 4)
s2 <- demo$screenshot("max-increase")
demo$set_values(min = -1)
s3 <- demo$screenshot("min-decrease")
knitr::include_graphics(c(s1, s2, s3))
demo$deploy()
```

[^observe]: Note that I've used `observeEvent()` here, although `observe()` would also work and would be more concise. I generally prefer `observeEvent()` because its arguments cleanly separate the event you're listening for from the action you want to take in response.

To help you get the hang of the update functions, I'll show a couple more simple examples. Then we'll dive into a more complicated case study using hierarchical select boxes, and I'll finish off by discussing circular references and the related problem of multiple sources of truth.

### Simple uses

The simplest uses of the update functions are to provide small conveniences for the user. For example, maybe you want to make it easy to reset parameters back to their initial value. The following snippet shows how you might combine an `actionButton()`, `observeEvent()` and `updateSliderInput()`, with the output shown in Figure \@ref(fig:update-reset).

```{r}
ui <- fluidPage(
  sliderInput("x1", "x1", 0, min = -10, max = 10),
  sliderInput("x2", "x2", 0, min = -10, max = 10),
  sliderInput("x3", "x3", 0, min = -10, max = 10),
  actionButton("reset", "Reset")
)

server <- function(input, output, session) {
  observeEvent(input$reset, {
    updateSliderInput(session, "x1", value = 0)
    updateSliderInput(session, "x2", value = 0)
    updateSliderInput(session, "x3", value = 0)
  })
}
```

```{r update-reset, fig.cap = demo$caption("The app on load (left), after dragging some sliders (middle), then clicking reset (right)."), out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("update-reset", ui, server)
demo$resize(350)
s1 <- demo$screenshot("onload")
demo$set_values(x1 = 5, x2 = -5)
s2 <- demo$screenshot("set")
demo$click("reset")
s3 <- demo$screenshot("reset")
knitr::include_graphics(c(s1, s2, s3))
demo$deploy()
```

A similar application is to tweak the text of an action button so you know exactly what it's going to do. Figure \@ref(fig:update-button) shows the results of the code below.

```{r}
ui <- fluidPage(
  numericInput("n", "Simulations", 10),
  actionButton("simulate", "Simulate")
)

server <- function(input, output, session) {
  observeEvent(input$n, {
    label <- paste0("Simulate ", input$n, " times")
    updateActionButton(session, "simulate", label = label)
  })
}
```

```{r update-button, fig.cap = demo$caption("The app on load (left), after setting simulations to 1 (middle), then settting simulations to 10 (right)."), out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("update-button", ui, server)
demo$resize(350)
s1 <- demo$screenshot("onload")
demo$set_values(n = 1)
s2 <- demo$screenshot("set1")
demo$set_values(n = 100)
s3 <- demo$screenshot("set100")
demo$deploy()
knitr::include_graphics(c(s1, s2, s3))
```

There are many ways to use update functions in this way; be on the look out for ways to give more information to the user when you are working on sophisticated applications. A particularly important application is making it easier to select from a long list of possible options, through step-by-step filtering. That's often a problem for "hierarchical select boxes".

### Hierarchical select boxes {#hierarchical-select}

A more complicated, but particularly useful, application of the update functions is to allow interactive drill down across multiple categories. I'll illustrate their usage with some imaginary data for a sales dashboard that comes from <https://www.kaggle.com/kyanyoga/sample-sales-data>.

```{r}
sales <- vroom::vroom("sales-dashboard/sales_data_sample.csv", col_types = list())
sales
```

For our purposes, I'm going to focus on a natural hierarchy in the data:

* Each territory contains customers.
* Each customer has multiple orders.
* Each order contains rows.

I want to create a user interface where you can:

* Select a territory to see all customers.
* Select a customer to see all orders.
* Select an order to see the underlying rows.

The essence of the UI is simple: I'll create three select boxes and one output table. The choices for the `customername` and `ordernumber` select boxes will be dynamically generated, so I set `choices = NULL`.

```{r}
ui <- fluidPage(
  selectInput("territory", "Territory", choices = unique(sales$TERRITORY)),
  selectInput("customername", "Customer", choices = NULL),
  selectInput("ordernumber", "Order number", choices = NULL),
  tableOutput("data")
)
```

In the server function, I work top-down:

1.  I create a reactive, `territory()`, that contains the rows from `sales` 
    that match the selected territory.

1.  Whenever `territory()` changes, I update the list of `choices` in 
    the `input$customername` select box.
    
1.  I create another reactive, `customer()`, that contains the rows from 
    `territory()` that match the selected customer.
    
1.  Whenever `customer()` changes, I update the list of `choices` in the
    the `input$ordernumber` select box.

1.  I display the selected orders in `output$data`.

```{r}
server <- function(input, output, session) {
  territory <- reactive({
    filter(sales, TERRITORY == input$territory)
  })
  observeEvent(territory(), {
    choices <- unique(territory()$CUSTOMERNAME)
    updateSelectInput(session, "customername", choices = choices) 
  })
  
  customer <- reactive({
    req(input$customername)
    filter(territory(), CUSTOMERNAME == input$customername)
  })
  observeEvent(customer(), {
    choices <- unique(customer()$ORDERNUMBER)
    updateSelectInput(session, "ordernumber", choices = choices)
  })
  
  output$data <- renderTable({
    req(input$ordernumber)
    customer() %>% 
      filter(ORDERNUMBER == input$ordernumber) %>% 
      select(QUANTITYORDERED, PRICEEACH, PRODUCTCODE)
  })
}
```

```{r update-nested, fig.cap = demo$caption('I select "EMEA" (left), then "Lyon Souveniers" (middle), then (right) look at the orders.'), out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("update-nested", ui, server, packages = "dplyr", data = list(sales = sales))
demo$resize(400, 500)
demo$set_values(territory = "NA")
demo$drop_down("territory", 2)
s1 <- demo$screenshot("territory")
demo$set_values(territory = "EMEA")
demo$drop_down("customername", 2)
s2 <- demo$screenshot("customername")
demo$set_values(customername = "Lyon Souveniers")
s3 <- demo$screenshot("orders")
knitr::include_graphics(c(s1, s2, s3))
demo$deploy()
```

Try out this simple example at `r demo$link()`, or see a more fully fleshed out application at <https://github.com/hadley/mastering-shiny/tree/master/sales-dashboard>.

### Circular references

There's an important issue we need to discuss if you want to use update functions to change the `value`[^other-params] of inputs. From Shiny's perspectve, using an update function to modify `value` is no different to the user modifying the value by clicking or typing. That means an update function can trigger reactive updates in exactly the same way that a human can. This means that you are now stepping outside of the bounds of pure reactive programming, and you need to start worrying about circular references and infinite loops. 

[^other-params]: This is generally only a concern when you are changing the `value`, but be some other parameters can change the value indirectly. For example, if you modify the `choices` for `selectInput()`, or `min` and `max` for `sliderInput()`, the current `value` will be modified if it's no longer in the allowed set of values.

For example, take the following simple app. It contains a single input control and a observer that increments its value by one. Every time `updateNumericInput()` runs, it changes `input$n`, causing `updateNumericInput()` to run again, so the app gets stuck in an infinite loop constantly increasing the value of `input$n`.

```{r}
ui <- fluidPage(
  numericInput("n", "n", 0)
)
server <- function(input, output, session) {
  observeEvent(input$n,
    updateNumericInput(session, "n", value = input$n + 1)
  )
}
```

You're unlikely to create such an obvious problem in your own app, but it can crop up if you updating multiple controls that depend on one another, as in the next example.

### Inter-related inputs

One place where it's easy to end up with circular references is when you have multiple "sources of truth" in an app. For example, imagine that you want to create a temperatue conversion app where you can either enter the temperature in Celsius or in Fahrenheit:

```{r}
ui <- fluidPage(
  numericInput("temp_c", "Celsius", NA),
  numericInput("temp_f", "Fahrenheit", NA)
)

server <- function(input, output, session) {
  observeEvent(input$temp_f, {
    c <- round((input$temp_f - 32) * 5 / 9)
    updateNumericInput(session, "temp_c", value = c)
  })
  
  observeEvent(input$temp_c, {
    f <- round((input$temp_c * 9 / 5) + 32)
    updateNumericInput(session, "temp_f", value = f)
  })
}
```

```{r update-temperature, include = FALSE}
# Screenshots don't work because the updates happen too quickly to show 
# the intermediate state
demo <- demo_inline("update-temperature", ui, server)
demo$resize(300)
demo$set_values(temp_f = 120)
demo$screenshot("f-120")
demo$set_values(temp_f = 119)
demo$screenshot("f-119")
demo$screenshot("f-118")
demo$deploy()
```

If you play around with this app, `r demo$link()`, you'll notice that it _mostly_ works, but you might notice that it'll sometimes trigger multiple changes. For example:

* Set 120 F, then click the down arrow.
* F changes to 119, and C is updated to 48.
* 48 C converts to 118 F, so F changes again to 118.
* Fortunately 118 F is still 48 C, so the updates stop there.

There's no way around this problem because you have one idea (the temperature) with two expressions in the app (Celsius and Fahrenheit). Here we are lucky that cycle quickly converges to a value that satisfies both constraints. In general, you are better off avoiding these situations, unless you are willing to very carefully analysis the convergence properties of the underlying dynamic system that you have created.

### Exercises

1.  Complete the user interface below with a server function that updates
    `input$date` so that you can only select dates in `input$year`.
    
    ```{r}
    ui <- fluidPage(
      numericInput("year", "year", value = 2020),
      dateInput("date", "date")
    )
    ```
    
1.  Complete the user interface below with a server function that updates
    `input$county` choices based on `input$state`. For an added challenge,
    also change the label from "County" to "Parrish" for Louisana and 
    "Borrough" for "Alaska".

    ```{r, messages = FALSE}
    library(openintro)
    states <- unique(county$state)
    
    ui <- fluidPage(
      selectInput("state", "State", choices = states),
      selectInput("county", "County", choices = NULL)
    )
    ```

1.  Complete the user interface below with a server function that updates
    `input$country` choices based on the `input$continent`. Use 
    `output$data` to display all matching rows.

    ```{r}
    library(gapminder)
    continents <- unique(gapminder$continent)
    
    ui <- fluidPage(
      selectInput("continent", "Continent", choices = continents), 
      selectInput("country", "Country", choices = NULL),
      tableOutput("data")
    )
    ```

1.  Extend the previous app so that you can also choose to select no continent,
    and hence see all countries. You'll need to add `""` to the list of
    choices, and then handle that specially when filtering.

1.  What is at the heart of the problem described at 
    <https://community.rstudio.com/t/29307>?

## Dynamic visibility

The next step up in complexity is to selectively show and hide parts of the UI. You'll learn a number of sophisticated approaches later, once you've learned a little JS and CSS, but there's a useful technique that you can use now: concealing optional UI in a tabset. This is a clever hack that allows you to show and hide UI as needed, without having to re-generate it from scratch (as you'll learn in the next section).

Currently[^tab-future] you need a smidgen of CSS to make this technique work: `tags$style("#switcher { display:none; }")`. If you adapt this for your own code make sure to repace `#switcher` with `#` followed the `id` of your the `tabsetPanel()`.

[^tab-future]: Hopefully, it will be built into `tabsetPanel()` in the future; follow <https://github.com/rstudio/shiny/issues/2680> for details.

```{r}
ui <- fluidPage(
  tags$style("#switcher { display:none; }"),
  sidebarLayout(
    sidebarPanel(
      selectInput("controller", "Show", choices = paste0("panel", 1:3))
    ),
    mainPanel(
      tabsetPanel(
        id = "switcher",
        tabPanel("panel1", "Panel 1 content"),
        tabPanel("panel2", "Panel 2 content"),
        tabPanel("panel3", "Panel 3 content")
      )
    )
  )
)

server <- function(input, output, session) {
  observeEvent(input$controller, {
    updateTabsetPanel(session, "switcher", selected = input$controller)
  })
}
```

```{r dynamic-panels, fig.cap = demo$caption("Selecting panel1 (left), then panel2 (middle), then panel3 (right)."), out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("dynamic-panels", ui, server)
demo$resize(300, 220)
s1 <- demo$screenshot()
demo$set_values(controller = "panel2")
s2 <- demo$screenshot("panel2")
demo$set_values(controller = "panel3")
s3 <- demo$screenshot("panel3")
knitr::include_graphics(c(s1, s2, s3))
# demo$deploy()
```

There are two main ideas here:

* Use tabset panel with hidden tabs.
* Use `updateTabsetPanel()` to switch tabs from the server.

This is a simple idea, but when combined with a little creativity, it gives you a considerable amount of power. The following two sections illustrate give a couple of small examples of how you might use it in practice.

### Conditional UI

Imagine that you want an app that allows the user to simulate from the normal, uniform, and exponential distributions. Each distribution has different parameters, so we'll need some way to show different controls for different distributions. Here, I'll put the unique user interface for each distribution in its on `tabPanel()`, and then arrange the three tabs into a `tabsetPanel()`.

```{r}
parameter_tabs <- tagList(
  tags$style("#params { display:none; }"),
  tabsetPanel(id = "params",
    tabPanel("normal",
      numericInput("mean", "mean", value = 1),
      numericInput("sd", "standard deviation", min = 0, value = 1)
    ),
    tabPanel("uniform", 
      numericInput("min", "min", value = 0),
      numericInput("max", "max", value = 1)
    ),
    tabPanel("exponential",
      numericInput("rate", "rate", value = 1, min = 0),
    )
  )
)
```

I'll then embed that inside a fuller UI which allows the user to pick the number of samples and shows a histogram of the results:

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("dist", "Distribution", 
        choices = c("normal", "uniform", "exponential")
      ),
      numericInput("n", "Number of samples", value = 100),
      parameter_tabs,
    ),
    mainPanel(
      plotOutput("hist")
    )
  )
)
```

Note that I've carefully matched the `choices` in `input$dist` to the names of the tab panels. That makes it easy to write the `observeEvent()` code below that automatically switches controls when the distribution changes. The rest of the app uses techniques that you're already familiar with. See the final result in Figure \@ref(fig:dynamic-conditional).

```{r}
server <- function(input, output, session) {
  observeEvent(input$dist, {
    updateTabsetPanel(session, "params", selected = input$dist)
  }) 
  
  sample <- reactive({
    switch(input$dist,
      normal = rnorm(input$n, input$mean, input$sd),
      uniform = runif(input$n, input$min, input$max),
      exponential = rexp(input$n, input$rate)
    )
  })
  output$hist <- renderPlot(hist(sample()))
}
```

```{r dynamic-conditional, fig.cap = demo$caption("Results for normal (left), uniform (middle), and exponential (right) distributions."), out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("dynamic-conditional", ui, server)
demo$resize(400)
demo$sleep(1)
s1 <- demo$screenshot("normal")
demo$set_values(dist = "uniform")
Sys.sleep(0.1)
s2 <- demo$screenshot("uniform")
demo$set_values(dist = "exponential")
Sys.sleep(0.1)
s3 <- demo$screenshot("exponential")
knitr::include_graphics(c(s1, s2, s3))
# demo$deploy()
```

### Wizard interface

You can also use this idea to a "wizard", a type of interface that makes it easier to collect a bunch of information by spreading it across multiple pages. Here we embed action buttons within each "page", making it easy to go forward and back.

```{r}
ui <- fluidPage(
  tags$style("#wizard { display:none; }"),
  tabsetPanel(id = "wizard",
    tabPanel("page1", 
      "Welcome!",
      actionButton("page12", "next")
    ),
    tabPanel("page2", 
      "Only one page to go",
      actionButton("page21", "prev"),
      actionButton("page23", "next")
    ),
    tabPanel("page3", 
      "You're done!",
      actionButton("page32", "prev")
    )
  )
)

server <- function(input, output, session) {
  switch_tab <- function(page) {
    updateTabsetPanel(session, "wizard", selected = page)
  }
  
  observeEvent(input$page12, switch_tab("page2"))
  observeEvent(input$page21, switch_tab("page1"))
  observeEvent(input$page23, switch_tab("page3"))
  observeEvent(input$page32, switch_tab("page2"))
}
```

### Exercises


## Dialog boxes {#dynamic-dialog}

The simplest way to create UI with code is to add a dialog box. A dialog box is "modal" because it creates a new "mode" of interaction; you can't interact with the main application until you have dealt with the dialog.

`modalDialog()` works much like `fluidPage()`: 

## Creating UI with code {#programming-ui}

Sometimes none of the techniques described above gives you the level of dynamism that you need. There's one last technique that gives you the ability to create any controls (both inputs and outputs) with code in the server function. You've always created your UI with code; this technique gives you the ability to re-generate while the app is running, not just define it when the app starts.

This technique has two components:

* You use `uiOutput()` to insert a placeholder in your user interface.
  This code is run when your app launches and it leaves a "hole" that your 
  server code can later fill in.
  
* You use `renderUI()` to fill in the placeholder with UI generated in the 
  server function.

Here's a simple example to illustrate the basic idea. It dynamically creates a different type of input control depending on an input. The resulting app is shown in Figure \@ref(fig:render-simple).

```{r}
ui <- fluidPage(
  textInput("label", "label"),
  selectInput("type", "type", c("slider", "numeric")),
  uiOutput("numeric")
)
server <- function(input, output, session) {
  output$numeric <- renderUI({
    if (input$type == "slider") {
      sliderInput("dynamic", input$label, value = 0, min = 0, max = 10)
    } else {
      numericInput("dynamic", input$label, value = 0, min = 0, max = 10)  
    }
  })
}
```

```{r render-simple, fig.cap = demo$caption("App after setting value to 5 (left), then changing type to numeric (middle), then label to 'my label'."), out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("render-simple", ui, server)
demo$resize(400)
s1 <- demo$screenshot("onload")
demo$set_values(type = "numeric")
s2 <- demo$screenshot("numeric")
demo$set_values(label = "My label")
s3 <- demo$screenshot("label")
knitr::include_graphics(c(s1, s2, s3))
demo$deploy()
```

If you run this code yourself, you'll notice that it takes a fraction of a second to appear after the app loads. That's because it's reactive: the app must load, trigger a reactive event, which calls the server function, yielding HTML to insert into the page. This is one of the downsides of `renderUI()`; relying on it too much can create a laggy UI. For good performance, strive to keep fixed as much of the user interface as possible, using the techniques described earlier in the chapter.

There's one other problem with this approach: when you change controls, you lose the currently selected value. Maintaining existing state is one of the big challenges of creating UI with code. This is one reason that selectively showing and hiding UI (as above) is a better approach if works for you - because you're not destroying and recreating the controls, you don't need to do anything to presrve the values. In many cases, we can fix the problem by setting the `value` of the new input to the current value of the existing control:

```{r}
server <- function(input, output, session) {
  output$numeric <- renderUI({
    if (input$type == "slider") {
      sliderInput("dynamic", input$label, value = isolate(input$dynamic))
    } else {
      numericInput("dynamic", input$label, value = isolate(input$dynamic))  
    }
  })
}
```

(I've removed the `min` and `max` arguments make it easier to fit everything on one line; you'd still need them in real code.)

The use of `isolate()` is important. We'll come back to exactly why it's needed in Chapter XYZ, but it ensures that we don't create a reactive dependency that would mean this code is re-run every time `input$dynamic` changes. We only want it to change when `input$type` or `inpu$label` changes.

Dynamic UI is most useful when you are generating an arbitrary number or type of controls. That means that you'll be generating UI with code, and I recommend using functional programming for this sort of task. You're certainly free to use base `lapply()` and `Reduce()` functions for this pupose, but I'm going to use `purrr::map()` and `purrr::reduce()`, because I find them a little easier to work with:

```{r}
library(purrr)
```

If you're not familiar with the basic `map()` and `reduce()` of functional programming, you might want to take a brief detour to read <https://adv-r.hadley.nz/functionals.html> before continuing. 

### Multiple controls

An important use of `renderUI()` is when you need to create an arbitrary number of controls. For example, imagine you'd like the user to be able to supply their own palette of colours. They'll first specify how many colours they want, and then supply a value for each colour. The `ui` is pretty simple: we have a `numericInput()` that controls the number of inputs, a `uiOutput()` where the generated text boxes will go, and a `textOutput()` that demonstrates that we've plumbed everything together correctly.

```{r}
ui <- fluidPage(
  numericInput("n", "Number of colours", value = 5, min = 1),
  uiOutput("col"),
  textOutput("palette")
)
```

There are three key ideas in the server function:

* I create a reactive, `col_names()`, that generates a character vector 
  giving the name of each of the colour inputs I'm about to generate. 

* I create the text boxes by using `map()` to create one `textInput()` for 
  input in `col_names()`. `output$col <- renderUI()` inserts these textboxes
  in the UI placeholder that I created earlier.
  
* To generate the output, I need to use use a new trick. So far we've always
  accessed the components of `input` with `$`, e.g. `input$col1`. But here we
  have the input names in a character vector, like `var <- "col1"`. 
  `$` no longer works in this scenario, so we need to swich to `[[`, i.e.
  `input[[var]]`. I use `map_chr()` to collect all values into a character
  vector, and display that in `output$pallete`.

```{r}
server <- function(input, output, session) {
  col_names <- reactive(paste0("col", seq_len(input$n)))
  
  output$col <- renderUI({
    map(col_names(), ~ textInput(.x, NULL))
  })
  
  output$palette <- renderText({
    map_chr(col_names(), ~ input[[.x]])
  })
}
```

You can see the results in Figure \@ref(fig:render-palette).

```{r render-palette, fig.cap = demo$caption("App on load (left), after setting n to 3 (middle), then entering some colours (right)."), out.width = "33%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("render-palette", ui, server, packages = "purrr")
demo$resize(400)
s1 <- demo$screenshot("onload")
demo$set_values(n = 3)
s2 <- demo$screenshot("change-n")
demo$set_values(col1 = "red", col2 = "yellow", col3 = "orange")
s3 <- demo$screenshot("set-cols")
knitr::include_graphics(c(s1, s2, s3))
demo$deploy()
```

If you run this app, you'll discover a really annoying behaviour: whenever you change the number of colours, all the data you've entered disappears. We can fix this problem by using the same technique as before: setting `value` to the (isolated) current value. I'll also tweak the appearance to look a little nicer, including displaying the selected colours in a plot. Sample screenshots are shown in Figure \@ref(fig:render-palette-full).

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      numericInput("n", "Number of colours", value = 5, min = 1),
      uiOutput("col"),
    ),
    mainPanel(
      plotOutput("plot")  
    )
  )
)
server <- function(input, output, session) {
  col_names <- reactive(paste0("col", seq_len(input$n)))
  
  output$col <- renderUI({
    map(col_names(), ~ textInput(.x, NULL, value = isolate(input[[.x]])) %||% "")
  })
  
  output$plot <- renderPlot({
    cols <- map_chr(col_names(), ~ input[[.x]])
    cols[cols == ""] <- NA
    
    barplot(
      rep(1, length(cols)), 
      col = cols,
      space = 0, 
      axes = FALSE
    )
  })
}
```

```{r render-palette-full, fig.cap = demo$caption("Filling out the colours of the rainbow (left), then reducing the number of colours to 3 (right); note that the existing colours are preserved."), out.width = "50%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("render-palette-full", ui, server, packages = "purrr")
demo$resize(800)
demo$set_values(
  col1 = "red", 
  col2 = "yellow", 
  col3 = "orange", 
  col4 = "green", 
  col5 = "blue"
)
s1 <- demo$screenshot("rainbow")
demo$set_values(n = 3)
s2 <- demo$screenshot("change-n")
knitr::include_graphics(c(s1, s2))
demo$deploy()
```

### Dynamic filtering

To finish off the chapter, I'm going to create an app that lets you dynamically filter any data frame. Each numeric input will get a range slider and each factor input will get a multi-select. i.e. if a data frame has 3 continuous variables and 2 factors, I'll generate an app with 3 sliders and 2 select boxes.

I'll start with a function that creates the UI for a single variable. It'll return a range slider for numeric inputs, a multi-select for factor inputs, and `NULL` (nothing) for all other types. 

```{r}
make_ui <- function(x, var) {
  if (is.numeric(x)) {
    rng <- range(x, na.rm = TRUE)
    sliderInput(var, var, min = rng[1], max = rng[2], value = rng)
  } else if (is.factor(x)) {
    levs <- levels(x)
    selectInput(var, var, choices = levs, selected = levs, multiple = TRUE)
  } else {
    # Not supported
    NULL
  }
}
```

And then I'll write the server side equivalent of this function: it takes the variable and value of the input control, and returns a logical vector saying whether or not to include each observation. I return a logical vector here because it'll make it easy to combine the results from multiple columns.

```{r}
filter_var <- function(x, val) {
  if (is.numeric(x)) {
    !is.na(x) & x >= val[1] & x <= val[2]
  } else if (is.factor(x)) {
    x %in% val
  } else {
    # No control, so don't filter
    TRUE
  }
}
```

I can then use these functions "by hand" to generate a simple filtering UI for the `iris` dataset:

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      make_ui(iris$Sepal.Length, "Sepal.Length"),
      make_ui(iris$Sepal.Width, "Sepal.Width"),
      make_ui(iris$Species, "Species")
    ),
    mainPanel(
      tableOutput("data")
    )
  )
)
server <- function(input, output, session) {
  selected <- reactive({
    filter_var(iris$Sepal.Length, input$Sepal.Length) &
      filter_var(iris$Sepal.Width, input$Sepal.Width) &
      filter_var(iris$Species, input$Species)
  })
  
  output$data <- renderTable(head(iris[selected(), ], 12))
}
```

```{r render-filter-1, fig.cap = "Simple filter interface for the iris dataset", echo = FALSE, out.width = "75%"}
demo <- demo_inline("render-filter-1", ui, server, data = list(make_ui = make_ui, filter_var = filter_var))
demo$resize(800)
demo$screenshot()
```

You might notice that I got sick of copying and pasting so that app only works with three columns. I can make it work with all the columns by using a little functional programming:

* In `ui` use `map()` to generate one control for each variable. 

* In `server()`, I use `map()` to generate the selection vector for each 
  variable. Then I use `reduce()` to take the logical vector for each variable
  and combine into a single logical vector by `&`-ing each vector together.

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      map(names(iris), ~ make_ui(iris[[.x]], .x))
    ),
    mainPanel(
      tableOutput("data")
    )
  )
)
server <- function(input, output, session) {
  selected <- reactive({
    each_var <- map(names(iris), ~ filter_var(iris[[.x]], input[[.x]]))
    reduce(each_var, `&`)
  })
  
  output$data <- renderTable(head(iris[selected(), ], 12))
}
```

```{r render-filter-2, fig.cap = "Using functional programming to build filtering app for the `iris` dataset", echo = FALSE, out.width = "75%"}
demo <- demo_inline("render-filter-2", ui, server, packages = "purrr", data = list(make_ui = make_ui, filter_var = filter_var))
demo$resize(800)
demo$screenshot()
```

From there, it's a simple generalisation to work with any data frame. Here I'll illustrate it using the data frames in the datasets package, but you can easily imagine how you might extend this to user uploaded data. See the result in Figure \@ref(fig:filtering-final).

```{r}
dfs <- keep(ls("package:datasets"), ~ is.data.frame(get(.x, "package:datasets")))

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("dataset", label = "Dataset", choices = dfs),
      uiOutput("filter")
    ),
    mainPanel(
      tableOutput("data")
    )
  )
)
server <- function(input, output, session) {
  data <- reactive({
    get(input$dataset, "package:datasets")
  })
  vars <- reactive(names(data()))
  
  output$filter <- renderUI(
    map(vars(), ~ make_ui(data()[[.x]], .x))
  )
  
  selected <- reactive({
    each_var <- map(vars(), ~ filter_var(data()[[.x]], input[[.x]]))
    reduce(each_var, `&`)
  })
  
  output$data <- renderTable(head(data()[selected(), ], 12))
}
```

```{r filtering-final, fig.cap = demo$caption("A dynamic user interface automatically generated from the fields of the selected dataset."), out.width = "100%", fig.show = "hold", fig.align = "default", echo = FALSE, message = FALSE, cache = FALSE}
demo <- demo_inline("filtering-final", ui, server, 
  packages = "purrr", 
  data = list(dfs = dfs, make_ui = make_ui, filter_var = filter_var)
)
demo$set_values(dataset = "Formaldehyde")
demo$resize(800)
demo$screenshot()
demo$deploy()
```

### Exercises

1.  Take this very simple app based on the initial example in the chapter:

    ```{r}
    ui <- fluidPage(
      selectInput("type", "type", c("slider", "numeric")),
      uiOutput("numeric")
    )
    server <- function(input, output, session) {
      output$numeric <- renderUI({
        if (input$type == "slider") {
          sliderInput("n", "n", value = 0, min = 0, max = 100)
        } else {
          numericInput("n", "n", value = 0, min = 0, max = 100)  
        }
      })
    }
    ```
    
    How could you instead implement it using dynamic visibility? If you
    implement dynamic visiblity, how could you keep the values in sync when
    you change the controls?

1.  Add support for date and date-time columns `make_ui()` and `filter_var()`.
    
1.  (Advanced) If you know the [S3 OOP](http://adv-r.hadley.nz/S3.html) system,
    consider how you could replace the `if` blocks in `make_ui()` and
    `filter_var()` with generic functions.

1.  (Hard) Make a wizard that allows the user to upload their own dataset.  
    The first page should handle the upload. The second should handle reading
    it, providing one drop down for each variable that lets the user select
    the column type. The third page should provide some way to get a summary
    of the dataset.
