# Reducing duplication with modules {#action-modules}

```{r, include = FALSE}
source("common.R")
```

Advantages of functions and modules:

*   Clearly isolated behaviour through specified inputs and outputs means it
    is easier to understand how parts of your app fit together, and you don't
    have to worry about spooky action at a distance where changing one part
    of your app changes the way an apparently unrelated part works.
   
*   Reducing duplication makes it easier to respond to changing needs because
    instead of having to track down and change every place you duplicated code,
    you can just change it in one place.
  
*   You can spread your app across multiple files, so that it can be more 
    easily digested in chunks. Because you're using functions and modules
    you can read the files independently. You don't have to load all the 
    pieces into your head to understand how the whole thing hangs together.

Often the hardest part is decomposing your big problem into smaller independent pieces. I include some case studies here to help you get a sense of how this feels, but ultimately it's a skill that can only be learned with practice. Try and set aside some time each week where you're not improving the behaviour or appearance of your app, but simply making it easier to understand. This will make your app easier to change in the future, and as you practice these skills your first attempt will become higher quality.

```{r setup}
library(shiny)

# We're experimenting with a new style of module construction that
# will appear in the next version of Shiny.
moduleFunction <- function(id, module) {
  callModule(module, id)
}
```

## Motivation

To illustrate why we need modules, and can't just use regular functions, consider the following simple app. It allows the user to input their birthday as a string: this is a little faster than using a `dateInput()` since there's no need to scroll through a calendar. But it means that we need to check that they've entered a correct date and give an informative message if they haven't.

```{r}
library(lubridate)

ui <- fluidPage(
  textInput("date", "When were you born? (yyyy-mm-dd)"),
  textOutput("error"),
  textOutput("age")
)

server <- function(input, output, session) {
  birthday <- reactive({
    req(input$date)
    ymd(input$date, quiet = TRUE)
  })
  
  output$error <- renderText({
    if (is.na(birthday())) {
      "Please enter valid date in yyyy-mm-dd form"
    }
  })
  age <- reactive({
    req(birthday())
    (birthday() %--% today()) %/% years(1)
  })
  output$age <- renderText({
    paste0("You are ", age(), " years old")
  })
}
```

It seems plausible that as your app gets bigger you might want to use this date control in multiple places, so lets have a go at extracting it out into functions. We'll need two functions: one to generate the UI, and one to do the computation on the server side:

```{r}
ymdInputUI <- function(label) {
  label <- paste0(label, " (yyyy-mm-dd)")
  
  fluidRow(
    textInput("date", label),
    textOutput("error")
  )
}

ymdInputServer <- function(input, output, session) {
  date <- reactive({
    req(input$date)
    ymd(input$date, quiet = TRUE)
  })
  
  output$error <- renderText({
    if (is.na(date())) {
      "Please enter valid date in yyyy-mm-dd form"
    }
  })

  date
}
```

That leads to the following app:

```{r}
ui <- fluidPage(
  ymdInputUI("When were you born?"),
  textOutput("age")
)

server <- function(input, output, session) {
  birthday <- ymdInputSever(input, output, session)
  
  age <- reactive({
    req(birthday())
    (birthday() %--% today()) %/% years(1)
  })
  
  output$age <- renderText({
    paste0("You are ", age(), " years old")
  })
}
```

There are two problems with this approach:

*   It always assumes that the control is called `date`. This means that 
    we can't have two controls in the same app.
  
    ```{r}
    ui <- fluidPage(
      ymdInputUI("When was your mother born?"),
      ymdInputUI("When was your father born?")
    )
    ```
  
*   The UI has a output with id `error` that you can't see from reading just
    the UI code. This makes it very easy to accidentally break the app.
  
    ```{r}
    ui <- fluidPage(
      ymdInputUI("When were you born?"),
      textOutput("error")
    )
    ```

    Debugging the problem that this creates will be painful because it will 
    reveal itself through failure of reactivity -- the output won't update as you
    expect, or you'll get weird errors because two controls are fighting for 
    the same input value. 

These problems arise because we've used functions to isolate local variables; the code is simpler to understand because any variables created inside of `ymdInputUI()` and `ymdInputServer()` can't be accessed outside. But there's another important way that Shiny code can interface: through the names of input and output controls.

This is the problem that modules are designed to solve: creating inputs and reactives that are completely isolated from the rest of your app. Learning how to use modules will take a little time, but it will pay off by giving you the ability to write components that are guaranteed to be isolated from everything else in your app.

## Using modules

A Shiny module is a pair of functions, corresponding to the front end UI and the backend server function. 

Modules are way to create an app within an app. They force isolation of behaviour so that one module can't affect another, and code outside of a module can only affect the inside in a way that the module explicitly allows.

### Making a module

To convert the code above into a module, we need to make two changes. First we need to add an `id` argument to our UI component, and use it with special `NS()` function. `NS` is short for namespace: it creates a "space" of "names" that is unique to the module. 

```{r}
ymdInput <- function(id, label) {
  label <- paste0(label, " (yyyy-mm-dd)")
  
  fluidRow(
    textInput(NS(id, "date"), label),
    textOutput(NS(id, "error"))
  )
}
```

(Notice I've called this `ymdInput()` to match Shiny's existing style.)

The key idea is that the first argument to `NS()` is supplied by the person using the component, and the arguments to the function it produces is supplied by the person who wrote the component. This two-part creation ensures that the final name combines properties needed by both the app author and the module author. This is a bit confusing at first, because you're likely to be both the app and module author.

We now need to specify an `id` when creating the UI. This is important because it puts this id in the same place as all the others, so it's easy to spot if you've used the same input id in multiple places. 

```{r}
ui <- fluidPage(
  ymdInput("birthday", "When were you born?"),
  textOutput("age")
)
```

We need to make a similar change to the server side of the module. Here instead of `NS()` we use `moduleFunction()`. `moduleFunction()` automatically tweaks the `input` and `output` so it looks for `date` inside the `id` namespace. You can think of it as doing `input[[id]]$birthday` (but it's actually `input[[paste(id, "-", birthday)]]`).

```{r}
ymdDate <- function(id) {
  moduleFunction(id, function(input, output, session) {
    date <- reactive({
      req(input$date)
      ymd(input$date, quiet = TRUE)
    })
    
    output$error <- renderText({
      if (is.na(date())) {
        "Please enter valid date in yyyy-mm-dd form"
      }
    })
  
    date
  })
}
```

(One difference with a regular server function is that this function returns a value)

Now the arguments to `ymdInput()` have changed: we pass in the `id`, and Shiny takes care of automatically plumbing up the input, output, and session in the appropriate namespaced way. That why I've removed the `Server` from the name - since the details are hidden from the interface.

```{r}
server <- function(input, output, session) {
  birthday <- ymdDate("birthday")
  
  age <- reactive({
    req(birthday())
    (birthday() %--% today()) %/% years(1)
  })
  
  output$age <- renderText({
    paste0("You are ", age(), " years old")
  })
}
```

To help cement the ideas of modules in your head, the following case studies use module to develop a few simple reusable components.

### Limited selection + other

Consider the following app, which provides a way to select gender that is sensitive to the many possible ways that people can express their gender.[^gender]

[^gender]: For a deeper dive on this issue, and a discussion of why many commonly used way of asking about gender can be hurtful to some people, I recommend reading "Designing forms for gender diversity and inclusion" by Sabrina Fonseca: <https://uxdesign.cc/d8194cf1f51>.

```{r}
ui <- fluidPage(
  radioButtons("gender", "Gender:",
    choiceValues = list("male", "female", "self-described", "na"),
    choiceNames = list(
      "Male",
      "Female",
      textInput("gender_self", NULL, placeholder = "Self-described"),
      "Prefer not to say"
    ),
    selected = "na",
  ),
  textOutput("txt")
)

server <- function(input, output, session) {
  observeEvent(input$gender_self, {
    req(input$gender_self)
    updateRadioButtons(session, "gender", selected = "self-described")
  })
  
  gender <- reactive({
    if (input$gender == "self-described") {
      input$gender_self
    } else {
      input$gender
    }
  })
  
  output$txt <- renderText({
    paste("You chose", gender())
  })
}
```

Convert to a module and generalise a little.

```{r}
radioButtonsExtraInput <- function(id, label, choices, selected = NULL, placeholder = NULL) {
  ns <- NS(id)
  
  radioButtons(ns("primary"), "Gender:",
    choiceValues = c(names(choices), "other"),
    choiceNames = c(
      unname(choices),
      list(textInput(ns("other"), NULL, placeholder = NULL))
    ),
    selected = selected
  )
}

radioButtonsExtra <- function(id) {
  moduleFunction(id, function(input, output, server) {
    observeEvent(input$primary, {
      req(input$other)
      updateRadioButtons(session, "primary", selected = "other")
    })
    
    reactive({
      if (input$primary == "other") {
        input$other
      } else {
        input$primary
      }
    })
  })
}

ui <- fluidPage(
  radioButtonsWithOtherUi("gender", 
    label = "Gender", 
    choices = list(
      male = "Male",
      female = "Female",
      na = "Prefer not to say"
    ), 
    placeholder = "Self-described", 
    selected = "na"
  ),
  textOutput("txt")
)

server <- function(input, output, session) {
  gender <- radioButtonsWithOther("gender")

  output$txt <- renderText({
    paste("You chose", gender())
  })
}
```

### Hierarchical select boxes

```{r, eval = FALSE}
library(tidyverse)
country_df <- countrycode::codelist %>% 
  as_tibble() %>% 
  select(iso3c, continent, country = cow.name) %>% 
  filter(!is.na(continent), !is.na(country))

continents <- sort(unique(country_df$continent))

ui <- fluidPage(
  selectInput("continent", "Continent", choices = continents),
  selectInput("country", "Country", choices = NULL)
)

server <- function(input, output, session) {
  countries <- reactive({
    country_df[country_df$continent == input$continent, , drop = FALSE]
  })
  
  observeEvent(input$continent, {
    updateSelectInput(session, "country", choice = countries()$country)
  })
}

shinyApp(ui, server)
```

### Modal

<https://gist.github.com/hadley/8d9ee5ea7991b0e5c400320abb9468de>

### Returning multiple reactives

* Leave in a list. 
* Use zeallot

### Exercises

1. The following app plots user selected variables from the `msleep` dataset 
   for three different types of mammals (carnivores, omnivores, and herbivores), 
   with one tab for each type of mammal. Remove the redundancy in the 
   `selectInput()` definitions with the use of functions.

```{r, eval = FALSE}
library(tidyverse)

ui <- fluidPage(
  selectInput(inputId = "x",
              label = "X-axis:",
              choices = c("sleep_total", "sleep_rem", "sleep_cycle", 
                          "awake", "brainwt", "bodywt"),
              selected = "sleep_rem"),
  selectInput(inputId = "y",
              label = "Y-axis:",
              choices = c("sleep_total", "sleep_rem", "sleep_cycle", 
                          "awake", "brainwt", "bodywt"),
              selected = "sleep_total"),
  tabsetPanel(id = "vore",
              tabPanel("Carnivore",
                       plotOutput("plot_carni")),
              tabPanel("Omnivore",
                       plotOutput("plot_omni")),
              tabPanel("Herbivore",
                       plotOutput("plot_herbi")))
)

server <- function(input, output, session) {

  # make subsets
  carni <- reactive( filter(msleep, vore == "carni") )
  omni  <- reactive( filter(msleep, vore == "omni")  )
  herbi <- reactive( filter(msleep, vore == "herbi") )

  # make plots
  output$plot_carni <- renderPlot({
    ggplot(data = carni(), aes_string(x = input$x, y = input$y)) +
      geom_point()
  })
  output$plot_omni <- renderPlot({
    ggplot(data = omni(), aes_string(x = input$x, y = input$y)) +
      geom_point()
  })
  output$plot_herbi <- renderPlot({
    ggplot(data = herbi(), aes_string(x = input$x, y = input$y)) +
      geom_point()
  })

}

shinyApp(ui = ui, server = server)
```

2. Continue working with the same app from the previous exercise, and further 
   remove redundancy in the code by modularizing how subsets and plots are 
   created.
   
3. Suppose you have an app that is slow to launch when a user visits it. Can  
   modularizing your app code help solve this problem? Explain your reasoning.
