# Reducing duplication with modules {#action-modules}

```{r, include = FALSE}
source("common.R")
```

Before reading chapter make sure that you're already with the previous chapter on using functions to reduce duplication in a Shiny app. Modules provide very similar benefits to functions, but have an explicit connection between ui and server components through **namespacing**.

There are two main reasons to use modules:

*   To isolate subcomponents of your app through specified inputs and outputs.
    This makes it easier to understand how your app works because you don't
    have to load all the pieces into your head to understand how the whole 
    thing hangs together. It also allows you to spread a large app across 
    multiple files.
    
*   To create reusable user interface components that encompass both 
    client-side HTML and server-side behaviour. 

These are not discrete cases, but points of a spectrum:

* No inputs or outputs
* Inputs or outputs
* Inputs and outputs

Often the hardest part is decomposing your big problem into smaller independent pieces. I include some case studies here to help you get a sense of how this feels, but ultimately it's a skill that can only be learned with practice. Try and set aside some time each week where you're not improving the behaviour or appearance of your app, but simply making it easier to understand. This will make your app easier to change in the future, and as you practice these skills your first attempt will become higher quality.

```{r setup}
library(shiny)

# This is a new style of module construction that will appear in shiny 1.5.0.
# For now, we're adding a simple shim that you can use in your own, and delete
# when shiny 1.5.0 is out.
moduleFunction <- function(id, module) {
  callModule(module, id)
}
```

## What is a module?

A module is just a pair of functions. One function is responsible for generating the UI components; one function is responsible for generating the server side. Both functions take "id" as the first argument, and use it in a standard way. This is the key difference with modules, and is what guarantees their isolation and allows you to use them multiple times in the same app.

Let's take a very simple app and divide it up into modules. This app is so simple that there's no real benefit to modularising it, but it will serve to illustrate the main differences with modules before we dive into more realistic use cases. This app draws a plot of a histogram for user selected variable and number of bins.

```{r}
ui <- fluidPage(
  selectInput("var", "Variable", names(mtcars)),
  numericInput("bins", "bins", 10, min = 1),
  plotOutput("hist")
)
server <- function(input, output, session) {
  data <- reactive(mtcars[[input$var]])
  output$hist <- renderPlot({
    hist(data(), breaks = input$bins, main = input$var)
  })
}
```

### Module UI

To turn this into a module we need to wrap both UI and server components in specially constructed functions. We'll start with the UI. There are two steps:

* We wrap it inside a function. The function must take an `id` argument.
  It can also take other arguments; we'll come back to that later
  
* Each existing ID needs to get "namespaced". Practically this means you
  transform (e.g.) `"var"` to `NS(id, "var")`. The key idea is that the first 
  argument to `NS()` is supplied by the user of the module, and the second 
  argument is supplied by the author of the module. This two-part creation 
  means that you as a module author don't need to worry about clashing with
  other UI components created by the user. You have your own "space" of names
  that you control. This is the key idea of namespaces.

This yields the following function:

```{r}
histogramUi <- function(id) {
  list(
    selectInput(NS(id, "var"), "Variable", names(mtcars)),
    numericInput(NS(id, "bins"), "bins", 10, min = 1),
    plotOutput(NS(id, "hist"))
  )
}
```

Note that here we wrap the UI components in a list; that means they will get included in the parent container. Returning a list is most flexible because it allows the caller of the module to put it in whatever container they want. But if in your app, you always place the module in the same container you might want to include that in the UI function and instead return a `column()` or a `fluidRow()`.

### Module server

Next we tackle the server function. This gets wrapped inside a another function which must have an `id` argument (again, it can have other arguments and we'll come back to those shortly). The module server calls `moduleFunction()` with the id, and a function that looks like a regular server function. This always has arguments `input`, `output`, and `session`.

```{r}
histogramServer <- function(id) {
  moduleFunction(id, function(input, output, server) {
    data <- reactive(mtcars[[input$var]])
    output$hist <- renderPlot({
      hist(data(), breaks = input$bins, main = input$var)
    })
  })
}
```

The code inside the module server is isolated - it can only read from inputs and write to output defined by the UI function with the matching ID. If you want to take input from reactives elsewhere in the app, you'll need to pass them to the module function explicitly; we'll come back to that later.

Note that the namespacing in the UI is explicit; you have to manually instrument each `id` argument with `NS()`. The namespace in the server is **implicit**; Shiny automatically modifies the `input` and `output` arguments so work it works automatically without your intervention.

### Updated app

Now that we have these pieces we can rewrite our app to use them:

```{r}
ui <- fluidPage(
  histogramUi("hist1")
)
server <- function(input, output, session) {
  histogramServer("hist1")
}
```

Note that, just like regular inputs/output and reactives the UI and server are linked by a shared ID. Otherwise they're black boxes and you can't see inside them. Shortly we'll see how you can add inputs and outputs to your reactives so that they can interact with other parts of your app.

## Dynamic UI

One extra little trick - you can get the current namespace through `session`.

### Advantages

There are two big advantages to using a module: isolation and reusability.

Firstly all the code inside the histogram is isolated. The UI below would still works because the histogram module is isolated from the new bins input. You can imagine how helpful this would be when working with large apps!

```{r}
ui <- fluidPage(
  histogramUi("hist1"),
  sliderInput("bins", "bins", 5, min = 1, max = 10)
)
```

The other advantage is that you can easily reuse in multiple places in the app.

```{r}
ui <- fluidPage(
  histogramUi("hist1"),
  histogramUi("hist2")
)
server <- function(input, output, session) {
  histogramServer("hist1")
  histogramServer("hist2")
}
```

## Isolated components

Common design for large apps is to arrange independent components with a tabset. This is useful design pattern for the user because it keeps related inputs close to their outputs, reducing the cognitive burden of using the app. Shiny modules allow you to also get the benefits on the backend: you can convert each tab to a separate pair of isolated UI and server functions that live in their own file. This makes your code easier to understand because its now split up across multiple files; each file can be analysed independently, because the outside connections are clearly demarcated.

```{r}
dists <- c(
  "Normal" = "norm",
  "Uniform" = "unif",
  "Log-normal" = "lnorm",
  "Exponential" = "exp"
)

ui <- fluidPage(
  titlePanel("Tabsets"),
  sidebarLayout(
    sidebarPanel(
      radioButtons("dist", "Distribution type:", dists),
      sliderInput("n", "Number of observations:", value = 500, min = 1, max = 1000)
    ),

    mainPanel(
      tabsetPanel(
        tabPanel("Plot", plotOutput("plot")),
        tabPanel("Summary", verbatimTextOutput("summary")),
        tabPanel("Table", tableOutput("table"))
      )
    )
  )
)
```

## Inputs and outputs

Should you pass a reactive or the current value of a reactive? Always reactive! Only take a constant if it's something that is used only during the setup of your module code.

How to return multiple values? Return a list of reactives.

You'll see quite a bit of advice online to use `session$userData`. I think this is generally bad advice because it takes an end run around namespacing and re-introduces many of the problems that you're trying to avoid by using modules. 

## Reusable components


## Using modules

### Making a module

To convert the code above into a module, we need to make two changes. First we need to add an `id` argument to our UI component, and use it with special `NS()` function. `NS` is short for namespace: it creates a "space" of "names" that is unique to the module. 

```{r}
ymdInput <- function(id, label) {
  label <- paste0(label, " (yyyy-mm-dd)")
  
  fluidRow(
    textInput(NS(id, "date"), label),
    textOutput(NS(id, "error"))
  )
}
```

(Notice I've called this `ymdInput()` to match Shiny's existing style.)


We now need to specify an `id` when creating the UI. This is important because it puts this id in the same place as all the others, so it's easy to spot if you've used the same input id in multiple places. 

```{r}
ui <- fluidPage(
  ymdInput("birthday", "When were you born?"),
  textOutput("age")
)
```

We need to make a similar change to the server side of the module. Here instead of `NS()` we use `moduleFunction()`. `moduleFunction()` automatically tweaks the `input` and `output` so it looks for `date` inside the `id` namespace. You can think of it as doing `input[[id]]$birthday` (but it's actually `input[[paste(id, "-", birthday)]]`).

```{r}
ymdDate <- function(id) {
  moduleFunction(id, function(input, output, session) {
    date <- reactive({
      req(input$date)
      ymd(input$date, quiet = TRUE)
    })
    
    output$error <- renderText({
      if (is.na(date())) {
        "Please enter valid date in yyyy-mm-dd form"
      }
    })
  
    date
  })
}
```

(One difference with a regular server function is that this function returns a value)

Now the arguments to `ymdInput()` have changed: we pass in the `id`, and Shiny takes care of automatically plumbing up the input, output, and session in the appropriate namespaced way. That why I've removed the `Server` from the name - since the details are hidden from the interface.

```{r}
server <- function(input, output, session) {
  birthday <- ymdDate("birthday")
  
  age <- reactive({
    req(birthday())
    (birthday() %--% today()) %/% years(1)
  })
  
  output$age <- renderText({
    paste0("You are ", age(), " years old")
  })
}
```

To help cement the ideas of modules in your head, the following case studies use module to develop a few simple reusable components.

### Limited selection + other

Consider the following app, which provides a way to select gender that is sensitive to the many possible ways that people can express their gender.[^gender]

[^gender]: For a deeper dive on this issue, and a discussion of why many commonly used way of asking about gender can be hurtful to some people, I recommend reading "Designing forms for gender diversity and inclusion" by Sabrina Fonseca: <https://uxdesign.cc/d8194cf1f51>.

```{r}
ui <- fluidPage(
  radioButtons("gender", "Gender:",
    choiceValues = list("male", "female", "self-described", "na"),
    choiceNames = list(
      "Male",
      "Female",
      textInput("gender_self", NULL, placeholder = "Self-described"),
      "Prefer not to say"
    ),
    selected = "na",
  ),
  textOutput("txt")
)

server <- function(input, output, session) {
  observeEvent(input$gender_self, {
    req(input$gender_self)
    updateRadioButtons(session, "gender", selected = "self-described")
  })
  
  gender <- reactive({
    if (input$gender == "self-described") {
      input$gender_self
    } else {
      input$gender
    }
  })
  
  output$txt <- renderText({
    paste("You chose", gender())
  })
}
```

Convert to a module and generalise a little.

```{r}
radioButtonsExtraUI <- function(id, label, choices, selected = NULL, placeholder = NULL) {
  ns <- NS(id)
  
  radioButtons(ns("primary"), "Gender:",
    choiceValues = c(names(choices), "other"),
    choiceNames = c(
      unname(choices),
      list(textInput(ns("other"), NULL, placeholder = NULL))
    ),
    selected = selected
  )
}

radioButtonsExtraServer <- function(id) {
  moduleFunction(id, function(input, output, server) {
    observeEvent(input$primary, {
      req(input$other)
      updateRadioButtons(session, "primary", selected = "other")
    })
    
    reactive({
      if (input$primary == "other") {
        input$other
      } else {
        input$primary
      }
    })
  })
}

ui <- fluidPage(
  radioButtonsExtraUI("gender", 
    label = "Gender", 
    choices = list(
      male = "Male",
      female = "Female",
      na = "Prefer not to say"
    ), 
    placeholder = "Self-described", 
    selected = "na"
  ),
  textOutput("txt")
)

server <- function(input, output, session) {
  gender <- radioButtonsExtraServer("gender")

  output$txt <- renderText({
    paste("You chose", gender())
  })
}
```

### Hierarchical select boxes

```{r, eval = FALSE}
library(tidyverse)
country_df <- countrycode::codelist %>% 
  as_tibble() %>% 
  select(iso3c, continent, country = cow.name) %>% 
  filter(!is.na(continent), !is.na(country))

continents <- sort(unique(country_df$continent))

ui <- fluidPage(
  selectInput("continent", "Continent", choices = continents),
  selectInput("country", "Country", choices = NULL)
)

server <- function(input, output, session) {
  countries <- reactive({
    country_df[country_df$continent == input$continent, , drop = FALSE]
  })
  
  observeEvent(input$continent, {
    updateSelectInput(session, "country", choice = countries()$country)
  })
}

shinyApp(ui, server)
```

### Modal

<https://gist.github.com/hadley/8d9ee5ea7991b0e5c400320abb9468de>

### Returning multiple reactives

* Leave in a list. 
* Use zeallot

### Exercises

1. The following app plots user selected variables from the `msleep` dataset 
   for three different types of mammals (carnivores, omnivores, and herbivores), 
   with one tab for each type of mammal. Remove the redundancy in the 
   `selectInput()` definitions with the use of functions.

```{r, eval = FALSE}
library(tidyverse)

ui <- fluidPage(
  selectInput(inputId = "x",
              label = "X-axis:",
              choices = c("sleep_total", "sleep_rem", "sleep_cycle", 
                          "awake", "brainwt", "bodywt"),
              selected = "sleep_rem"),
  selectInput(inputId = "y",
              label = "Y-axis:",
              choices = c("sleep_total", "sleep_rem", "sleep_cycle", 
                          "awake", "brainwt", "bodywt"),
              selected = "sleep_total"),
  tabsetPanel(id = "vore",
              tabPanel("Carnivore",
                       plotOutput("plot_carni")),
              tabPanel("Omnivore",
                       plotOutput("plot_omni")),
              tabPanel("Herbivore",
                       plotOutput("plot_herbi")))
)

server <- function(input, output, session) {

  # make subsets
  carni <- reactive( filter(msleep, vore == "carni") )
  omni  <- reactive( filter(msleep, vore == "omni")  )
  herbi <- reactive( filter(msleep, vore == "herbi") )

  # make plots
  output$plot_carni <- renderPlot({
    ggplot(data = carni(), aes_string(x = input$x, y = input$y)) +
      geom_point()
  })
  output$plot_omni <- renderPlot({
    ggplot(data = omni(), aes_string(x = input$x, y = input$y)) +
      geom_point()
  })
  output$plot_herbi <- renderPlot({
    ggplot(data = herbi(), aes_string(x = input$x, y = input$y)) +
      geom_point()
  })

}

shinyApp(ui = ui, server = server)
```

2. Continue working with the same app from the previous exercise, and further 
   remove redundancy in the code by modularizing how subsets and plots are 
   created.
   
3. Suppose you have an app that is slow to launch when a user visits it. Can  
   modularizing your app code help solve this problem? Explain your reasoning.


## Single object modules

When many people encounter modules for the first time, they immediately attempt to rewrite into a single "thing", a module object that has both UI and server components. If that didn't occur to you, feel free to skip this last part of the chapter, because it explains why that approach (which occurs to many people) isn't a good fit for Shiny.

Shiny explicitly considered and rejected this design because it has some major limitations. For this to work, you have to define the module outside of the server and the UI. That means there's no way to make any of the inputs to the module reactive. 

In Shiny, UI and server are inherently disconnected; Shiny doesn't know which UI invocation belongs to which server session. You can see this pattern throughout Shiny: for example, `plotOutput()` and `renderPlot()` are connected only by shared ID. Writing modules as separate functions reflects that reality: they're distinct functions that are not connected other than through a shared ID.

To illustrate the problem, lets generalise the motivating example from the first part of the chapter so that the data frame is now a parameter:

```{r}
histogramUi <- function(id, df) {
  list(
    selectInput(NS(id, "var"), "Variable", names(df)),
    numericInput(NS(id, "bins"), "bins", 10, min = 1),
    plotOutput(NS(id, "hist"))
  )
}

histogramServer <- function(id, df) {
  moduleFunction(id, function(input, output, server) {
    data <- reactive(df[[input$var]])
    output$hist <- renderPlot({
      hist(data(), breaks = input$bins, main = input$var)
    })
  })
}
```

And that leads to the following app:

```{r}
ui <- fluidPage(
  tabsetPanel(
    tabPanel("mtcars", histgramUI("mtcars", mtcars)),
    tabPanel("iris", histogramUI("iris", iris))
  )
)
server <- function(input, output, session) {
  histgramServer("mtcars", mtcars)
  histgramServer("iris", iris)
}
```

It seems undesirable that we have to repeat both the ID and the name of the dataset in both the UI and server, so it's natural to want to wrap into a single function that returns both the UI and the server:

```{r}
histogramModule <- function(id, df) {
  list(ui = histogramUi(id, df), server = histogramServer(id, df))
}
```

Then we define the module outside of the UI and server, extracting elements from the list as needed:

```{r}
hist1 <- histogramModule("mtcars", mtcars)
hist2 <- histogramModule("iris", iris)

ui <- fluidPage(
  tabsetPanel(
    tabPanel("mtcars", hist1$ui()),
    tabPanel("iris", hist2$ui())
  )
)
server <- function(input, output, session) {
  hist1$server()
  hist2$server()
}
```

This works, but has a substantial drawback: what if we want to allow the user to select the dataset, i.e. we want to make the `df` argument reactive.


```{r}
ui <- function(id, df = NULL) {
  list(
    selectInput(NS(id, "var"), "Variable", if (!is.null(df)) names(df)),
    numericInput(NS(id, "bins"), "bins", 10, min = 1),
    plotOutput(NS(id, "hist"))
  )
}
server <- function(id, df) {
  moduleFunction(id, function(input, output, server) {
    observeEvent(df, {
      updateSelectInput(session, "var", choices = names(df()))
    })
    data <- reactive(df()[[input$var]])
    output$hist <- renderPlot({
      hist(data(), breaks = input$bins, main = input$var)
    })
  })
}
```

If you find your fingers itching to make a single module object that incorporates both UI and server components it's almost certainly a sign that you've decomposed your repeated code into a module in a suboptimal way.
