# Introduction

```{r include = FALSE}
source("common.R")
```

## What is Shiny?

## What you will learn

If you've never used Shiny before, welcome! 

The first part of the book, "Getting started", is designed to get you up and running with Shiny as quickly as possible. This is pretty quick, because Shiny is carefully designed to let you get started writing apps as quickly as possible. We'll teach you a few recipes to follow that let you turn your existing R code into basic apps by wrapping key bits of code in `reactive({...})` or `renderXXX({...})`.

In "Shiny in Action", you'll learn a bunch of common patterns for building Shiny apps.

Then in "Mastering UI", you'll dive further into the many options you have for the front-end of your app.

In "Mastering reactivity", you'll go deep in to the theory and practice of reactive programming, the programming paradigm that underlies Shiny. If you're an existing Shiny user, you'll get the most value out of this chapter as it will give you a solid theoretical underpinning that will allow you to create new tools specifically tailored for your problems.

The biggest drawback of reactive programming is that it's a fundamentally new programming paradigm. Even experienced R users can have trouble getting their heads around reactive programming, and those with deep experience in software engineering may feel uncomfortable with so much "magic". But once you've formed an accurate mental model, you'll see that there's nothing up Shiny's sleeves: the magic comes from simple concepts combined in consistent ways.

Finally, in "Taming Shiny" we'll finish up a survey of useful techniques for making your Shiny apps work well in production. You'll learn how to measure and improve performance, debug problems when they go wrong, and manage your app's dependencies.

## What you won't learn

The focus of this book is making effective Shiny apps and understanding the underlying theory of reactivity. I'll do my best to show off best practices for data science, R programming, and software engineering at the same time as well, but you'll need other references to master these other important components.

## Prerequisites {#prerequisites}

The first thing you'll need to do is install the software you'll need, if you don't have it already:

### R {-}

If you don't have R installed already, you may be reading the wrong book; you need to be proficient with R if you want to write Shiny apps. If you'd like to learn how to use R, I'd recommend my [_R for Data Science_](https://r4ds.had.co.nz/) book which is designed to get you up and running with R with minimum of fuss.

### RStudio {-}

RStudio is a free and open source *integrated development environment* (IDE) for R: a single program that includes a code editor, R console, graphics device, and many features for working productively with R.

While you can write and use Shiny apps with any R environment (including R GUI and [ESS](http://ess.r-project.org)), RStudio has some nice features specifically for authoring, debugging, and deploying Shiny apps. We recommend giving it a try, but it's not required to be successful with Shiny or with this book.

You can download RStudio Desktop at <https://www.rstudio.com/products/rstudio/download>

### Shiny {-}

Shiny is an R package; you install it the same way you install any R package. From the R console:

```{r eval=FALSE}
install.packages("shiny")
```

### Other R packages {-}

This book is heavy on code, and we will often use other R packages in our examples. You can install them all now by running this code:

```{r eval=FALSE}
install.packages(c("magrittr", "lubridate", "readr", "dplyr", "ggplot2", "gt"))
```

TODO: Update this list before final draft; see [DESCRIPTION](https://github.com/hadley/mastering-shiny/blob/master/DESCRIPTION) for the definitive list.

## Cheat sheet

You may find it helpful to print a copy of our [Shiny "cheat sheet"](https://github.com/rstudio/cheatsheets/raw/master/shiny.pdf), a reference card for many of the most important concepts and functions in Shiny. It won't all make sense to you yet, but as you work through the chapters you'll find it more and more helpful to refresh your memory.

## If you get stuck {#reprex}

Do you have a problem you can't figure out, or a question that this book (and Google) can't answer? The best place to ask is our community site,  <https://community.rstudio.com/c/shiny>. 

To get the most useful help as quickly as possible, you will need to create a reprex, or **repr**oducible **ex**ample. The goal of a reprex is to provide the smallest possible snippet of R code that illustrates the problem and that can easily be run on someone else's computer. It's common courtesy (and in your own best interest) to create a reprex: if you want someone to help you, you should make it as easy as possible for them.

### Reprex basics

A reprex is just some R code that works when you copy and paste it into a fresh R session. Here's a simple reprex for a Shiny app:

```{r, eval = FALSE}
library(shiny)
ui <- fluidPage(
  selectInput("n", "N", 1:10),
  plotOutput("plot")
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    n <- input$n * 2
    plot(head(cars, n))
  })
}
shinyApp(ui, server)
```

Anyone can copy and paste this code into their R session and see the error: "non-numeric argument to binary operator". 

Clearly illustrating the problem is the first step to getting help, and because anyone can reproduce the problem by just copying and pasting the code, they can easily explore your code and test possible solutions. (In this case, you need `as.numeric(input$n)` since `selectInput()` creates a string.)

### Making a reprex

The first step in making a reprex is to create a single self-contained file that contains everything needed to run your code. You should check it by starting a fresh R session and then running the code. Make sure you haven't forgotten to load any packages that make your app work.

You'll also need to check for data that's only presented on your computer. You can make it easier for others by providing a data set in a way that can be copied and pasted. This can be either your original data set, or a simplified version of it that still causes the issue. For example, if you have a data frame with 1,000 rows, you might try to take the first 20 rows and see if that will still cause the problem. In general it is best to provide a simplified version of the data set, so that it is easy for others to understand it.

*   Use a built-in data set like `mtcars` or `iris`.

*   Provide regular R code that generates a data set. For example:

    ```{r}
    mydata <- data.frame(x = 1:5, y = c("a", "b", "c", "d", "e"))
    ```

*   Use `dput()` to turn an existing data frame into executable code.
    `dput(mydata)` generates the code you need to recreate `mydata`:
    
    ```{r}
    dput(mydata)
    ```
  
    Once you have that code, you can put this in your reprex to generate 
    `mydata`:
    
    ```{r}
    mydata <- structure(list(x = 1:5, y = structure(1:5, .Label = c("a", "b", 
    "c", "d", "e"), class = "factor")), class = "data.frame", row.names = c(NA, 
    -5L))
    ```

    If `dput()` on your original data generates too much code, try taking a 
    subset of your data and see if it will reproduce the problem.

### Making a minimal reprex

Creating the smallest possible reprex is particularly important for Shiny apps, which can be quite large and complicated. Rather than forcing the person trying to help you to understand all the details of your app, you are more likely to get higher quality help faster if you extract out the precise piece of the app that you're struggling with. As an added benefit, this process will often lead you to discover what the problem is, so you don't have to wait for help from someone else!

Reducing a bunch of code to the essential problem is a skill, and you probably won't be very good it at first. That's ok! Even the smallest reduction in code complexity helps the person helping you, and over time your compexity reduction skills will approve.

If you don't know what part of your code is triggering the problem, a good way to find it is to remove sections of code from your application, piece by piece, until the problem goes away. If removing a particular piece of code makes the problem stop, it's likely that that code is related to the problem. Once you've isolated the code that triggers the problem, you can go on to the next step.

To create a minimal example, you can go one of two ways: you can either keep removing code from your application until all the extraneous parts are gone, or you can construct a new Shiny application and copy the problem code into the new application. It's generally easier for others to understand the problem when you construct a new example, although either way is much better than not having a reproducible example at all.

### Case study

<!-- 
https://community.rstudio.com/t/error-in-eval-object-variable-not-found-data-table-feeding-plot-in-shiny-app/38517/8
-->

## Acknowledgements

This book was written in [bookdown](http://bookdown.org/) inside [RStudio](http://www.rstudio.com/ide/). The [website](http://mastering-shiny.org/) is hosted with [netlify](http://netlify.com/), and automatically updated after every commit by [travis-ci](https://travis-ci.org/). The complete source is available from [GitHub](https://github.com/hadley/mastering-shiny). 

```{r, echo = FALSE}
ruler <- function(width = getOption("width")) {
  x <- seq_len(width)
  y <- dplyr::case_when(
    x %% 10 == 0 ~ as.character((x %/% 10) %% 10),
    x %% 5 == 0  ~ "+",
    TRUE         ~ "-"
  )
  cat(y, "\n", sep = "")
  cat(x %% 10, "\n", sep = "")
}
```
```{r}
ruler()
```
