# Introduction

```{r include = FALSE}
source("common.R")
```

## What is Shiny?

If you've never used Shiny before, welcome! Shiny is an R package that allows you to easily create rich, interactive, web apps. Shiny allows you to take your work in R and expose it via a web browser so that anyone can use it. Shiny makes you look amazing because it allows you to create a polished web app with a minimum amount of pain.

People use Shiny to:

* Create dashboards that track important performance indicators, and faciliate
  drill down into suprising results.

* Replace hundreds of pages of PDFs with interactive apps that allow the 
  user to jump to the exact data slice they care about.
  
* Communicate complex models to a lay audience by exposing their behaviour
  in an interactive sandbox.

* Provide self serve data analysis for common workflows. Instead of 
  
* Teach statistics concepts through interactive widgets.

Web apps are traditionally hard to write because:

* You need a deep knowledge of web technologies like HTML, CSS and javascript. 

* Responding to user interactions requires carefully thought out code so that 
  interactive parts of the app update if, and only if, the values they depend
  on change.
  
Shiny helps with both these problems by:

* Providing a carefully curated set of user interface (or **UI** for short)
  functions that generate the HTML, CSS, and javascript that you need for 
  common tasks. You don't need to learn about HTML/CSS/JS until you want to
  go beyond the basics that Shiny provides for you.
  
* Introducing a new style of programming called __reactive programming__ that 
  automatically tracks the dependencies of a code chunk. This means that 
  Shiny automatically tracks the dependencies of your code, so that whenever
  an input change, Shiny can do the minimal amount of work to update the 
  related outputs.

By solving these problems, Shiny makes it possible for you, the R user, to create rich, interactive web applications without having to involve another team or other people with web programming expertise. This substantially shortens the iteration time.

## What will you learn?

The book is divided into five parts:

1.  In "Getting started", you'll learn the basics of Shiny so you can get up and 
    running as quickly as possible. You'll learn about the basic the app 
    structure, useful UI components, and the basics of reactive programming.
    
1.  "Shiny in action" builds on the basics to help you solve common problems
    including giving feedback to the user, uploading and downloading data,
    generate UI with code, reducing duplication, and using Shiny with the 
    tidyverse.
    
1.  "Mastering UI" dives into the details of the user interface. You'll learn
    about some of the most important extension packages, and then learn the
    basics of HTML and CSS so that you can customise Shiny to precisely meet 
    your needs.
    
1.  In "Mastering reactivity", you'll go deep into the theory and practice of 
    reactive programming, the programming paradigm that underlies Shiny. If 
    you're an existing Shiny user, you'll get the most value out of this 
    chapter as it will give you a solid theoretical underpinning that will 
    allow you to create new tools specifically tailored for your problems.

1.  Finally, in "Taming Shiny" we'll finish up a survey of useful techniques 
    for making your Shiny apps work well in production. You'll learn how to 
    measure and improve performance, debug problems when they go wrong, and 
    manage your app's dependencies.

## What won't you learn?

The focus of this book is making effective Shiny apps and understanding the underlying theory of reactivity. I'll do my best to show off best practices for data science, R programming, and software engineering at the same time as well, but you'll need other references to master these other important components. If you enjoy my writing, you might enjoy my books on these topics: [R for data science](http://r4ds.had.co.nz/), [Advanced R](http://adv-r.hadley.nz/), and [R packages](http://r-pkgs.org/).

## Prerequisites {#prerequisites}

The first thing you'll need to do is install the software you'll need, if you don't have it already:

### R {-}

If you don't have R installed already, you may be reading the wrong book; you need to be proficient with R if you want to write Shiny apps. If you'd like to learn how to use R, I'd recommend my [_R for Data Science_](https://r4ds.had.co.nz/) book which is designed to get you up and running with R with minimum of fuss.

### RStudio {-}

RStudio is a free and open source *integrated development environment* (IDE) for R: a single program that includes a code editor, R console, graphics device, and many features for working productively with R.

While you can write and use Shiny apps with any R environment (including R GUI and [ESS](http://ess.r-project.org)), RStudio has some nice features specifically for authoring, debugging, and deploying Shiny apps. We recommend giving it a try, but it's not required to be successful with Shiny or with this book.

You can download RStudio Desktop at <https://www.rstudio.com/products/rstudio/download>

### R packages {-}

This book is heavy on code, and we will often use other R packages in our examples. You can install them all now by running this code:

```{r, echo = FALSE}
deps <- desc::desc_get_deps()
pkgs <- sort(deps$package[deps$type == "Imports"])
pkgs2 <- strwrap(paste(encodeString(pkgs, quote = '"'), collapse = ", "), exdent = 2)

install <- paste0(
  "install.packages(c(\n  ", 
  paste(pkgs2, "\n", collapse = ""), 
  "))"
)
```

```{r code = install, eval=FALSE}
```

## Acknowledgements

This book was written in [bookdown](http://bookdown.org/) inside [RStudio](http://www.rstudio.com/ide/). The [website](http://mastering-shiny.org/) is hosted with [netlify](http://netlify.com/), and automatically updated after every commit by [travis-ci](https://travis-ci.org/). The complete source is available from [GitHub](https://github.com/hadley/mastering-shiny). 

```{r, echo = FALSE}
ruler <- function(width = getOption("width")) {
  x <- seq_len(width)
  y <- dplyr::case_when(
    x %% 10 == 0 ~ as.character((x %/% 10) %% 10),
    x %% 5 == 0  ~ "+",
    TRUE         ~ "-"
  )
  cat(y, "\n", sep = "")
  cat(x %% 10, "\n", sep = "")
}
```
```{r}
ruler()
```
