# Why reactivity?

```{r setup, include=FALSE}
source("common.R")
```

## Introduction

## Why reactive programming? {#motivation}

Reactive programming is a style of programming that emphasizes values that change over time, and calculations and actions that depend on those values. 

Shiny apps are interactive: users change input controls (sliders, textboxes, checkboxes), which causes logic to run on the server (reading CSVs or performing database queries, subsetting data, running simulations or models), ultimately resulting in outputs updating (plots, tables, blocks of text, downloadable PDFs). For Shiny apps to be useful, we need two things:

* To ensure that input and output stay in sync, all expressions and outputs 
  that depend on an input value must be automatically updated every time it
  changes.
  
* To ensure that apps respond quickly to user input, _only_ the relevant 
  expressions should run, and _only_ the affected outputs should update. It'd 
  be a frustrating experience if, say, changing a plot y-axis from linear to 
  log scale caused the expensive logic behind a totally unrelated summary 
  table to be rerun.

It's relatively easy to satisify one of the two conditions, but much harder to satisfy both. To see why, and to see how we might attack the basic problem with other styles of programming lets take it back to a very simple example, eliminating all additional complexity of a web app, and focussing on the underlying code.

### Why can't you use variables?

In one sense, all programmers are used to "values that change over time": they're called "variables". And yes, variables in R represent values and they can change over time. But they're not particularly designed to help you when their values change. Take this simple example of converting a temperature in Celsius to Fahrentheit:

```{r}
temp_c <- 10
temp_f <- (temp_c + 32) * 9 / 5 
temp_f
```

So far so good: the `temp_c` variable has the value `10`, and the `temp_f` variable has the value 75.6. But changing `temp_c` does not affect `temp_f`:

```{r}
temp_c <- 30
temp_f
```

Variables can change over time, but they never change automatically as 

### What about functions?

You could instead attack this problem with a function:

```{r}
temp_c <- 10
temp_f <- function() {
  message("Converting") 
  (temp_c + 32) * 9 / 5 
}
temp_f()

temp_celsius <- -3
temp_f() 
```

(This is a slightly weird function because it doesn't have any arguments, but instead accesses `temp_c` from its enclosing environment, but it's perfectly valid R code.)

This solves the first problem that reactivity is trying to solve: whenever you access `temp_f()` you get the latest computation. It doesn't, however solve the second problem of trying to do as little computation as possible. Every time you call `temp_f()` it recomputes:

```{r}
temp_f() 
```

That isn't a problem here in this trivial example, but minimising recomputation is a substantial challenge in real apps.

### Event-driven programming {#event-driven}

Since neither of the built-in tools of variables or functions work, we have to create something new. In previous decades, there wouldn't have been any question about what we should add: we'd implement something to support _event-driven programming_. It's an appealingly simple paradigm: you register callback functions to be executed in response to events (e.g. a mouse click, or a textbox's value being changed). 

```{r, echo = FALSE}
DynamicValue <- R6::R6Class("DynamicValue", list(
  value = NULL,
  on_update = NULL,
  get = function() self$value,
  set = function(value) {
    self$value <- value
    if (!is.null(self$on_update)) 
      self$on_update()
    invisible(self)
  },
  onUpdate = function(action) {
    self$on_update <- rlang::as_function(rlang::enquo(action))
    invisible(self)
  }
))
```

Had Shiny been invented five years earlier, we might have written something like this:

```{r}
temp_c <- DynamicValue$new()
temp_c$onUpdate({
  message("Converting") 
  temp_f <<- (temp_c$get() + 32) * 9 / 5 
})

temp_c$set(10)
temp_f

temp_c$set(-3)
temp_f
```

Note that we can no longer rely on an ordinary variable, and instead have to create an (imaginary) `DynamicValue`. This object has three important methods: `get()` and `set()` to access and change the underlying value, and `onUpdate()` lets you register code to run every time the value is modified. Here we're using `<<-` to update the value of `temp_f()` in the global environment every time the temperature changes.

Unfortunately, this kind of programming is _simple_, not _easy_! As your application adds more features, it becomes very difficult to keep track of what inputs affect what calculations, and what calculations affect each other, and what input and calculations affect what outputs and actions. Before long, you start to trade off correctness (just update everything whenever anything changes) against performance (try to update only the necessary parts, and pray you didn't miss any edge cases), because it's so difficult to optimize for both.

### Reactive programming

Reactive programming elegantly solves both problems by combining features of each of the solutions above. Now we can show you some real Shiny code. We're going to do this using a special Shiny mode, `consoleReactive(TRUE)`, enabling reactivity in the console so you that you can experiment with it directly, outside of an app. This mode is isn't enabled by default because it makes a certain class of bug harder to spot in an app, and it's primary benefit is to help you understand reactivity.

```{r, cache = FALSE}
library(shiny)
consoleReactive(TRUE)
```

As with event-driven programming, we still need some way to indicate that we have a special dynamic variable: `shiny::reactiveVal()`. This creates a single __reactive value__ that has a special syntax for getting and setting its value[^mutable-cell]. You use it much like a reactive expression, but it has the additional property that you can set its value.

[^mutable-cell]: If you've programmed in languages like rust or clojure this might look familiar: this is very similar to a mutable cell.

```{r}
temp_c <- reactiveVal(10)
temp_c()   # get
temp_c(20) # set
temp_c()   # get
```

Now we can create a reactive expression that depends on this value. As you've seen previously, a reactive expression automatically tracks all of its dependencies:

```{r}
temp_f <- reactive({
  message("Converting") 
  (temp_c() + 32) * 9 / 5 
})
temp_f()
```

So that later if `temp_celsius` changes, `is_freezing()` will be up to date:

```{r}
temp_c(-3)
temp_f()
temp_f()
```

Notice that the conversion only happens if we request the value of `temp_f()`, and the computation happens only once. Shiny strives to be as lazy as possible! In the next chapter you'll learn precisely how this dependency tracking works.

## A brief history of reactive programming 

<!--
https://www.cocoawithlove.com/blog/reactive-programming-what-and-why.html#appendix-a-little-bit-of-history
-->

The chances are you've already actually done a bunch of reactive programming outside of R already: 

Spreadsheets. Original and most popular instantiation of reactivity. When you update a cell all of it dependencies automatically update.

FRP, 1997. Mostly of research interest.

Reactive programming roared into the programming mainstream in the 2010s, via the blisteringly fast-paced world of JavaScript UI frameworks. Pioneering frameworks like [Knockout](https://knockoutjs.com/), [Ember](https://emberjs.com/), and (Joe Cheng's personal inspiration) Meteor demonstrated that reactive programming could make UI programming dramatically easier. Within a few short years, reactive programming has come to dominate UI programming on the web, with hugely popular frameworks like React, Vue.js, and Angular that are either inherently reactive or designed to work with reactive backends.

Before we go further, please be aware that the term "reactive programming" is a fairly general term in the programming world. While all reactive programming libraries/frameworks/languages are broadly about writing programs that respond to changing values, they vary enormously in their terminology, designs, and implementations.

In this book, whenever we refer to "reactive programming", we are referring specifically to reactive programming as implemented in Shiny. Conversely, if you read any material about reactive programming that isn't specifically about Shiny, it's unlikely that those concepts or even terminology will be relevant to writing Shiny apps. For readers who do have some experience with other reactive programming frameworks, our approach is similar to [Meteor](https://www.meteor.com/) and [MobX](https://mobx.js.org/), and very different than the [ReactiveX](http://reactivex.io/) family or anything that labels itself Functional Reactive Programming (often called FRP for short).
