# Reactive components {#reactivity-components}

```{r setup, include=FALSE}
source("common.R")
```

## Introduction

## Creating components

<https://github.com/hadley/shinySignals/>

### Isolate

Generally, you don't need to use `reactiveValues()` yourself.
But then it can be useful for achieving specific types of coordination that would otherwise be inaccessible, particularly maintaining state.
You need to be extra careful when reading to and writing from `reactiveValues()` because unlike the rest of Shiny, there's nothing to stop you from getting caught in an infinite loop.

```{r}
count <- function(signal) {
  val <- reactiveVal(0)
  
  observe({
    signal()
    val(val() + 1)
  })

  reactive(val())
}
```

```{r}
count <- function(signal) {
  val <- reactiveVal(0)
  
  observeEvent(signal(), {
    val(val() + 1)
  })

  reactive(val())
}
```

```{r}
count <- function(signal) {
  val <- reactiveVal(0)
  
  observe({
    signal()
    val(isolate(val()) + 1)
  })

  reactive(val())
}
```

Whenever you modify a reactive value based on a previous value you'll need to use isolate to avoid getting stuck in an infinite loop.

### Temporal dependence

```{r}
dropRepeats <- function(signal) {
  val <- reactiveVal()

  observe({
    cur_val <- isolate(val())
    new_val <- signal()
    if (!identical(cur_val, new_val)) {
      val(new_val)
    }
  })
  
  reactive(val)
}
```

<https://plotly-r.com/linking-views-with-shiny.html#reactive-vals>

```{r}
brush <- function(n, mode = "|") {
  selected <- reactiveVal(rep(FALSE, n))
  
  list(
    brush = function(new) {
      selected(do.call(mode, isolate(selected()), new))
    },
    reset = function() {
      selected(rep(FALSE, n))
    },
    selected = reactive(selected)
  )
}
```

### Combining reactives

```{r}
merge <- function(signal1, signal2) {
  val <- reactiveVal()

  observe(val(signal1()))
  observe(val(signal2()))

  reactive(val())
}
```
