# Why reactivity?

```{r setup, include=FALSE}
source("common.R")
```

## Introduction

The initial impression of Shiny is often that it's "magic".
This is great when you get started because you can make simple apps very very quickly.
But magic in software usually leads to disillusionment.
Without a solid mental model to reason with, it's extremely difficult to predict how the software will act when you venture beyond the borders of its demos and examples.
And when things don't go the way you expect, debugging is almost impossible.
Even experienced R users can have trouble getting their heads around reactive programming, and those with deep experience in software engineering may feel uncomfortable with so much "magic".

Fortunately shiny is "good" magic.
As Tom Dale said of his Ember.js JavaScript framework: "We do a lot of magic, but it's *good magic*, which means it decomposes into sane primitives." This is the quality that the Shiny team aspires to for Shiny, especially when it comes to reactive programming.
When you peel back the layers of reactive programming, you won't find a pile of heuristics, special cases, and hacks; instead you'll find a clever, but ultimately fairly straightforward mechanism.
Once you've formed an accurate mental model, you'll see that there's nothing up Shiny's sleeves: the magic comes from simple concepts combined in consistent ways.

In this part of the book, you'll dive into the theory of reactivity:

-   In this chapter, you'll learn why reactivity is needed, and a little bit about the history of reactive programming outside of R.

-   In Chapter \@ref(reactive-graph), you'll learn exactly how and when reactive components are updated.

-   Chapter \@ref(reactivity-scoping) goes into the details of object lifetimes including how you can share data across users, and how and why you might create an app within a function.

-   In Chapter \@ref(reactivity-components), , and how you can compose them to solve new problems.

You certainly don't need to understand the details of the theory for routine development of Shiny apps.
But understanding what's going on will help you create more complex apps and give you greater powers to extract out repeated patterns into reusable functions.

## Why reactive programming? {#motivation}

Reactive programming is a style of programming that focuses on values that change over time, and calculations and actions that depend on those values.
This is important for Shiny apps because they're interactive: users change input controls (dragging sliders, typing in textboxes, and checking checkboxes) which causes logic to run on the server (reading CSVs, subsetting data, and fitting models) ultimately resulting in outputs updating (plots replotting, tables updating).

For Shiny apps to be useful, we need two things:

-   Expressions and outputs should update whenever one of their input values changes.
    This ensures that input and output stay in sync.

-   Expressions and outputs should update *only* when one of their inputs changes.
    This ensures that apps respond quickly to user input, doing the minimal amount.

It's relatively easy to satisfy one of the two conditions, but much harder to satisfy both.
To see why, and to see how we might attack the basic problem with other styles of programming we'll use a very simple example, eliminating the additional complexity of a web app, and focussing on the underlying code.

### Why can't you use variables?

In one sense, you already know how to handle "values that change over time": they're called "variables".
Variables in R represent values and they can change over time, but they're not designed to help you when they change.

Take this simple example of converting a temperature from Celsius to Fahrenheit:

```{r}
temp_c <- 10
temp_f <- (temp_c * 9 / 5) + 32
temp_f
```

So far so good: the `temp_c` variable has the value `10`, the `temp_f` variable has the value 75.6, and we can change `temp_c`:

```{r}
temp_c <- 30
```

But changing `temp_c` does not affect `temp_f`:

```{r}
temp_f
```

Variables can change over time, but they never change automatically.

### What about functions?

You could instead attack this problem with a function:

```{r}
temp_c <- 10
temp_f <- function() {
  message("Converting") 
  (temp_c * 9 / 5) + 32
}
temp_f()
```

(This is a slightly weird function because it doesn't have any arguments, but instead accesses `temp_c` from its enclosing environment, but it's perfectly valid R code.)

This solves the first problem that reactivity is trying to solve: whenever you access `temp_f()` you get the latest computation:

```{r}
temp_c <- -3
temp_f() 
```

It doesn't, however, solve the second problem of trying to do as little computation as possible.
Every time you call `temp_f()` it recomputes:

```{r}
temp_f() 
```

That isn't a problem in this trivial example, but minimising re-computation is a substantial challenge in real apps.

### Event-driven programming {#event-driven}

Since neither variables nor functions work, we need to create something new.
In previous decades, there wouldn't have been any question about what we should create: we'd implement something to support *event-driven programming*.
Event-driven programming is an appealingly simple paradigm: you register callback functions to be executed in response to events (e.g. a mouse click, or a textbox's value being changed).

We could implement a very simple event-driven toolkit using R6.
A `DynamicValue` has three important methods: `get()` and `set()` to access and change the underlying value, and `onUpdate()` lets you register code to run every time the value is modified.
If you're not familiar with R6, don't worry about it, and instead focus on the interface as shown in the next examples.

```{r}
DynamicValue <- R6::R6Class("DynamicValue", list(
  value = NULL,
  on_update = NULL,

  get = function() self$value,

  set = function(value) {
    self$value <- value
    if (!is.null(self$on_update)) 
      self$on_update(value)
    invisible(self)
  },
  
  onUpdate = function(on_update) {
    self$on_update <- on_update
    invisible(self)
  }
))
```

So if Shiny had been invented five years earlier, it might have looked more like this:

```{r}
temp_c <- DynamicValue$new()
temp_c$onUpdate(function(value) {
  message("Converting") 
  temp_f <<- (value * 9 / 5) + 32
})

temp_c$set(10)
temp_f

temp_c$set(-3)
temp_f
```

Now `temp_c` is a dynamic value that uses `<<-` to automatically update `temp_f` whenever it changes.

Unfortunately, this kind of programming is *simple*, not *easy*!
As your application adds more features, it becomes very difficult to keep track of what inputs affect what calculations, and what calculations affect each other, and what input and calculations affect what outputs and actions.
Before long, you start to trade off correctness (just update everything whenever anything changes) against performance (try to update only the necessary parts, and hope that you didn't miss any edge cases) because it's so difficult to do both.

## Reactive programming

Reactive programming elegantly solves both problems by combining features of the solutions above.
Now we can show you some real Shiny code, using a special Shiny mode, `reactiveConsole(TRUE)`, that enable reactivity in the console so you can experiment with it directly.
This mode isn't enabled by default because it makes a certain class of bugs harder to spot in an app, and its primary benefit is to help you understand reactivity.
(It's only available in Shiny 1.6.0 so you may need to upgrade if it's not available for you.)

```{r, cache = FALSE}
library(shiny)
reactiveConsole(TRUE)
```

As with event-driven programming, we need some way to indicate that we have a special special type of variable.
In Shiny, we use a **reactive value**, which has a special syntax[^reactivity-motivation-1] for getting and setting its value.
We create the object with `shiny::reactiveVal()`, get its value by calling it like a zero-argument function, and set its value by calling it like a one-argument function[^reactivity-motivation-2]
.

[^reactivity-motivation-1]: If you've programmed in languages like rust or clojure this might look familiar: a reactive value is very similar to a mutable cell.

[^reactivity-motivation-2]: If you've ever used R's active bindings, you might notice that the syntax is the very similar.
    This is not a coincidence.

```{r}
temp_c <- reactiveVal(10) # create
temp_c()                  # get
temp_c(20)                # set
temp_c()                  # get
```

Now we can create a reactive expression that depends on this value.
As you've learned when creating apps, a reactive expression automatically tracks all of its dependencies:

```{r}
temp_f <- reactive({
  message("Converting") 
  (temp_c() * 9 / 5) + 32
})
temp_f()
```

So that later, if `temp_c` changes, `temp_f` will automatically update:

```{r}
temp_c(-3)
temp_c(-10)
temp_f()
temp_f()
```

The reactive expression has two important properties that you can verify by looking at when "converting" is printed:

-   It's **lazy**: it doesn't do any work until it's called.

-   It's **cached**: it doesn't do any work the second and subsequent times its called because it caches the previous result.

### Reactive values

You might wonder why it's taken so long to mention `reactiveVal()` --- we've done 20+ chapters of Shiny programming without needing it more than a handful of places.
That's because you most often work with a special `reactiveValues()` object:

```{r}
vals <- reactiveValues(a = 1, b = 2)
vals$a
vals$a <- 100
vals$a
```

From the outside `vals` appears to be a regular list.
But if you refer to the values inside a reactive expression, it will automatically update when the values change:

```{r}
sum <- reactive({
  message("Summing")
  vals$a + vals$b
})
sum()

vals$a <- 10
sum()
```

This might look familiar --- ever app you've written works with `input` in this way.
That's because `input` is a special type of read-only `reactiveValues()`.
You're not allowed to modify the values because Shiny automatically updates them based on the users action in the browser.

It's unfortunate that you have to use different syntax to get and set the value of a standalone reactive value and a reactive value in a list.
Unfortunately there's no easy way to make them behave the same, and since you `inputs` frequently and `reactiveVal()` rarely, the decision was to make `input` as familiar as possible.

### Observers

Along with reactive values and expressions, there is one more fundamental building block of reactivity: **observers**.
An observer sets up a block of code that is run every time one of the reactive values or expressions it uses is updated:

```{r}
y <- reactiveVal(10)
observe({
  message("`y` is ", y())
})

y(5)
y(4)
```

Observers differ from expressions in a couple of important ways:

-   They are designed to work with functions that change the world in some way, like `print()`, `plot()`, or `write.csv()`.
    In programming terminology, changing the world is called a **side-effect**.
    Unlike pharmaceuticals where side effects are always unintentional and usually negative, we simply mean any effects apart from a function's return value.

-   They are eager and forgetful --- they run as soon as possibly can and they don't remember their previous action.
    This eagerness is "infectious" because if they use a reactive expression, that reactive expression will also be evaluated immediately.

Observers are often coupled with reactive values in order to track state changes over time.
But if you're not careful, you quickly hit a problem.
For example, take this code which tracks how many times `x` changes:

```{r, eval = FALSE}
count <- reactiveVal(0)
x <- reactiveVal(1)
observe({
  x()
  count(count() + 1)
})
```

If you were to run it, you'd immediately get stuck in an infinite loop because the observer will take a reactive dependency on `x` **and** `count`; and since it modifies `count`, it will immediately re-run.

Fortunately, Shiny provides a function to resolve this problem: `isolate()`.
This function allows you to access the current value of a reactive value or expression without taking a dependency on it:

```{r}
count <- reactiveVal(0)
x <- reactiveVal(1)
observe({
  x()
  isolate(count(count() + 1))
})

x(1)
x(2)
count()
x(3)
count()
```

Another way to avoid this problem is to use the `observeEvent()` function that you learned about in Section \@ref(observers).
It decouples listening from reaction, so you could rewrite the above as:

```{r, eval = FALSE}
observeEvent(x(), {
  count(count() + 1))
})
```

In Chapter \@ref(reactivity-components) we'll come back to the problem of managing state over time in much more detail.

### Outputs {#outputs-theory}

If you've being paying close attention you might have noticed that I've explained `input`, reactive expressions, and observer, but I haven't said anything about `output`.
How do they fit into the picture?
For the purposes of discussion, take this simple output:

```{r eval = FALSE}
output$text <- renderText({
  paste(up_to_x(), collapse = ", ")
})
```

Is this an observer or a reactive expression?
It looks like a reactive expression because we're assigning the result of `renderText()`.
On the other hand, it feels like a observer, because it has a side-effect: updating the contents of a text box.
It turns out that outputs are neither reactive expressions nor observers, and indeed aren't even a fundamental building block.
In fact, they're a Shiny feature built on top of the foundations that behave somewhat in between reactive expressions and observers.

Reactive expressions, observers, and outputs are all reactive consumers.
They differ primarily in their execution model: observers are eager, reactive expressions are lazy, and outputs are somewhere in between.
When an output's UI element is visible in the browser, outputs execute eagerly; that is, once at startup, and once anytime their relevant inputs or reactive expressions change.
However, if the UI element is hidden then Shiny will automatically suspend (pause) that output from reactively executing[^reactivity-motivation-3].

[^reactivity-motivation-3]: In rare cases, you may prefer to process even outputs that aren't hidden.
    You can use the `outputOptions()` function's `suspendWhenHidden` to opt out of the automatic suspension feature on an output-by-output basis.

Outputs also have a somewhat complicated relationship with side-effects, particularly printing and plotting because `renderPrint()` and `renderPlot()` capture printing and plotting side-effects and convert them into special values that can be sent to the browser.
Apart from these special side-effects handled by matching render functions, outputs should be free of side-effects, as you have no guarantee of when they'll be performed.

## A brief history of reactive programming

If you want to learn more about reactive programming in other languages, a little history can be helpful.
You can see the genesis of reactive programming over 40 years ago in [VisiCalc](https://en.wikipedia.org/wiki/VisiCalc), the first spreadsheet:

> I imagined a magic blackboard that if you erased one number and wrote a new thing in, all of the other numbers would automatically change, like word processing with numbers.
> --- [Dan Bricklin](https://youtu.be/YDvbDiJZpy0)

Spreadsheets are closely related to reactive programming: you declare the relationship between cells (using formulas), and when one cell changes, all of its dependencies automatically update.
So the chances are that you've already done a bunch of reactive programming without knowing it!

While the ideas of reactivity have been around for a long time, it wasn't until 1997 before they were seriously studied in academic computer science.
Research in reactive programming was kicked off by FRAN [@fran], **f**unctional **r**eactive **an**imation, a novel system for incorporating changes over time and user input into a functional programming language.
This spawned a rich literature [\@rp-survey], but had little impact on the practice of programming.

It wasn't until the 2010s that reactive programming roared into the programming mainstream via the blisteringly fast-paced world of JavaScript UI frameworks.
Pioneering frameworks like [Knockout](https://knockoutjs.com/), [Ember](https://emberjs.com/), and (Joe Cheng's personal inspiration) [Meteor](https://www.meteor.com) demonstrated that reactive programming could make UI programming dramatically easier.
Within a few short years, reactive programming has come to dominate UI programming on the web, with hugely popular frameworks like [React](https://reactjs.org), [Vue.js](https://vuejs.org), and [Angular](https://angularjs.org) which are either inherently reactive or designed to work hand-in-hand with reactive backends.

It's worth bearing in mind that "reactive programming" is a fairly general term.
While all reactive programming libraries, frameworks, and languages are broadly about writing programs that respond to changing values, they vary enormously in their terminology, designs, and implementations.
In this book, whenever we refer to "reactive programming", we are referring specifically to reactive programming as implemented in Shiny.
Conversely, if you read any material about reactive programming that isn't specifically about Shiny, it's unlikely that those concepts or even terminology will be relevant to writing Shiny apps.
For readers who do have some experience with other reactive programming frameworks, Shiny's approach is similar to [Meteor](https://www.meteor.com/) and [MobX](https://mobx.js.org/), and very different than the [ReactiveX](http://reactivex.io/) family or anything that labels itself Functional Reactive Programming.
