# Reactive building blocks {#reactivity-objects}

```{r setup, include=FALSE}
source("common.R")
```

Now that you understand the theory underpinning the reactive graph and you have some practical experience, it is a good time to talk in more detail about how reactivity fits into R the programming language.
There are three fundamental building blocks of reactive programming: reactive values, reactive expressions, and observers.
You've already seen most of the important parts of reactive values and expressions, so this chapter will spend more time on observers and outputs (which as you'll learn are a special type of observer).
You'll also learn two other tools for controlling the reactive graph: isolation and timed invalidation.

This chapter will again use the reactive console so that we can experiment with reactivity directly in the console without having to launch a Shiny app each time.

```{r}
library(shiny)
reactiveConsole(TRUE)
```

## Reactive values

There are two types of reactive values:

-   A single reactive value, created by `reactiveVal()`.

-   A list of reactive values, created by `reactiveValues()`.

They have slightly different interfaces for getting and setting values:

```{r}
x <- reactiveVal(10)
x()       # get
x(20)     # set
x()       # get

r <- reactiveValues(x = 10)
r$x       # get
r$x <- 20 # set
r$x       # get
```

It's unfortunate that these two similar objects have rather different interfaces, but there's no way to standardise them.
However, while they look different, they behave the same, so you can choose between them based on which syntax you prefer.
In this book I use `reactiveValues()` because the syntax is easier to understand at a glance, but in my own code I tend to use `reactiveVal()` because the syntax makes it's clear that something weird is going on.

It's important to note that both types of reactive values have so called reference semantics.
Most R objects have copy-on-modify[^reactivity-foundations-1] semantics which means that if you assign the same value to two names, the connection is broken as soon as you modify one:

[^reactivity-foundations-1]: See more details at <https://adv-r.hadley.nz/names-values.html#copy-on-modify>

```{r}
a1 <- a2 <- 10
a2 <- 20
a1 # unchanged
```

This is not the case with reactive values --- they always keep a reference back to the same value so that modifying any copy modifies all values:

```{r}
b1 <- b2 <- reactiveValues(x = 10)
b1$x <- 20
b2$x
```

We'll come back to why you might create your own reactive values in Chapter \@ref(reactivity-components).
Otherwise, most of the reactive values you'll encounter will come from the `input` argument to the server function.
These are a little different to the `reactiveValues()` that you create yourself because they're read-only: you're can't modify the values because Shiny automatically updates them based on user actions in the browser.

### Exercises

1.  What are the differences between these two lists of reactive values?
    Compare the syntax for getting and setting individual reactive values.

    ```{r}
    l1 <- reactiveValues(a = 1, b = 2)
    l2 <- list(a = reactiveVal(1), b = reactiveVal(2))
    ```

2.  Design and perform a small experiment to verify that `reactiveVal()` also has reference semantics.

## Reactive expressions

Recall that a reactive has two important properties: it's lazy and cached.
This means that it only does work when it's actually needed, and if called twice in a row, it returns the previous value.

There are two important details that we haven't yet covered: what reactive expressions do with errors, and why `on.exit()` works inside of them.

### Errors

Reactive expressions cache errors in exactly the same way that they cache values.
For example, take this reactive:

```{r, error = TRUE}
r <- reactive(stop("Error occured at ", Sys.time(), call. = FALSE))
r()
```

If we wait a second or two, we can see that we get the same error as before:

```{r, error = TRUE}
Sys.sleep(2)
r()
```

Errors are also treated the same way as values when it comes to the reactive graph: errors propagate through the reactive graph exactly the same way as regular values.
The only difference is what happens when an error hits an output or observer:

-   An error in an output will be displayed in the app[^reactivity-foundations-2].
-   An error in an observer will cause the current session to terminate. If you don't want this to happen, you'll need to wrap the code in `try()` or `tryCatch()`.

[^reactivity-foundations-2]: By default, you'll see the whole error message.
    You can show a generic error message by turning error sanitising on.
    See <https://shiny.rstudio.com/articles/sanitize-errors.html> for details.

This same system powers `req()` (Section \@ref(cancelling-execution-with-req)), which emits a special type of error[^reactivity-foundations-3].
This special error causes observers and outputs to stop what they're doing but not otherwise fail.
By default, it will cause outputs to reset to their initial blank state, but if you use `req(..., cancelOutput = TRUE)` they'll preserve their current display.

[^reactivity-foundations-3]: Technically, a custom condition.
    See <https://adv-r.hadley.nz/conditions.html#custom-conditions> for more details on the underlying theory.

### `on.exit()`

You can think of `reactive(x())` as a shortcut for `function() x()`, automatically adding laziness and caching.
This is mostly of importance if you want to understand how Shiny is implemented, but means that you can use functions that only work inside functions.
The most useful of these is `on.exit()` which allows you to run code when a reactive expression finishes, regardless of whether the reactive successfully returns an error or fails with an error.
This is what makes `on.exit()` work in Section \@ref(removing-on-completion).

### Exercises

1.  Use the reactlog package to observe an error propagating through the reactives in the following app, confirming that it follows the same rules as value propagation.

    ```{r}
    ui <- fluidPage(
      checkboxInput("error", "error?"),
      textOutput("result")
    )
    server <- function(input, output, session) {
      a <- reactive({
        if (input$error) {
          stop("Error!")
        } else {
          1
        }
      })
      b <- reactive(a() + 1)
      c <- reactive(b() + 1)
      output$result <- renderText(c())
    }
    ```

2.  Modify the above app to use `req()` instead of `stop()`.
    Verify that events still propagate the same way.
    What happens when you use the `cancelOutput` argument?

## Observers and outputs {#observers-details}

Observers and outputs are terminal nodes in the reactive graph.
They differ from reactive expressions in two important ways:

-   They are eager and forgetful --- they run as soon as they possibly can and they don't remember their previous action.
    This eagerness is "infectious" because if they use a reactive expression, that reactive expression will also be evaluated.

-   The value returned by an observer is ignored because they are designed to work with functions called for their side-effects, like `cat()` or `write.csv()`.

Observers and outputs are powered by the same underlying tool: `observe()`.
This sets up a block of code that is run every time one of the reactive values or expressions it uses is updated.
Note that the observer runs immediately when you create it --- it must do this in order to determine its reactive dependencies.

```{r}
y <- reactiveVal(10)
observe({
  message("`y` is ", y())
})

y(5)
y(4)
```

I rarely use `observe()` in this book, because it's the low-level tool that powers the user-friendly `observeEvent()`.
Generally, you should stick with `observeEvent()` unless it's impossible to get it to do what you want.
In this book, I'll only show you one case where `observe()` is necessary, Section \@ref(pausing-animations).

`observe()` also powers reactive outputs.
Reactive outputs are a special type of observers that have two important properties:

-   They are defined when you assign them into `output`, i.e. `output$text <- ...` creates the observer.

-   They have some limited ability to detect when they're not visible (i.e. they're in non-active tab) so they don't have to recompute[^reactivity-foundations-4].

[^reactivity-foundations-4]: In rare cases, you may prefer to process even outputs that are hidden.
    You can use the `outputOptions()` function's `suspendWhenHidden` to opt out of the automatic suspension feature on an output-by-output basis.

It's important to note that `observe()` and the reactive outputs don't "do" something, but "create" something (which then takes action as needed).
That mindset helps you to understand what's going on in this example:

```{r}
x <- reactiveVal(1)
y <- observe({
  x()
  observe(print(x()))
})
x(2)
x(3)
```

Each change to `x` causes the observer to be triggered.
The observer itself calls `observe()` setting up *another* observer.
So each time `x` changes, it gets another observer, so its value is printed another time.

As a general rule, you should only ever create observers or outputs at the top-level of your server function.
If you find yourself trying to nest them or create an observer inside an output, sit down and sketch out the reactive graph that you're trying to create --- there's almost certainly a better approach.
It can be harder to spot this mistake directly in a more complex app, but you can always use the reactlog: just look for unexpected churn in observers (or outputs), then track back to what is creating them.

## Isolating code

To finish off the chapter I will discuss two important tools for controlling exactly how and when the reactive graph is invalidated.
In this section, I'll discuss `isolate()`, the tool that powers `observeEvent()` and `eventReactive()`, and that lets you avoid creating reactive dependencies when not needed.
In the next section, you'll learn about `invalidateLater()` which allows you to generate reactive invalidations on a schedule.

### `isolate()`

Observers are often coupled with reactive values in order to track state changes over time.
For example, take this code which tracks how many times `x` changes:

```{r, eval = FALSE}
r <- reactiveValues(count = 0, x = 1)
observe({
  r$x
  r$count <- r$count + 1
})
```

If you were to run it, you'd immediately get stuck in an infinite loop because the observer will take a reactive dependency on `x` **and** `count`; and since the observer modifies `count`, it will immediately re-run.

Fortunately, Shiny provides `isolate()` to resolve this problem.
This function allows you to access the current value of a reactive value or expression **without** taking a dependency on it:

```{r}
r <- reactiveValues(count = 0, x = 1)
class(r)
observe({
  r$x
  r$count <- isolate(r$count) + 1
})

r$x <- 1
r$x <- 2
r$count

r$x <- 3
r$count
```

Like `observe()`, a lot of the time you don't need to use `isolate()` directly because there are two useful functions that wrap up the most common usage: `observeEvent()` and `eventReactive()`.

### `observeEvent()` and `eventReactive()`

When you saw the code above, you might have remembered Section \@ref(observers) and wondered why I didn't use `observeEvent()`:

```{r, eval = FALSE}
observeEvent(r$x, { 
  r$count <- r$count + 1 
}) 
```

And indeed, I could have because `observeEvent(x, y)` is equivalent to `observe({x; isolate(y)})`.
It elegantly decouples what you want to listen to from what action you want to take.
And `eventReactive()` performs the analogous job for reactives: `eventReactive(x, y)` is equivalent to `reactive({x; isolate(y)}).`

`observeEvent()` and `eventReactive()` have additional arguments that allow you to control the details of their operation:

-   By default, both functions will ignore any event that yields `NULL` (or in the special case of action buttons, 0).
    Use `ignoreNULL = FALSE` to also handle `NULL` values.

-   By default both functions will run once when you create them.
    Use `ignoreInit = TRUE` to skip this run.

-   For `observeEvent()` only, you can use `once = TRUE` to run the handler only once.

These are rarely needed but good to know about so that you can look up the details from the documentation when you need them.

### Exercises

1.  Complete the app below with a server function that updates `out` with the value of `x` only when the button is pressed.

    ```{r}
    ui <- fluidPage(
      numericInput("x", "x", value = 50, min = 0, max = 100),
      actionButton("capture", "capture"),
      textOutput("out")
    )
    ```

## Timed invalidation {#timed-invalidation-adv}

`isolate()` reduces the time the reactive graph is invalidated.
This topic of this section, `invalidateLater()` does the opposite: it lets you invalidate the reactive graph when no data has changed.
You saw an example of this in Section \@ref(timed-invalidation) with `reactiveTimer()`, but the time has come to discuss the underlying tool that powers it: `invalidateLater()`.

`invalidateLater(ms)` causes any reactive consumer to be invalidated in the future, after `ms` milliseconds.
It is useful for creating animations and connecting to data sources outside of Shiny's reactive framework that may be changing over time.
For example, this reactive will automatically generate 10 fresh random normals every half a second[^reactivity-foundations-5]:

[^reactivity-foundations-5]: Assuming that it's used by some output or observer; otherwise it will stay in its initial invalidated state forever.

```{r}
x <- reactive({
  invalidateLater(500)
  rnorm(10)
})
```

And this observer will increment a cumulative sum with a random number:

```{r, eval = FALSE}
sum <- reactiveVal(0)
observe({
  invalidateLater(300)
  sum(isolate(sum()) + runif(1))
})
```

In the following sections you'll learn how to use `invalidateLater()` to read changing data from disk, how to avoid getting `invalidateLater()` stuck in an infinite loop, and some occasionally important details of exactly when the invalidation happens.

### Polling

A useful application of `invalidateLater()` is to connect Shiny to data that is changing outside of R.
For example, you could use the following reactive to re-read a csv file every second:

```{r}
data <- reactive({
  on.exit(invalidateLater(1000))
  read.csv("data.csv")
})
```

This connects changing data into Shiny's reactive graph, but it has a serious downside: when you invalidate the reactive, you're also invalidating all downstream consumers, so even if the data is the same, all the downstream work has to be redone.

To avoid this problem, Shiny provides `reactivePoll()` which takes two functions: one that performs a relatively cheap check to see if the data has changed and another more expensive function that actually does the computation.
We can use `reactivePoll()` to rewrite the previous reactive as below.

```{r}
server <- function(input, output, session) {
  data <- reactivePoll(1000, session, 
    function() file.mtime("data.csv"),
    function() read.csv("data.csv")
  )
}
```

Here we used to `file.mtime()`, which returns the last time the file was modified, as a cheap check to see if we need to reload the file:

Reading a file when it changes is a common task, so Shiny provides an even more specific helper that just needs a file name and a reader function:

```{r}
server <- function(input, output, session) {
  data <- reactiveFileReader(1000, session, "data.csv", read.csv)
}
```

If you need to read changing data from other sources (e.g. a database), you'll need to come up with your own `reactivePoll()` code.

### Long running reactives

If you're performing a long running computation, there's an important question you need to consider: when should you execute `invalidateLater()`?
For example, take this reactive:

```{r, eval = FALSE}
x <- reactive({
  invalidateLater(500)
  Sys.sleep(1)
  10
})
```

Assume Shiny starts the reactive running at time 0, it will request invalidation at time 500.
The reactive takes 1000ms to run, so it's now time 1000, and it's immediately invalidated and must be recomputed, which then sets up another invalidation: we're stuck in an infinite loop.

On other hand, if you run `invalidateLater()` at the end, it will invalidate 500ms after completion, so the reactive will be re-run every 1500 ms.

```{r, eval = FALSE}
x <- reactive({
  on.exit(invalidateLater(500), add = TRUE)
  Sys.sleep(1)
  10
})
```

This is the main reason to prefer `invalidateLater()` to the simpler `reactiveTimer()` that we used earlier: it gives you greater control over exactly when the invalidation occurs.

### Timer accuracy

The number of milliseconds specified in `invalidateLater()` is a polite request, not a demand.
R may be doing other things when you asked for invalidation to occur, so your request has to wait.
This effectively means that the number is a minimum and invalidation might take longer than you expect.
In most cases, this doesn't matter because small differences are unlikely to affect user perception of your app.
However, in situations where many small errors will accumulate, you should compute the exact elapsed time and use it to adjust your calculations.

For example, the following code computes distance based on velocity and elapsed time.
Rather than assuming `invalidateLater(100)` always delays by exactly 100 ms, I compute the elapsed time and use it in my calculation of position.

```{r, eval = FALSE}
velocity <- 3
r <- reactiveValues(distance = 1)

last <- proc.time()[[3]]
observe({
  cur <- proc.time()[[3]]
  time <- last - cur
  last <<- cur
  
  r$distance <- isolate(r$distance) + velocity * time
  invalidateLater(100)
})
```

If you're not doing careful animation, feel free to ignore the inherent variation in `invalidateLater()`.
Just remember that it's a polite request, not a demand.

### Exercises

1.  Why will this reactive never be executed?
    Your explanation should talk about the reactive graph and invalidation.

    ```{r}
    server <- function(input, output, session) {
      x <- reactive({
        invalidateLater(500)
        rnorm(10)
      })  
    }
    ```

2.  If you're familiar with SQL, use `reactivePoll()` to only re-read an imaginary "Results" table whenever a new row is added.
    You can assume the Results table has a `timestamp` field that contains the date-time that a record was added.

## Summary {#how-it-works}

In this chapter you've learned more about the building blocks that make Shiny work: reactive values, reactive expressions, observers, and timed evaluation.
Now we'll turn our attention to a specific combination of reactive values and observers that allows us to escape some of the constraints (for better and worse) of the reactive graph.
