# Security {#scaling-security}

Most Shiny apps are deployed within a company firewall, and you can generally assume that your colleagues aren't going to try and hack your app[^scaling-safety-1].
If, however, you want to expose an app to the public, you will need to put a little more thought about security.
This is a discussion that you should have with your IT folks, since they will have security professionals who can help you ensure that your app is secure.
But they're unlikely to be familiar with the details of R, so here

[^scaling-safety-1]: If you can't assume that, you have bigger problems!

Defence in depth.

Greatest risk is accessing data that they shouldn't be able to.
And the easiest way this might happen is via an **injection attack**, where you accidentally allow the user to run arbitrary R code.

```{r setup}
library(shiny)
```

## Deployment

The first layer of defence is your deployment environment.
It should be designed so that apps are isolated --- most importantly it shouldn't be possible for one app to access the data of another app.
But it should also be designed so that one app can't steal all the resources from a server.

## Injection attacks

It's hopefully obvious that allowing the user to run arbitrary code is dangerous:

```{r}
ui <- fluidPage(
  textInput("code", "Enter code here"),
  textOutput("results")
)
server <- function(input, output, session) {
  output$results <- renderText({
    eval(parse(text = input$code))
  })
}
```

In general, any use of `parse()`[^scaling-safety-2] or `eval()` in your Shiny app is a warning sign.
(They can be ok, but only as long as they don't involve user input.)

[^scaling-safety-2]: Or any rlang `parse_` function.

But there are a few less obvious places that might surprise you:

-   Modelling functions (including model argument to `ggplot2::geom_smooth()`)

-   If you allow a user to supply a [glue](http://glue.tidyverse.org/) string to label output data, you might expect them to write something like `{title}-{number}`.
    But anything inside `{}` is evaluated by glue, so they can now execute any R code that they like.

-   You can't generally allow a user to supply arbitrary transformations to dplyr or ggplot2.
    You might expect they'll write `log10(x)` but they could also write something dangerous.
    (In particularly, this means that you shouldn't use the older `aes_string()` with user supplied input).
    You'll be safe if you use the techniques in Chapter \@ref(action-tidy).

-   If you allow the user to upload an Rmd, and you `render()` it, obviously can run arbitrary code.

Also note that Shiny input controls use client-side validation, i.e. the checks are performed in the browser, not by R.
This means it's possible for someone who understands how Shiny works to send values that you don't expect.
For example, you might expect the only possible values that a `selectInput()` could return are those listed in `choices`, but it's actually possible for them to be anything.
So avoid code like this:

```{r, eval = FALSE}
confidential <- read_csv("secrets.csv")

server <- function(input, output, session) {
  mine <- reactive(filter(confidential, user %in% input$user_id))
}

```

In general, if different users need access to different data, you should work with your IT team to design a secure mechanism so that people can't see data that .
<https://solutions.rstudio.com/auth/kerberos/>

<https://db.rstudio.com/best-practices/deployment/>

## SQL injection attack

Another common vector for injection attacks is SQL.
Don't generate SQL strings that include user input by pasting them together.
Instead use a system that escapes user input, e.g. generate the SQL with dbplyr, or use `glue::glue_sql()`.

<https://xkcd.com/327/>

For example, if you construct SQL like this:

```{r}
find_student <- function(name) {
  paste0("SELECT * FROM Students WHERE name = ('", name, "');")
}
find_student("Hadley")
```

Then "Little Bobby tables" still generates a valid SQL query:

```{r}
find_student("Robert'); DROP TABLE Students; --")
```

This query has three components components:

-   `SELECT * FROM Students WHERE name = ('Robert');` --- finds a student with name Robert

-   `DROP TABLE Students;` --- drops the `Students` table

-   `--'` --- is a comment needed to prevent the extra `'` from creating a syntax error.

## Security grab bag

-   Never put passwords in code.
    Instead either put them in environment variables, or if you have many use the [config](https://github.com/rstudio/config) package and ensure that your `config.yml` is included in `.gitignore`.

-   If you implement bookmarking (Chapter \@ref(action-bookmark)), check that you're not putting any confidential data in the URL.

-   If you use on disk caching (Section XYZ), be aware that it's shared between users.
    This can also introduce a subtle timing attack, where you can tell if someone else has (e.g.) looked at a plot by noticing how long it takes it to load for you.
