# Functions {#scaling-functions}

```{r, include = FALSE}
source("common.R")
```

Without discipline, as your app gets bigger, it will get harder and harder to understand. In turn, this makes it harder to add new features, and harder to find a solution when something goes wrong (i.e. it's harder to debug). If you're not careful, over time your app will become less and less enjoyable to work on.

Shiny allows you to have one giant `app.R` file, containing a massive `ui` specification and large `server` function. But it's hard to understand anything that runs over a single screen of code. So to continue to grow your app, you need to start to break it up into independent pieces.

This chapter will help you to take the first step in that journey, by showing you how to tactically use functions to pull out pieces into unrelated concerns. This tends to have slightly different flavours for UI and server components:

* In the UI, you have components that are repeated in multiple places with
  minor variations. Pulling out repeated code into a function reduce duplication
  (making it easier to update many control from one place), and can be
  combined with functional programming techniques to generate many controls
  at once.
  
* In the server, complex reactives are hard to debug because you need to be
  in the midst of the app. Pulling out a reactive into a separate function,
  even if that function is only called in one place, makes it substantially
  easier to debug, because you can experiment with important computational
  part of your app independent of reactivity.

I assume that you're already familiar with the basics of functions[^if-not]. The of this chapter goal is to activate your existing skills, showing you some specific cases where using functions can substantially improve the clarity of your app. We'll focus on functions that extract out code that you've already written. 

[^if-not]: If you're not, and you'd like to learn the basics, you might try reading the [Functions chapter](https://r4ds.had.co.nz/functions.html) of _R for Data Science_.

Once you've mastered the ideas in this chapter, the next step is to learn how to writing code that requires coordination across the front end and back end. This requires **modules**, which you'll learn about in Chapter \@ref(scaling-modules).

```{r setup}
library(shiny)
```

## File organisation

Before we go on to talk about exactly how you might use functions in your app, I want to point our one big immediate benefit: functions can live outside of `app.R`. There are two places you might put them depending on how big they are:

* I recommend putting large functions (and any smaller helper functions that 
  they need) into their own in `R/{function-name}.R` file.
  
* You might want to collect smaller, simpler, functions into one place.
  I often use `R/utils.R` for this, but if they're primarily used in your ui
  you might use `R/ui.R`. 
  
If you're using Shiny 1.5.0, running the app will automatically load all files in the `R/` directory, so there's nothing else to do!

If you've made an R package before, you might notice that Shiny uses the same convention for storing files containing functions. And indeed, if you're making a complicated app, particularly in conjunction with others, there are substantial advantages to making a full fledged package. If you want to do this, I recommend the reading the ["Buidling Big Shiny Apps"](https://thinkr-open.github.io/building-shiny-apps-workflow/) book and using the accompanying [golem](https://thinkr-open.github.io/golem/) package. We'll touch on packages again when we talk more about testing because packages provide useful infrastructure for testing.

## UI functions

Functions are a powerful tool to reduce duplication in your UI code. Let's start with a concrete example of some duplicated code. Imagine that you're creating a bunch of sliders that each need to each from 0 to 1, starting at 0.5, with a 0.1 step. You _could_ do a bunch of copy and paste to generate all the sliders:

```{r}
ui <- fluidRow(
  sliderInput("alpha", "alpha", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("beta",  "beta",  min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("gamma", "gamma", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("delta", "delta", min = 0, max = 1, value = 0.5, step = 0.1)
)
```

But I think it's worthwhile to recognise the repeated pattern and extract out a function. That makes the UI code substantially simpler:

```{r}
sliderInput01 <- function(id) {
  sliderInput(id, label = id, min = 0, max = 1, value = 0.5, step = 0.1)
}

ui <- fluidRow(
  sliderInput01("alpha"),
  sliderInput01("beta"),
  sliderInput01("gamma"),
  sliderInput01("delta")
)
```

Functions provide a few benefits even in this simple case:

* We can give the function a evocative name, making it easier to understand 
  what's going on when we re-read the code in the future.
  
* If we need to change the behaviour, we only need to do it in one place.
  For example, if we decided that we needed a finer resolution for the steps,
  we only need to write `step = 0.01` in one place, not four.

### Functional programming

If you're comfortable with functional programming, you could reduce the code still further:

```{r}
library(purrr)

vars <- c("alpha", "beta", "gamma", "delta")
sliders <- map(vars, sliderInput01)
ui <- fluidRow(sliders)
```

There are two big ideas here:

* `map()` calls `sliderInput01()` once for each string stored in `vars`. 
  It returns a list of sliders.
  
* When you pass a list to an html container, it automatically unpacks
  so that elements of the list because the children of the container.

If you're like to learn more about `map()` (or its base equivalent, `lapply()`), you might enjoy the [Functionals chapter](https://adv-r.hadley.nz/functionals.html) of _Advanced R_.

It's possible to generalise this further if the controls have more than one varying input. First, we create an inline data frame that defines the parameters of each control, using `tibble::tribble()`. Explicitly describing UI structure as data is a useful pattern.

```{r}
vars <- tibble::tribble(
  ~ id,   ~ min, ~ max,
  "alpha",     0,     1,
  "beta",      0,    10,
  "gamma",    -1,     1,
  "delta",     0,     1,
)
```

Then we create a function where the argument names match the column names:

```{r}
mySliderInput <- function(id, label = id, min = 0, max = 1) {
  sliderInput(id, label, min = 0, max = 1, value = 0.5, step = 0.1)
}
```

Then finally we use `purrr::pmap()` to call `mySliderInput()` once for each row of `vars`:

```{r}
sliders <- pmap(vars, mySliderInput)
```

Don't worry if this code looks like gibberish to you: you can continue to use copy and paste instead. But in the long-run, I'd recommend learning more about functional programming, because it gives you such a wonderful ability to concisely express otherwise long-winded concepts.

See Section \@ref(programming-ui) for more examples of using these techniques to generate dynamic UI in response to user actions.

### Other applications

Whenever you use the same variant of an input control in multiple places, make a function. For example,

*   If you're using a customised `dateInput()` for your country, pull out into 
    one place so that you can use consistent arguments. For example, imagine 
    you wanted a date control for Americans to use to select week days:
  
    ```{r}
    usWeekDateInput <- function(inputId, ...) {
      dateInput(inputId, ..., format = "dd M, yy", daysofweekdisabled = c(0, 6))
    }
    ```
    
    Note the use of `...`; it means that you can still pass along any other
    arguments to `dateInput()`.
  
*   Or maybe you want a radio button that makes it easier to provide icons:

    ```{r}
    iconRadioButtons <- function(inputId, label, choices, selected = NULL) {
      names <- lapply(choices, icon)
      values <- if (is.null(names(choices))) names(choices) else choices
      radioButtons(inputId,
        label = label,
        choiceNames = names, choiceValues = values, selected = selected
      )
    }
    ```


*   Or if there are multiple selections you reuse in multiple places:

    ```{r}
    stateSelectInput <- function(inputId, ...) {
      selectInput(inputId, ..., choices = state.name)
    }
    ```
    
    (If you're developing a lot of Shiny apps in your organisation, this sort
    of function is really useful to include in a shared package that everyone
    can use to make their apps more consistent.)

## Server functions

Whenever you have a long reactive (say >10 lines) you should consider pulling it out into a separate function. One of the most common mistakes I see people making is having a very large server function. Then when something goes wrong, you have use advanced debugging skills, because the failure occurs in the middle of your shiny app (and you have to do a bunch of interacting with the UI to trigger it). It's better to pull out complex pieces into their own functions so you can debug them with your usual skills.

A big downside of complex `reactive()`s is that there's no easy way to see exactly what inputs it takes; i.e. what other app state affects the output of the reactive? That's because reactives live in the global environment of your app. 

The key benefits of a function in the UI tend to be around reducing duplication. The key benefits of functions in a server tend to be around isolation and testing. When looking at a reactive expression or output, there's no way to easily tell exactly what values it depends on, except by carefully reading the code block. The function definition is a nice signpost that tells you  exactly what to inspect.  

### Reading uploaded data

Even without repetition, it's sometimes useful to pull complex code out of a reactive, and move it into a function. Take this server from Section \@ref(uploading-data). It contains a moderately complex `reactive()`:

```{r}
server <- function(input, output, session) {
  data <- reactive({
    req(input$file)
    
    ext <- tools::file_ext(input$file$name)
    switch(ext,
      csv = vroom::vroom(input$file$datapath, delim = ","),
      tsv = vroom::vroom(input$file$datapath, delim = "\t"),
      validate("Invalid file; Please upload a .csv or .tsv file")
    )
  })
  
  output$head <- renderTable({
    head(data(), input$n)
  })
}
```

If this was a real app, I'd seriously considering extracting out a function specifically for reading uploading files into its own function:

```{r}
load_file <- function(name, path) {
  ext <- tools::file_ext(name)
  switch(ext,
    csv = vroom::vroom(path, delim = ","),
    tsv = vroom::vroom(path, delim = "\t"),
    validate("Invalid file; Please upload a .csv or .tsv file")
  )
}
```

(Should it take `name` and `path` arguments or a single list?)

When extracting out such helpers, where possible avoiding taking reactives as input or returning outputs. Instead, pass them in through the arguments, and assume the caller will turn into a reactive. This isn't a hard and fast rule; sometime it will make sense for your functions to input or output reactives. But generally, I think it's better to keep the reactive and non-reactive parts of your app as separate a possible. In this case, I'm still using `validate()`; that works because outside of Shiny `validate()` works similar to `stop()`. But I keep the `req()` in the server, because it shouldn't the be responsibility of the file parsing code to know when it's run.

Since this is now an independent function, it could live in its own file (`R/load_file.R`, say), keeping the `server()` svelte. This helps keep the server function focussed on the big picture
of reactivity, rather than the smaller details underlying each component.

```{r}
server <- function(input, output, session) {
  data <- reactive({
    req(input$file)
    load_file(input$file$name, input$file$datapath)
  })
  
  output$head <- renderTable({
    head(data(), input$n)
  })
}
```

The other big advantage is that you can play with `load_file()` at the console, outside of your Shiny app. If you move towards formal testing of your app (see Chapter XXX), this also makes that code easier to test.

